Dear Loretta,

I'm writing, as you asked, to report the changes that Joe Harris and I have made since we sent you our manuscript ``Practical Curves'' to send to the reviewers.

1. We have changed the name to ``The Practice of Algebraic Curves''

2. We took the ``genus 1'' part of the old chapter 3 and combined it with the ``smooth plane curve'' part of chapter 14 and combined them in a new chapter 4 (this augments the subsequent chapter numbers by 1).

3. We substantially shortened the treatment of special curves of genus 6.

4. We slightly simplified the proof of the Brill-Noether Theorem to avoid using the duality of Grassmanians, and added some new references there.

5. We many a large number of local fixes and improvements.

6. We implemented all the suggestions in the reviews you sent us -- in particular we fixed the large number of typos and other local glitches in the annotated mss sent by one of the reviewers -- please convey our thanks for that work!

7. We have created a Macaulay2 package implementing the material in Chapters 4 and 15 (Plane curves); it will be distributed and maintained with the Macaulay2 program, which is freely available, and will be linked from the AMS web page of the book.

8. Finally, we (mostly Joe) have been edititng exercises and adding many hints. It is our plan to copy the exercises with the hints (slightly reformatted) to a separate document, which we would make available on the  AMS web page of the book. Then the hints will be deleted from the book itself. (We have added a reference to the hints document to the introduction of the book.

We hope that these changes will meet the approval of the committee!

Cordially,

David Eisenbud and Joe Harris






