             \input{header}

\chapter{Syzygies}
\label{SyzygiesChapter}


%$$
%\vbox{\offinterlineskip %\baselineskip=15pt
%\halign{\strut\hfil# \ \vrule\quad&# \ &# \ &# \ &# \ &# \ &# \ 
%&# \ &# \ &# \ &# \ &# \ 
%\cr
%degree&\cr
%\noalign {\hrule}
%0&1&--&--&--&--&--&--\cr
%1&--&17&46&45&4&--&--\cr
%2&--&--&--&--&25&18&4\cr
%\noalign{\bigskip}
%\omit&\multispan{8}{\bf Conjectural shape of $F_\bullet$}\cr
%\noalign{\smallskip}
%}}
%$$
%
%
%
%
%\centerline{\scriptsize
%\begin{tabular}{r|ccc} 
%$j\backslash i$&0&1&2\\ 
%\hline 
%0&1&$-$&$-$\\ 
%1&$-$&3&2\\ 
%\end{tabular}}
%
%
%$$
%\begin{matrix} 
%j\backslash i&\vline&0   &  1    & \cdots & n    \cr\hline
%\vdots&\vline&\vdots&\vdots & \cdots    &\vdots     \cr 
%       0&\vline&\beta_{0,0}&\beta_{1,1}&\cdots&\beta_{n,n}\cr
%       1&\vline&\beta_{0,1}&\beta_{1,2}&\cdots&\beta_{n,n+1}\cr
%\vdots&\vline&\vdots&\vdots & \cdots    &\vdots     \cr 
%\end{matrix}
%$$         
%
%
\def\length{{\rm length}}
\section{Introduction} 
%Motivation: canonical embedding turns intrinsic invariants into projective invariants. Hilbert Function. Projective Normality; Canonical Module.
%
%What are the projective invariants that correspond to Clifford index? Conjecturally, Green's conjecture. Inequality from Eagon-Northcott.
%
%Hilbert Syzygy theorem, Hilbert function derivation, Unique minimal resolution, Betti table, 
%\fix{ introduce tools as they are used}

In this Chapter we will study  invariants associated to a free resolution, or syzygies, of the homogeneous coordinate ring of a curve in projective space, with an emphasis on their relation to the varieties (or schemes) containing the curve. We have two cases in mind: curves of  (relatively) high degree, and canonical curves.


\section{Syzygies and the Hilbert Syzygy Theorem} \fix{all or part of this section should be an appendix}
\fix{introduce free resolution, minimal free resolution over a local ring, then move to
 $S = k[x_{0}, \dots, x_{n}]$ and homogeneous coordinate rings.}
 
The following result assures us that invariants of the minimal free resolution of $M$ are invariants of $M$:

\begin{theorem}[Uniqueness]\cite[Theorem ***]{E}\label{uniqueness} With hypotheses as above, the minimal free resolution of of a finitely generated graded $S$-module is unique up to (non-unique) isomorphism.
\end{theorem}

This and the other results in this section are usually stated for
modules over a local ring, and in that form the results here can be found in \cite[Ch ?]{E}. The proofs are essentially identical in the graded case presented here.

The numbers $\beta_{i,j}$ in  the minimal free resolution of $M$ are called the \emph{graded Betti numbers} of $M$; it follows from Theorem~\ref{uniqueness} that the graded Betti numbers, and all the invariants of the matrices $d_{i}$, are invariants of $M$. 
We sometimes refer to $\beta_{i,j}$ informally as the \emph{number of $i$-th syzygies of degree $j$}; more properly, it is the number of minimal generators of degree $j$ required by the module of $i$-th syzygies of $M$.

For convenience, the graded Betti numbers are usually displayed in a compact \emph{Betti table}, with the nonzero $\beta_{i,j}$ appearing in the $i$ column and the $(j-i)$-th row. (The reason for the initially non-intuitive choice $j-i$ instead of $j$ is that, for the resolution $\FF$ to be minimal, it is necessary that if $\beta_{i,j}\neq 0$, then some $\beta_{i-1,k}\neq 0$ for some $k<j$. Thus shift by $-j$ makes the diagram more compact.) This useful convention seems to go back to Bayer and Stillman and their work on the program Macaulay in the 1980s. In those days of small displays, efficiency was even more important than now.

We sometimes speak of the \emph{Betti table of a variety $X\subset \PP^{n}$}, by which we will mean the Betti table of the minimal free resolution
of the homogeneous coordinate ring $S_{X}$ of $X$, as a module over the homogeneous coordinate ring of $\PP^{n}.$

\begin{example}\label{canonical in P3}
For example, the minimal free resolution of the homogeneous coordinate ring $S_{C}$ of a canonical curve $C$ of genus 4 in $\PP^{3}$ whose ideal is generated by a quadric $q$ and a cubic $f$, as a module over the homogeneous coordinate ring $S = \CC[x_{0},\dots,x_{3}]$ of $\PP^{3}$, is the Koszul complex
\small
$$
S \lTo^{
\begin{pmatrix}
  q& f
\end{pmatrix}}
S(-2) \oplus S(-3) \lTo^{
\begin{pmatrix}
f\\-q 
\end{pmatrix}
}
S(-5)\lTo 0.
$$
\normalsize
since
$q$ and $f$ are relatively prime, and thus form a regular sequence
Thus it has  Betti table:

\centerline{\small %\scriptsize
\begin{tabular}{r|ccc} 
$j\backslash i$&0&1&2\\ 
\hline 
0&1&$-$&$-$\\ 
1&$-$&1&$-$\\ 
2&$-$&1&$-$\\ 
3&$-$&$-$&1\\ 
\end{tabular}}
\noindent where, for clarity, we have suppressed the $\beta_{i,j}$ that are 0, and replaced some of
them with $-$.
\end{example}

From now on in this section, we let $S = \CC[x_{0},\dots, x_{n}]$ be the homogeneous coordinate ring of $\PP^{n}$, and we consider only free resolutions of finitely generated graded $S$-modules. In this case something quite special happens \cite[****]{E}:

\begin{theorem}[Hilbert's Syzygy Theorem]\label{hst}
The minimal $S$-free resolution $\FF$ of any finitely generated graded $S$-module $M$ is finite: in fact, $F_{i}=0$ for all $i>n+1$.
\end{theorem}

\begin{fact}
 There are in fact finite resolutions for all $S$-modules; however there is generally no reasonable notion of a  minimal resolution outside the positively graded or finitely generated local cases. See \cite{???}.
\end{fact}

 In particular since we know the Hilbert function of $S$, and thus of $S(-j)$,
$$
H_{S(-j)}(m) = 
\begin{cases}
 {n+m-j \choose n} & \hbox{for $m\geq j$}
  \\ 
 0 & \hbox{for $m<j$}
\end{cases}
$$
we can compute the Hilbert function for a finitely generated graded $S$-module $M$ from its graded Betti numbers 
$$
H_{M}(m) = \sum_{i,j} (-1)^{i}\beta_{i,j}{n+m-j\choose n}.
$$
\begin{example} [The homogeneous coordinate ring of a plane curve] 
For the simplest non-trivial example, if $C$ is a plane curve with equation $F(x_{0},x_{1}x_{2})$ of 
degree $d$, then the minimal free resolution of the homogeneous coordinate ring $S_{C}$ has the form
$$
\FF: S\lTo^{F}S(-d)\lTo 0
$$
Thus the only nonzero graded Betti numbers of $S_{C}$ as an $S$-module are
 $\beta_{0,0} = \beta_{1,d} = 1$, so
$$
 H_{S_{C}}(m) = {2+m\choose 2} - {2-d+m\choose 2}
 $$
 which has the value  $dm + 1+(2-d)(2-d-1)/2 = dtm+ (d-1)(d-2)/2 +1$. Since the degree is the leading coefficient and the arithmetic genus \fix{where is this defined?} is the constant term minus 1, we get
$\deg C = d,\ p_{a}(C) = (d-1)(d-2)/2$ as we should expect.  Note that this method of computation is not sensitive at all to whether the curve is smooth and irreducible; it works for any divisor in $\PP^{2}$.
\end{example}

\begin{exercise} 
\begin{enumerate}
 \item Show that any curve in $\PP^{3}$ whose homogeneous coordinate ring has the
 form given in Example ~\ref{canonical in P3} has degree 6 and (arithmetic) genus 4.
 
\item In Section **** we computed the genus of a smooth complete intersection curve inductively, by using the adjunction formula for curves on a smooth surface along with Bertini's Theorem.
Show that the same formula gives the arithmetic genus of an arbitrary 1-dimensional complete intersection $C$ hypersurfaces 
$X_{i} := \{F_{i} = 0\}$ of degrees $d_{i}$,
using the fact that the minimal free resolution of the homogeneous coordinate ring 
$S/(F_{1}, \dots, F_{n-1})$ of $C$  is the Koszul complex (see \cite{???}):
%\scriptsize
\small
$$
 S \lTo \oplus_{1\leq i<n}S(-d_{j}) \lTo  \oplus_{1\leq j<k<n}S(-d_{j}-d_{k}) \cdots S(-\sum_{1\leq j<n}d_{j})\lTo 0.
 $$
\normalsize
\end{enumerate}
\end{exercise}

Computing the Hilbert function was the application for which Hilbert proved the Syzygy Theorem; but if that were the only use of free resolutions, we would not be introducing them here. A more interesting feature that the Betti table of the homogeneous coordinate ring (or homogeneous ideal) of a variety $X$ reflects is the presence of a variety $Y$ containing 
$X$ and having low degree syzygies. 

Here is a  simple example:

We have seen in Chapter~\ref{Scrolls}, the homogeneous ideal $I$ of the twisted cubic curve in $\PP^{3}$ is generated by the three $2\times 2$ minors of the matrix
$$
M = \begin{pmatrix}
 x_{0}&x_{1}&x_{2}\\
  x_{1}&x_{2}&x_{3}
\end{pmatrix}
$$
We shall soon compute the Betti tables of all rational normal scrolls. In particular, we shall see that the Betti table of the twisted cubic has the form:

\centerline{\small
\begin{tabular}{r|ccc} 
$j\backslash i$&0&1&2\\ 
\hline 
0&1&$-$&$-$\\ 
1&$-$&3&2\\ 
\end{tabular}}
\noindent
where the dashes indicate that the corresponding $\beta_{i,j}$ is zero. In plain language, the table indicates  that  the ideal $I$ of the curve is minimally generated by 3 elements of degree 2, and that there are two linear relations on the 3 (that is, generators of total degree $2+1=3$, and that these freely generate the module of  syzygies of $I$ (the two relations are in fact just the columns of the transpose of the matrix $M$).

The homogeneous ideal $J$ of any set of points $\Gamma$ that lies on the twisted cubic contains $I$. Supposing that $\Gamma$ does not lie on a plane, the ideal $J$ contains no linear forms. The 2 linear relations on the generators of $I$ must be consequences of relations on the generators of $J$. But since the relations on $I$ are linear, they are of the lowest possible degree of any relations on $J$, so there must be at least 2 linear relations on the quadratic generators of $J$. Here is the way this can be used:
\begin{fact}
 A lemma of Castelnuovo \fix{exercise in Ch 8; is it really Castelnuovo?} shows  that every set of 6 points in linearly general position
in $\PP^{3}$ lies on a twisted cubic;  a set of points $\Gamma\subset \PP^{3}$ in linearly general position of cardinality $n\geq 7$ lies on a twisted cubic curve if and only if
the Betti table of $\Gamma$ is termwise $\geq$ that of the twisted cubic.\end{fact}

\section{What Makes a Complex Exact}

Given that the resolution of any module over a polynomial ring $S = k[x_1,\dots,x_n]$ (or regular local ring) has  length bounded by $n$, the resolution of an $S$-module that is a $k$-th syzygy---that is, the image of the $k$-th map in a resolution---must have length $\leq n-k$. What is special about the presentation matrix of such a module? -- put differently, what is it about the $k$-th map in a resolution that is different than the first map?

One answer to this question was given by David Hilbert in his orginal paper \cite{} and another can be given in terms of Gr\"obner bases and initial ideals \cite{}. Here we give a third answer, one that lends itself better to conceptual proofs, such as the proof of exactness of the Eagon-Northcott complex resolving the ideal of minors of a matrix given in Section~\ref{}. We give the result in a special case; in its general form in \cite{} it applies to finite free complexes over any ring.

\begin{theorem}\label{WMACE} Let 
$$
{\bf F}:  \quad 0\rTo F_m\rTo^{\phi_m} \cdots\rTo^{\phi_2} F_1\rTo^{\phi_1} F_0
$$
be a finite complex of finitely generated free modules over a polynomial ring or, more generally, a Cohen-Macaulay ring. The following conditions are equivalent:
\begin{enumerate}

\item The complex is exact (and thus a resolution of $\coker \phi_1$);

\item The ranks $r_i$ of the maps $\phi_i$ satisfy $r_i+r_{i+1}  = \rank F_i$ and the $r_i\times r_i$ minors of a matrix representation of
$\phi_i$ generate an ideal of codimension $\geq i$.

\item For every prime ideal $P$ of codimension $c$, the localize truncated complex 
$$
({\bf F}_{\geq c})_P: \quad 0\rTo (F_m)_P\rTo^{(\phi_m)_P} \cdots\rTo^{(\phi_{c+1})_P} (F_c)_P
$$
is split exact (and thus is a free resolution of a projective module.)
\end{enumerate}
\end{theorem}

The proof that exactness implies items 2,3 is easy to summarize: if ${\bf F}$ is a free resolution of an $S$-module $M$ and $P$ is a prime of codimension $c$
then $({\bf F})_P$ is a free resolution of
$M_P$ over $S_P$, a regular ring of dimension $c$.Thus the kernel of $(\phi_{c-1})_P$ must be projective (actually, free) over $S_P$, 
and $({\bf F}_{\geq c})_P$ is a free resolution of this module, and thus split exact. Taking $P = 0$ gives the condition on ranks in
item 1), while the condition on codimension follows because the cokernel of each $(\phi_i)_P$ is projective for $i\geq c$. For a detailed treatment and the opposite implications, we refer to \cite{book}.

\section{Castelnuovo's theorem: $2g+1$ is projectively normal.}

\fix{I'm inclined to make this a ``cheerful fact''}

 Regularity in terms of resolution and in terms of local cohomology. (has the relation of local to global coho been done already?)

\fix{We'll need local cohomology, local duality, 
WMACE -- we need this for Hilbert-Burch in 10-DualityAndLinkage}

recall the $2g+1$ theorem from Ch 4. Say that it's ACM+smooth. relate it to regularity as in "geom of syz". Say that the projective
normality of smooth CI's is really ACM for all CI's. Say that ACM is measured by the length of the free resolution.




\subsection{How syzygies can reflect geometry}\label{syzy and geom}

One of the main ways in which syzygies can be seen to reflect the geometry of an embedded  reduced, irreducible curve $C\subset \PP^r$
(or other variety) is linked to the possibility of factoring the line bundle $\sO_C(1)$ as the tensor product of two bundles with sections. Suppose for example that $\sO_C(1) = \sL_1\otimes \sL_2$. Choose 2 independent global sections
$\sigma_1, \sigma_2$ of  $H^0(\sL_1)$ and a basis $\tau_1,\dots, \tau_n$ of $H^0(\sL_2)$. Set
$l_{i,j}= \sigma_i\tensor \tau_j$ and consider the matrix 
$$
M = 
\begin{pmatrix}
 l_{1,1}&l_{1,2}&\dots&l_{1,n}\\
  l_{2,1}&l_{2,2}&\dots&l_{2,n}
\end{pmatrix}.
$$
We claim that  the $2\times 2$ minors $l_{1,j} l_{2,j'}-l_{1,j'}l_{2,j}$ are in the homogeneous ideal of $I_C$ of $C$ in $\PP^r$. 

\begin{example}
The most familiar example is that of the twisted cubic. In this case the global sections $x_0\dots x_3$ of $\sO_C(1)$ may be identified with the forms $s^3, s^2t, st^2, t^3 \in k[s,t]$, and if $p\in C \cong \PP^1$ then the multiplication of sections
in the factorization  $\sO_C(1) = \sO_C(p) \otimes \sO_C(2p)$ 
$$
\bordermatrix{
 &s^2&st&t^2\cr
 s& s^3&s^2t&st^2\cr
 t& s^2t&st^2&t^3
}
$$
 leads to the familiar matrix
$$
\begin{pmatrix}
x_0&x_1&x_2\\
x_1&x_2&x_3 
\end{pmatrix}.
$$
We have $I_2(M) = I_C$, and the same idea works for the rational normal curve of any degree.

\end{example}

To see that, in general, $I_C$ contains the ideal $I_2(M)$ generated by the $2\times 2$ minors of $M$, 
let $K(C)$ be the ring of rational functions on $C$ \fix{have we made this definition somewhere?}. Choosing identifications $\sL_i\otimes K(C) \cong K(C)$ we see that the $\sigma_i$ and the $\tau_j$ commute with each other as elements of $K(C) \otimes_{\sO_X} K(C)$, and thus 
$$
\bigl(l_{1,j} l_{2,j'}-l_{1,j'}l_{2,j}\bigr)|_C = \sigma_1\tau_j\sigma_2\tau_j' - \sigma_1\tau_j'\sigma_2\tau_j =0.
$$

In case $C$ is reduced and irreducible the matrix above has a special property: $K(X)$ is a domain, so no product of a nonzero
section of $\sL_1$ with a nonzero section of $\sL_2$ can be zero. We can state this without any reference to $C$:

\begin{definition}
Let $R$ be a commutative ring. A map $M:R^n\to R^m$ is 1-generic if the kernel of the corresponding
 map $R^{n}\otimes R^{m*} \to R$  contains no pure tensor $a\otimes b$. In more concrete terms, a matrix
$M$ is \emph{1-generic} if there are no invertible matrices $A,B$ such that  $AMB$ has some entry equal to 0.
\end{definition}

By the material in Chapter~\ref{scrolls}, the ideal $I_2(M)$ of a 1-generic matrix of linear forms is the homogeneous ideal of a rational normal 
scroll
of codimension $n-1$ and degree $n$. 

In the next section we will show that it has a free resolution of a special form called the 
Eagon-Northcott complex that is a subcomplex of the minimal free resolution of $I_C$. The presence of such a variety containing $C$ or
a subcomplex of this special form in the minimal free resolution of $C$ is thus necessary for the 
factorization of the line bundle $\sO_C(1)$ as above, and it is sometimes sufficient, as well.

\section{The Eagon-Northcott Complex of a $2\times n$ matrix of linear forms}

The Eagon-Northcott complex is a complex of free modules associated to any matrix over any commutative ring. The most familiar special case is the Koszul complex, which one may think of as the Eagon-Northcott complex of a $1\times n$ matrix, and  even in the general case the Eagon-Northcott complex is in a sense built out of the Koszul complexes. A full treatment of the Eagon-Northcott complex and a whole family of related constructions can be found in 
\cite[Appendix ***]{E}, and, from a more conceptual and general point of view, in \cite{Weyman}. Here we will only
make use of the case of a matrix such as the one above, we will present a simplified account in that case only. Here is the result we need:

\begin{theorem}\label{Eagon-Northcott}
 Let $S = k[x_0,\dots, x_r]$ be a polynomial ring,  and let $M: F\to G$ be a homomorphism with
 $F = S^n(-1), G= S^2$. If $M$ is 1-generic, then the minimal free resolution of $S/I_2(M)$ has the form:
\begin{align*}
EN(M) := 
S \lTo{\bigwedge^2 M} 
 \bigwedge^2 F&
 \lTo^{\delta_{2}}
 S^{2*}\otimes \bigwedge^3 F  \lTo^{\delta_{3}}
  (\Sym^2S^{2})^*\otimes\bigwedge^4F  \\
 &\lTo^{\delta_{4}}\cdots\lTo^{\delta_{n-1}} 
(\Sym^{n-2}S^{2})^*\otimes\bigwedge^nF 
 \lTo 0.
\end{align*}
\end{theorem}

From Chapter **** we know also that the ideal of minors defines a rational normal scroll.

\begin{proof} We first show that $r\geq n$; more precisely, we show that the span of the entries of $M$ has dimension $\geq n+1$. As noted above, to say that the $2\times n$ matrix of linear forms $M$ is 1-generic means that
the kernel of the corresponding map $ \phi: k^2\otimes k^n \to S_1$ contains no pure tensors. In the projective
space $\PP^{2n-1} = \PP(k^2\otimes k^n)$ the pure tensors form a variety isomorphic to $\PP^1\times \PP^{n-1}$, and thus of dimension $n$. Consequently the kernel of $\phi$ can have dimension at most $n-1$, whence the image of $\phi$ 
in $S_1 = k^{r+1}$ has dimension at least $2n-(n-1) = n+1$. 

We begin the discussion of $EN(M)$ by defining the maps $\delta_i$ and and proving that the given sequence is indeed a complex---that is, consecutive maps compose to 0. For simplicity of notation, we choose a generator of $\wedge^2 S^2$
 and identify it with $S$, which gives a sense to the map labeled $\bigwedge^2M$.
 
  Although it is not hard to do this directly, the dual maps
 $$
 \partial_i: \Sym^{i-2} G \otimes \bigwedge^i F^* \rTo \Sym^{i-1} G \otimes \bigwedge^{i+1} F^*
 $$
 have a more familar-looking description, so we define these instead. Indeed, the map $M$ corresponds to an
 element $\mu\in G\otimes F^*$. We may think of $ \Sym^{i-2} G \otimes \bigwedge^i  F^*$
 as a (bigraded) component of the exterior algebra over $ \Sym G$ of 
 $$
  \Sym G \otimes \bigwedge_S  F^*= \bigwedge_{ \Sym G} (\Sym G \otimes  F^*).
 $$
We define $\partial_i$ to be  multiplication by $\mu$ in the sense of this exterior algebra. Since $\mu$ has degree 1
in this sense, its square is 0. 

To show that $(\bigwedge^2 M)\circ \delta_2$ is zero, it is simplest to choose a matrix representing $M$.
Direct computation using only the usual expansion of a determinant
along a row shows that, up to sign,
pure basis vector $e\otimes f_i\wedge f_j\wedge f_k$ of $G^*\otimes \bigwedge^3 F$
maps under the composition $(\bigwedge^2) M\circ \delta_2$ to the determinant
of the $3\times 3$ matrix obtained from $M$ by repeating the row corresponding to $e$ and
the columns $i,j,k$. This determinant is 0 because it has a repeated row.

We next prove the split exactness of a complex of the form $EN(M')$ where $M'$ is surjective, so that we
may write $F = G\oplus F'$ and the map $M': G\oplus F' \to G$ as projection on the first factor. 
Of course
it suffices to prove the split exactness of the dual sequence, $EN(M')^*$:
\begin{align*}
EN(M')^* := 
S \rTo{\bigwedge^2 M'^*} 
 \bigwedge^2 F^*&
 \rTo^{\partial_{2}}
 G\otimes \bigwedge^3 F^*  
 \rTo^{\partial_{3}}
  \Sym^2G\otimes\bigwedge^4F^*  \\
 &\rTo^{\partial_{4}}\cdots\rTo^{\partial_{n-1}} 
\Sym^{n-2}G\otimes\bigwedge^nF^* 
 \rTo 0.
\end{align*}
In this case the proof
is an exercise in multilinear algebra. 
We begin by proving split exactness at the 
positions $\Sym^{i} G \otimes \bigwedge^{i+2}  F^*$ where $i\geq 1$.

The module
$ \Sym^{i} G \otimes \bigwedge^{i+2}  F^*$
decomposes as
\begin{align*}
&\Sym^{i} G \otimes \bigwedge^2 G^* \otimes \bigwedge^{i} F'^*\oplus \\
&\Sym^{i} G \otimes  G^*\otimes \bigwedge^{i+1} F'^* \oplus \\
&\Sym^{i} G \otimes  \bigwedge^{i+2} F'^* 
\end{align*}
Note that under our hypothesis, the element $\mu' \in G\otimes F^* = G\otimes G^* \oplus G\otimes F'^*$
has the form $(\mu_G, 0)$, where $\mu_G$ represents the identity map $G \to G$. Thus the complex
$EN(M')^*$ is a direct sum over $i$ of 3-term complexes of the form
%$$
%\Sym^{i+1} G \otimes \bigwedge^2 G^* 
%\lTo^{-\wedge \mu'} 
%\Sym^{i} G \otimes  G^*
%\lTo^{-\wedge \mu'} 
%\Sym^{i-1} G
%$$
$$
\Sym^{i-1} G 
\rTo^{-\wedge \mu'} 
\Sym^{i} G \otimes  G^*
\rTo^{-\wedge \mu'} 
\Sym^{i+1} G \otimes \bigwedge^2 G^* 
$$
tensored with various $\bigwedge^j F'^*$, and it suffices to show that the former are split exact when
$i\geq 0$. Now $\Sym G$ may be identified with $R:= S[x,y]$, where $x,y$ are a basis of $S^2$, and
as such the sequences above may be identified with components of the Koszul complex of $x,y$ over $R$,
%$$
%R\otimes \wedge^2 G\lTo R\otimes G \lTo R .
%$$
$$
0\to R \rTo R\otimes G \rTo R\otimes \wedge^2 G
$$
The only homology of this sequence is $R/(x,y)R$ at the right so if we replace  $R\otimes \wedge^2 G \cong R$ by the ideal $(x,y)R$, this sequence is a split exact sequence of free $S$-modules. This is the desired result.

It remains to treat the beginning of the complex $EN(M')^*$,
$$
S \rTo{\bigwedge^2 M'^*} 
 \bigwedge^2 F^*
 \rTo^{-\wedge \mu'}
G\otimes \bigwedge^3 F^*
$$
which, in our case, may be written:
%$$
%\wedge^2 F^* = \bigwedge^2 G^* \oplus (G^*\otimes F'^*) \oplus \bigwedge^2 F'*
%$$
%Consider the pair of maps
\begin{align*}
S \rTo{\bigwedge^2 M'^*} 
 &\bigwedge^2 G^* \ \oplus\ (G^*\otimes F'^*)\ \oplus\ \bigwedge^2 F'^*
 \rTo^{-\wedge \mu'} \\
 &G\otimes \bigwedge^2 G^*\otimes F'^*\ \oplus\ (G\otimes G^*\otimes \bigwedge^2 F'^*)\ \oplus\ G\otimes \bigwedge^3 F'^*
\end{align*}
The map $\bigwedge^2 M'$ is the projection to $\bigwedge^2 G$ composed with the chosen isomorphism
$\bigwedge^2 G \cong S$, and is thus a split monomorphism. To complete the argument, we must show that
 the map marked $-\wedge \mu'$ is a monomorphism on $(G^*\otimes F'^*) \oplus \bigwedge^2 F'^*$.
 But this map is the direct sum of the two maps
  $$
 (G^* \rTo^{-\wedge \mu'} G\otimes \bigwedge^2 G^*)  \otimes F'^*
 $$
 and
 $$
(S  \rTo^{\mu'} G\otimes G^*) \otimes \bigwedge^2 F'^*
 $$
 which are evidently split monomorphisms. 
completing the proof of split exactness of $EN(M')^*$ and thus of $EN(M')$.


To go further we use a basic result, proven in a more general form (and with a slightly different statement) in \cite[Theorem ***]{E}. We make the convention
that the codimension of the empty set is infinity.

\begin{theorem}\label{WMACE}
 Let $S = k[x_0,\dots, x_r]$, and
 $$ 
\FF:  F_0\lTo^{\phi_1}F_1 \lTo \cdots \lTo F_{n-1}\lTo^{\phi_n} F_n\lTo 0
 $$
be a finite complex of free $S$-modules. Set
$$
X_i = \{p\in \AA^{n+1} \mid  H_i(\FF \otimes \kappa(p)) \neq 0\}
$$
The complex $\FF$ is \emph{acyclic} (that is, $H_i(\FF) = 0$ for all $i>0$) if and only if
$$
\codim X_i \geq i
$$
for all $i>0$. Moreover, $X_{0}\supseteq X_{1}\supseteq \cdots \supseteq X_{n}$
\qed
\end{theorem}

For example, Nakayama's Lemma implies that $X_{0}$ is the support of $\coker \phi_{1}$; thus $X_{0}$ is the set defined by the $\rank F_{0}$-sized minors of $\phi_{1}$. Similarly, 
and that $X_{n}$ is the support of the cokernel of the dual of $\phi_{n}$. 

Also, if $n=1$, the theorem simply says that a map $F_1\to F_0$ is a monomorphism iff it becomes a monomorphism after tensoring with the field of rational functions $K$, which follows from the flatness of
localization and the fact that $F_1$ is torsion-free, so that
$F_1 \subset F_1 \otimes K$. 

\begin{fact}
Theorem~\ref{WMACE} is true in this form over any Cohen-Macaulay ring; for more general
rings, ``codimension'' must be replaced by ``grade'', as in the given reference.
The Theorem can be generalized
to case where the $F_i$ are not free, but are sufficiently ``like'' free modules, too.
\end{fact}

Conclusion of the proof of Theorem~\ref{Eagon-Northcott}.
Let $X_{i}\subset \AA^{r+1}$ be the variety defined from the complex $EN(M)$ as in 
Theorem~\ref{WMACE}. Since $EN(M)$ becomes split exact after inverting any $2\times 2$ minor of $M$
$X_{i}$ is
contained in the closed set defined by $I_{2}(M)$, for all $i$. Thus if $I_{2}(M)$ has codimension $n-1$,
then $EN(M)$ is acyclic. 
\end {proof}

\fix{START COMMENTED OUT MATERIAL}
We will also use a special case of the Auslander-Buchsbaum formula connecting projective dimension and depth:

\begin{theorem}\label{Auslander-Buchsbaum}
If $R$ is a regular local ring of dimension $d$, and $M$ is a finitely generated $R$-module, then the projective dimension of $M$ is $\leq d$ with equality only if
$M$ contains a submodule of finite length. 
\end{theorem}

\begin{corollary}\label{associated primes}
If $R$ is a regular local ring of dimension $d$, and $M$ is a finitely generated $R$-module, then the codimension of an associated prime of $A$ is at most the projective dimension of $A$. 
\end{corollary}
\begin{proof}[Proof of  Corollary~\ref{associated primes}]
 Projective dimension can only decrease under localization, and
 the associated primes $P$ of $A$ are those for which $A_{P}$ contains a submodule
 of finite length.
\end{proof}

With this and the multi-linear algebra above we can  prove the basic acyclicity result for an Eagon-Northcott complex:

\fix{the Theorem as now stated doesn't need the following. I've copied the short proof
of acyclicity into the end of the proof above.}
\begin{proposition}\label{acyclicity}
Let $S = k[x_0,\dots, x_r]$ be a polynomial ring,  and let $M: F\to G$ be a homomorphism with
 $F = S^n(-1), G= S^2$.
 $$
 S^n \cong F \rTo^M G \cong S^2
 $$
 is a (not necessarily homogeneous) map of free $S$-modules.
 The Eagon-Northcott complex $EN(M)$ is acyclic if and only if $\codim I_2(M) \geq n-1$, in which case the dual complex is also acyclic and
 the associated primes of $I_2(M)$ are all minimal and of codimension $n-1$.
 \end{proposition}

\begin{proof}[Proof of Proposition~\ref{acyclicity}]
Let $X_{i}\subset \AA^{r+1}$ be the variety defined from the complex $EN(M)$ as in 
Theorem~\ref{WMACE}. Since $EN(M)$ becomes split exact after inverting any $2\times 2$ minor of $M$
$X_{i}$ is
contained in the closed set defined by $I_{2}(M)$, for all $i$. Thus if $I_{2}(M)$ has codimension $n-1$,
then $EN(M)$ is acyclic. 

In this case the projective dimension
of $S/I_{2}(M)$ is $n-1$, so all the associated primes of $I_{2}(M)$ have codimension
exactly $n-1$.

If $EN(M)$ is  acylic then, by Theorem~\ref{WMACE}, the codimension of $X_{n-1}$ is at least $n-1$. Thus to prove that the acyclicity of $EN(M)$ implies  $\codim I_{2}(M) \geq n-1$ (and thus
$\codim I_{2}(M))  n-1$, it suffices to show that $X_{n-1} = X_{0}$ as algebraic sets.

To see this, note that
the ideal of $2\times 2$ minors of $M$. By definition, $X_{n-1}$ 
is the set of points $p$ where $\kappa(p)\otimes \delta_{n-1}$ is not an inclusion, 
or equivalently, that 
that 
$$
\kappa(p)\otimes F\otimes \Sym^{n-3}G 
\cong 
\kappa(p)\otimes \bigwedge^{n-1} F^{*} \otimes  \Sym^{n-3}G 
\rTo^{\partial_{n-1}}
\kappa(p)\otimes \bigwedge^{n} F^{*} \otimes  \Sym^{n-2}G
\cong
\kappa(p)\otimes \Sym^{n-2}G
$$
is not a split surjection, and it is easy to see that the composite map takes
$a\otimes b$ to $\kappa(p) \otimes M(a)\cdot b$, so the cokernel is the $(n-2)$-nd symmetric power
of the cokernel of the map $\kappa(p) \otimes M$. Thus $X_{n}$ is equal to the support
of the cokernel of $M$ itself. 

By Nakayama's Lemma, $X_{0}$ is the support of $M$; furthermore, the localization of $\coker M$ at $p$ is 0 if and only if
one of the 
$2\times 2$ minors of $M$ is a unit locally at $p$ so $X_{0}$, so this is defined
set-theoretically by $I_{2}(M)$.

It now follows from Theorem~\ref{WMACE} that all the $X_{i}$ are equal, so $EN(M)$ is 
acyclic if and only if $EN(M)^{*}$ is acyclic. 


Since $M$ is 1-generic the entries of of the second row of $M$ are linearly independent, and since the dimension of the span of all the linear forms is at least $n+1$, some element in the first row is outside the span of the the elements in the second. After a permutation of columns we may assume that $l_{1,1}, l_{2,1}, l_{2,2},\dots l_{2,n}$ are linearly independent, and we may take them to be a subset of the variables, say $x_{0},\dots x_{n+1}$

We next show by induction on $n$ that  $I_{2}(M)$ is prime. In the case $n=2$ we have $I_{2}(M) =  (x_{0}x_{2}-x_{1}l_{1,2})$ which obviously does not factor. 

Now suppose that $n>2$, and let $M'$ be the matrix $M$ with the first column omitted. we know by induction that $I_{2}(M')$ is prime of codimension $n-2$. Since $I:=I_{2}(M)$ does not have the maximal ideal as an associated prime, it is saturated. The ideal $I_{2}(M)+x_{0}$ properly contains $I_{2}(M')$ and thus has codimension $\geq n-1$ in $S/x_{0}$, whence we see that every component of $I_{2}(M)$ meets the open set
$x_{0} = 1$. Restricting to this open set
\end{proof}


The first non-trivial example of a finite free resolution is the Koszul complex on 3 variables, which is the minimal $S = k[x,y,z]$-free resolution of the module $S/(x,y,z)$:
$$
0\to S(-3) \rTo^{
\begin{pmatrix}
x\\y\\z 
\end{pmatrix}}
 S^3(-2) \rTo^{\begin{pmatrix}
0&-z&y\\
z&0&-x\\
-y&x &0
\end{pmatrix}}
S^3(-1) \rTo^{
\begin{pmatrix}
x&y&z
\end{pmatrix}}
S
$$
In fact this is the first example that Hilbert presented in his famous paper \cite{}. 																											
\fix{END COMMENTED OUT MATERIAL}

\section{Canonical Curves}
We follow the treatment in \cite{Schreyer}, and treats a more general situation than that of the 
images of smooth curves under the canonical embeddings.

We define a \emph{canonical curve} in $\PP^{g-1}$ to be a purely one-dimensional, nondegenerate closed subscheme  such that
$$
 h^{0}(\sO_{C}) = 1,\ h^{0}(\sO_{C}(1) = g, \hbox{ and } \omega_{C} = \sO_{C}(1).
$$

The first of these conditions is always satisfied when $C$ is reduced and connected. The last two conditions imply that $C$ is locally Gorenstein, and that $C$ is embedded in $\PP^{g-1}$ by $g$ independent sections of the dualizing, line bundle.

We say that a canonical curve $C$ has a \emph{simple}  $g-2$ secant if $C$ contains $g-2$ smooth points spanning a $(g-3)$ dimensional
plane in $\PP^{g-1}$ that meets $C$ only in the $g-2$ points. This hypothesis is always satisfied if $C$ is irreducible. More generally:

\begin{lemma}
 If $C\subset \PP^n$ is a reduced, irreducible, nondegenerate curve, and $m\leq n-2$, then the linear span $L := \overline{p_1,\dots, p_m}$
 of $m$ general points of $C$ is a simple $m$-secant; that is, a plane of dimension $m-1$ such that
 $C\cap L = \{p_1,\dots,p_m\}$ scheme-theoretically.
 \end{lemma}
 
 
\begin{proof}
The plane $L$ is contained in a hyperplane $H$, and since the points are general, we may take this to be a general hyperplane. By Bertini's Theorem, $C\cap H$ is reduced, so $C\cap L$ is also reduced.
 If $C\cap L$ had length $>m$, then by Theorem~\ref{uniform position??} every set of $m+1$ points of $C\cap H$ would be dependent,
 and the span of $C\cap H$ would thus have dimension $\leq m-1<n-1$, and we could choose a hyperplane section $C\cap H'$ with more points than $C\cap H$, which is absurd.
\end{proof}

\begin{theorem}[Max Noether]\label{canonical curves are ACM}
A canonical curve in $\PP^{g-1}$ has degree $2g-2$ and arithmetic genus $g$. If the curve has a simple
$g-2$ secant, then it is arithmetically Cohen-Macaulay; that is,
$\HH^{1}\sI_{C/\PP^{g-1}}(m) = 0$ for all $m\in \ZZ$.
\end{theorem}
 

Note that for a canonically embedded irreducible curve, the hypothesis of having a simple $g-3$-dimensional $g-2$ secant plane $\Lambda$ is the same as having a base-point free pencil of degree $g-2$; this is the linear series 
cut out by the hyperplanes containing $\Lambda$.

 Note that these hypotheses imply that $C$ is connected and Gorenstein.
 
\begin{proof}
We begin by showing that $\sO(-m)$ has no global sections for $m>0$.
If $D$ is a divisor equivalent to $n$ times the hyperplane section, we have an exact sequence
$$
0\to \HH^{0}(\sO_{C}(-n)) \to \HH^{0}(\sO_{C}) \to \HH^{0}(\sO_{D}) \to \cdots.
$$
By hypothesis, the vector space $\HH^{0}\sO_{C}$ is spanned by the constant functions, and these
restrict non-trivially to $\sO_{D}$, and $\HH^{0}(\sO_{C}(-n)) = 0$ as claimed.

Using the Riemann-Roch Theorem we can now compute the Hilbert function $\chi_{C}(m)$:
We have 
$$
\chi_{C}(0) = h^{0}(\sO_{C}) - h^{1}(\sO_{C}) = h^{0}(\sO_{C}) - h^{0}(\omega_{C}) = 1-g.
$$
On the other hand 
$$
\chi_{C}(0) = h^{0}(\sO_{C}(1)) - h^{1}(\sO_{C}(1)) = h^{0}(\sO_{C}(1_{}) - h^{0}(\sO_{C}) = g-1.
$$
Since $\chi_{C}(m)$ is a polynomial of degree1 in $m$, we deduce
$\chi_{C}(m) = (2g-2)m -g+1$, whence we see that the degree of $C$ is $2g-2$ and $\p(C) = g$ as claimed.

To show that
$C$ is arithmetically Cohen-Macaulay we use the sequence
$$
\cdots \to \HH^{0}(\sO_{\PP^{n}}(m)) \to \HH^{0}(\sO_{C}(m))
\to \HH^{1}(\sI_{C}(m))\to \HH^{1}(\sO_{\PP^{n}}(m)) \to\cdots .
$$
Since $\HH^{0}(\sO_{\PP^{n}}(m)) = 0$, it
is enough to show that the natural map 
$$
\HH^{0}(\sO_{\PP^{n}}(m)) \to \HH^{0}(\sO_{C}(m))
$$
 is surjective for all $m\in \ZZ$. For $m=0,1$ this is immediate from the hypothesis.

For $m <0$ we must show $\HH^{0}(\sO_{C}(m))=0.$ 
If $D$ is a divisor equivalent to $-m$ times the hyperplane section, we have an exact sequence
$$
0\to \HH^{0}(\sO_{C}(m)) 
\to \HH^{0}(\sO_{C}) 
\to \HH^{0}(\sO_{D}) \to \cdots.
$$
By hypothesis, the vector space $\HH^{0}\sO_{C}$ is spanned by the constant functions, and these
restrict non-trivially to $\sO_{D}$, so the kernel, $\HH^{0}(\sO_{C}(m))$, is 0 as claimed. 

To prove surjectivity for $m\geq 2$ we use the remaining hypothesis, the existence of
a simple $g-3$-dimensional $g-2$ secant plane $\Lambda$  and an idea sometimes called the \emph{base-point-free pencil trick}. Let $p_{0},\dots p_{g-3}$ be the points in which $\Lambda$ meets $C$.  Since the
$p_{i}$ are linearly independent by hypothesis, we may choose homogeneous coordinates $x_{i} \in \HH^{0}(\sO_{C}(1))$ so that
$x_{i}(p_{j} \neq 0$ if and only if $i = j$. It follows that the sections
$x_{i}^{m}$ of $\sO_{C}(m)$ span $\HH^{0}(\sO_{C}(m)|_{\{p_{0}, \dots, p_{g-3}\}}$. Let 
$V\subset \HH^{0}(\sO_{C}(1))$ be the two-dimensional subspace of linear forms vanishing on
$\Lambda$, and thus on the $p_{i}$. 

For $m\geq 2$ there are maps of vector spaces
$$
\wedge^{2} V\otimes \HH^{0}(\sO_{C}(m-2)) \to V\otimes \HH^{0}(\sO_{C}(m-1)) 
\to \HH^{0}(\sO_{C}(m))
$$
where the right hand map is multiplication and the left hand map sends
$s_{1}\wedge s_{2}\otimes \sigma$ to $s_{1}\sigma-s_{2}\sigma$ for any local section $\sigma$.
The sequence is exact because the sections $s_{1},s_{2}$ that span $V$ never vanish simultaneously except on the $p_{i}$, and has image  consisting of sections that vanish on the points $p_{i}$

\end{proof}

\begin{corollary}
If $C\subset \PP^{g-1}$ is a canonical curve with a simple $g-3$-secant  
\end{corollary}
Theorem~\ref{canonical curves are ACM} implies, in particular, that the homogeneous coordinate ring of $C$ can be identified with $\oplus_{n\in \ZZ}\HH^0\sO_C(n)$. It follows from the Riemann-Roch theoall canonically embedded smooth curves of given genus have the same Hilbert functions. 

\section{Syzygies and the Clifford index}

For a "modern" treatment of Petri's thm, recommended by Michael Kemeny:
 E. Arbarello and E. Sernesi, Petri’s approach to the study of the ideal associated to a special divisor,
Inventiones Math. 49 (1978), 99-119.

Green's Conjecture, Voisin's Theorem, Ein-Lazarsfeld Theorem --- all cheerful facts. proof that low gonality implies long linear part.

As an application of the ideas above, suppose that $C\subset \PP^{g-1}$ a canonically embedded (and thus not hyper-elliptic) curve
of genus $g>2$, and suppose that there is a map from $C$ to $\PP^1$ of degree $d$. Equivalently, suppose that
$C$ has a line bundle $\sL$ of degree $d$ generated by 2 global sections. It follows from the Riemann-Roch theorem that
the complementary line bundle $\omega_C \otimes \sL^{-1}$ has at least $2+g-d-1$ sections. The construction of 
Section~\ref{syz and geom} shows that the homogeneous ideal $I_C$ of $C$ contains the ideal of $2\times 2$ minors of 
a $2\times 2+g-d-1$ 1-generic matrix $M$ of linear forms, and it follows that the Eagon-Northcott complex resolving the
minors of $M$ is a direct summand. Since $C$ is nondegenerate, the generators of this ideal can be taken to be 

\subsection{Low genus canonical embeddings} Schreyer's table (include $g=9$)?

\section{Cheerful facts about Stillman's problem}
%\section{Low degree}
\input{footer}
