\input{header}

\chapter{Embeddings of Hyperelliptic and Trigonal curves}
\label{ScrollsChapter}
\section{Hyperelliptic curves}
\fix{maybe start a new Chapter?}

In our early encounters with curves, we frequently assumed that the curve we were considering was non-hyperelliptic, since the behavior of hyperelliptic curves is so atypical. In this section, we'll describe the geometry of hyperelliptic curves.

\subsection{Basic models of hyperelliptic curves}\fix{move this section to ch 2; add discussion of adjoints--- perhaps as exercises?}

We start by establishing some basic facts about hyperelliptic curves. Many of these follow from general theorems like Riemann-Roch; but since they can be established by direct examination we will carry that out here.

Suppose $C$ is a smooth, projective hyperelliptic curve of genus $g \geq 2$. By definition, $C$ admits a degree 2 map $\pi : C \to \PP^1$; and as we've observed (\ref{**}) this map is unique.

By Riemann-Hurwitz, \fix{attibution?} the map $\pi : C \to \PP^1$ will have $2g+2$ distinct simple branch points, say $\lambda_1,\dots,\lambda_{2g-2} \in \PP^1$. An open subsect $C^\circ$ of $C$ can then be realized as the smooth projective completion of the affine curve given as
$$
C^\circ = \big\{ (x,y) \in \AA^2 \; \mid \; y^2 = \prod_{i=1}^{2g+2} (x - \lambda_i) \big\}.
$$ 
\fix{if two of the $\lambda_i$ coinncide, then the curve develops a singular point. Much of what we will do carries over to the singular case.} \fix{say the smooth model has 2 points at $\infty$.} Note that if we simply take the closure of this locus in $\PP^2$, the resulting curve will be highly singular at the point $[1,0,0]$, as can be seen either  directly by making an appropriate change of variables, or by invoking the genus formula for plane curves: if the closure were smooth, it would have genus $\binom{2g+1}{2}$. We can, however, complete the curve simply in $\PP^1 \times \PP^1$, for example by setting \fix{this is a rabbit from a hat. Consider either saying that by the previous section, if there's an emb in P3 then its on P1 x P1 as a divisor of type
2,g+1; and then "finding" this embedding as below; or moving this page to the early place where hyperelliptic curves are first mentioned.}
$$
y' = \frac{y}{\prod_{i=1}^{g+1} (x - \lambda_i)};
$$
we can then write the equation of a still smaller open subset of $C$ as
$$
{y'}^2 \cdot \prod_{i=1}^{g+1} (x - \lambda_i) \; = \; \prod_{i=g+2}^{2g+2} (x - \lambda_i).
$$
If we now take the closure of this locus in $\PP^1 \times \PP^1$, we get a curve of type $(2,g+1)$ on $\PP^1 \times \PP^1$; this curve is smooth, as can be seen again either directly in coordinates or by invoking the genus formula for curves on $\PP^1 \times \PP^1$. In other words,
$$
C \; = \; V\Big(Y_0^2\cdot \prod_{i=1}^{g+1} (X_1 - \lambda_iX_0) - Y_1^2 \cdot \prod_{i=g+2}^{2g+2} (X_1 - \lambda_iX_0) \Big)
$$

Next, let's describe the space of regular differentials on $C$. For this, it's convenient to work with the affine model $C^\circ = V(f) \subset \AA^2$, where
$$
f(x,y) = y^2 - \prod_{i=1}^{2g-2} (x - \lambda_i).
$$

We'll denote the two points at infinity---that is, the two points of $C \setminus C^\circ$---as $p$ and $q$.

To start, consider the simple differential $dx\in \Omega_{C^\circ/k}$. This is clearly regular on $C^\circ$, with zeros at the ramification points $r_i = (\lambda_i, 0)$. But it does not extend to a regular differential on all of $C$: it will have double poles at $p$ and $q$, as can be seen either directly or by degree considerations: as we said, $dx$ has $2g+2$ zeros, while the degree of $K_C$ is $2g-2$, meaning that there must be poles at the points $p$ and $q$.

To kill these poles, we can of course divide by $x^2$ (or any quadratic polynomial in $x$). But that just introduces new poles in the finite part $C^\circ$ of $C$. Instead, we want to multiply $dx$ by a rational function with zeros at $p$ and $q$, but \emph{whose poles occur only at the points where $dx$ has zeroes}---that is, the points $r_i$.  A natural choice is simply the reciprocal of the partial derivative $f_y = \partial f/ \partial y = 2y$, which vanishes exactly at the points $r_i$, and has correspondingly a pole of order $g+1$ at each of the points $p$ and $q$ (reason: the involution $y\to -y$ fixes $C^\circ$ and $x$), and exchanges the points $p,q$. In other words, the differential
$$
\omega = \frac{dx}{f_y}
$$
is regular, with divisor
$$
(\omega) = (g-1)p + (g-1)q.
$$
The remaining regular differentials on $C$ are now easy to find: Since $x$ has only a simple pole
at the two points at infinity \fix{say why}. we can  multiply $\omega$ by any $x^k$ with $k = 0, 1, \dots, g-1$. Since this gives us $g$ independent differentials, these  form a basis for $H^0(K_C)$.


 1) special linear series are mult $g^1_2$+basepoints. 2) Given an embedding, there's a union of lines. If the embedding is complete, we get a matrix...that defines the union of lines. Scrolls in all dimensions as unions of spans of divisors.
 
 
\subsection{General embeddings of degree
genus$+3$} 

It's a divisor on a quadric in $\PP^{3}$ of type $(2,g+1)$


\section{Trigonal curves}

\subsection{Special linear series on trigonal curves}

In analyzing special linear series on a hyperelliptic curve, we made crucial use of the facts that the canonical image of a hyperelliptic curve is a rational normal curve, and that any collection of points on a rational normal curve $C \subset \PP^n$ either are linearly independent or span $\PP^n$. In a similar (though necessarily less complete) way, we can use the fact that the canonical image of a trigonal curve lies on a rational normal surface scroll to describe special linear series on it.

 

\begin{lemma}
Let $S = S_{a,b} \subset \PP^n$ be a rational normal surface scroll. Any hyperplane section $H \cap S$ consists of the union of a rational normal curve $E$, which is a section of the scroll, and a union of lines of the ruling of the scroll.
\end{lemma}

Note that the curve $E$ must be a reduced component of $S \cap H$, but the lines $L_i$ may coincide, i.e., may be non-reduced components of the intersection. In the following proof, we'll assume for clarity that the lines $L_i$ are distinct (that is, $S \cap H$ is reduced); we leave it as an exercise to rewrite the proof to accommodate the remaining cases.

\begin{proof}
Let $F \in \Pic(S)$ be the class of a line of the ruling. Since $F^2 = 0$ and $H\cdot F = 1$, exactly one of the components of $S \cap H$ must have intersection number 1 with $F$; all other components must have intersection number 0 with $F$ and so must be lines of the ruling.

It remains to show that the unique component $E$ of $H \cap S$ having intersection number 1 with $F$ is a rational normal curve. This can be seen directly, but there's a shortcut. Suppose that we have
$$
S \cap H = E \cup L_1 + \dots + L_k,
$$
so that in particular $\deg(E) = n-1-k$. Since each of the lines $L_i$ of the ruling must meet $C$, we have that
\begin{align*}
n-1 &= \dim(\overline{S \cap H}) \\
&\leq \dim(\overline {E}) + k\\
&\leq (n-1-k) + k \\
&= n-1.
\end{align*}
We conclude that $\dim(\overline E) = n-k-1$, and hence that $E$ is a rational normal curve.
\end{proof}

Note that if $S = S_{a,b}$ with $a \leq b$, we must have either $0 \leq k \leq a$ or $k = b$: as soon as $k > a$, the span of the lines $L_i$ will contain the directrix of the scroll, and so must consist of the union of the directrix with $n-1-a = b$ lines.

Now let $C$ be a trigonal curve of genus $g \geq 5$, embedded in $\\P^{g-1}$ as a canonical curve, and let $S$ be the scroll containing $C$. We want to describe special linear series $\cD = |D|$. If our linear series has base points, we can delete them; so we'll assume that $|D|$ and $|K-D|$ are base point free. Note that this implies that   both  $r(D) \geq 1$ and $r(K-D) \geq 1$. In addition, it follows by Bertini that a general divisor $D \in \cD$ is reduced, that is, consists of distinct points $p_1,\dots,p_d$.

Now, the first hypothesis, that $r(D) \geq 1$, says that the points $p_1,\dots,p_d$ are linearly dependent. The second hypothesis, that $r(K-D) \geq 1$, says that the points $p_i$ span a subspace of codimension at least 2 in $\PP^{g-1}$ They therefore lie on at least a pencil of hyperplanes; let $H$ be a general hyperplane containing $D$.




. is effective, says that the divisor $D$ lies in a hyperplane section $C \cap H$; let $H$ be a general such hyperplane. At the same time 

canonical image lies on a 2-dim scroll (non -subcanonical embedding only on 3-dim scrolls).  embedding of a trigonal curve lies on the same scroll.Stratification of trigonal curves by Maroni invariants. Dimensions via automorphism groups of scrolls.

\section{Castelnuovo's Theorem}
(Statement only) \fix{we'll need the existence of smooth curves in given classes --- base point freeness of certain divisor classes on the scroll. Theorem: bpf iff they meet both a,b rational normal curves positively. reference to Montreal? better to make a tex file of the essential bit and put it in, as appendix. or ACGH?}

\input{footer}