\input{header}

\chapter{Inflection points and the Brill Noether Theorem}
\label{InflectionsChapter}


In this chapter, we will introduce the \emph{inflection points} of a linear system, and use them to give a proof of one implication in the statement of the Brill-Noether theorem.\fix{I changed from "classical B-N" -- ok?}

Just as a point $p \in C$ on a smooth plane curve $C \subset \PP^2$ is called a \emph{flex point} if there is a line $L \subset \PP^2$ having contact of order 3 or more with $C$ at $p$, a point on a smooth, nondegenerate curve $C \subset \PP^r$ will be called an \emph{inflection point} if there is a hyperplane $H \subset \PP^r$ having contact of order $r+1$ or more with $C$ at $p$. This notion can be extended to arbitrary linear series on smooth curves (as opposed to very ample ones); we'll see below that (in characteristic 0) every linear series has finitely many inflection points, and how to count them.


\section{Inflection points,  Pl\"ucker formulas and Weierstrass points}

\subsection{Definitions}
To define the inflection points of a linear series $\sD = (\sL, V)$ on a curve $C$, we will use the following result:

\begin{proposition} Let $V$ be a vector space of global sections of an invertible sheaf $\sL$ on a smooth curve $C$, and let $p \in C$ be a point. There exists a basis $\sigma_0, \dots, \sigma_r$ of $V$ consisting of sections vanishing to different orders at $p$. Thus the set
$$
\{ \ord_p(\sigma) \mid \sigma \neq 0 \in V \}
$$
 has cardinality $\dim V$
\end{proposition}

\begin{proof} Let $\tau_0, \dots, \tau_r$ be any basis of $V$. If  $\tau_i$ and $\tau_j$ vanish to the same order $a$ then as germs of sections of $\sL$ at $p$, they both represent nonzero classes in the one-dimensional vector space
$\frak{m}_{C,p}^a\sL/\frak{m}_{C,p}^{a+1}\sL$. \fix{what is our notation for the maximal ideal? What is our notation for the localization of $\sL$ at $p$?} Thus some nonzero linear combination $\tau_i' := a\tau_i+b\tau_j$  vanishes to strictly higher order. Since the coefficients $a$ and $b$ are both necessarily nonzero we may change bases, replacing $\tau_i$ with $\tau_i'$. As long as our basis contains elements vanishing to the same order we can repeat this process; since the order $\ord_p(\sigma)$ of any section at $p$ is bounded above by the degree of $\sL$, this process must terminate.
\end{proof}

Using the Proposition, we may write
$$
\{ \ord_p(\sigma) \mid \sigma \neq 0 \in V \} = \{a_0,\dots,a_r\} \; \text{ with } \; 0\leq a_0 < a_1 < \dots < a_r.
$$
The sequence $a_i = a_i(\cD,p)$ is called the \emph{vanishing sequence} of $\cD$ at $p$.  Since $a_i \geq i$, the numbers $\alpha_i = \alpha_i(\cD,p) = a_i - i$ are often more interesting, and the sequence $0 \leq \alpha_0 \leq \alpha_1 \leq \dots \leq \alpha_r$ is called the \emph{ramification sequence} of $\cD$ at $p$. 

We say that $p$ is an \emph{inflection point} of the linear series $\cD$ if $(\alpha_0,\dots,\alpha_r) \neq (0,\dots,0)$, and we define the weight of $p$ to be
$$
w(\cD, p) = \sum_{i=0}^r \alpha_i(\cD, p).
$$

If $\cD$ is very ample, so that it may be viewed at the linear series cut on $C$ by hyperplanes for some embedding $C \subset \PP^r$, then this coincides with the notion above: $p$ is an inflection point if $a_r > r$; that is, if there is a hyperplane $H \subset \PP^r$ having contact of order $r+1$ or more with $C$ at $p$.

The first two terms in the ramification sequence are particularly important: $\alpha_0(\cD, p)$, is nonzero if and only if $p$ is a base point of $\cD$; and $\alpha_1(\cD, p) = 0$ if and only if, in addition, the map $\phi_\cD$ is an immersion (that is, has nonzero derivative) at $p$.


\subsection{The Pl\"ucker formula}

In characteristic 0,a linear series on a smooth projective curve can have only finitely many inflection points, and in fact the sum of the weights of all the inflection points depends only on the genus of the curve and the degree of the the linear series. The result is called the Pl\"ucker Formula:

\begin{theorem}
 \begin{equation}\label{Plucker formula}
\sum_{p \in C} w(\cD, p) \; = \; (r+1)d + r(r+1)(g-1).
\end{equation}
\end{theorem}
For a proof, see for example~\cite{} (3264?). 

\begin{exercise}
Show that in case $r=1$ this is exactly the Riemann-Hurwitz formula for branched covers of $\PP^1$.
\end{exercise}

The result also holds in positive characteristic under the hypothesis that the number of inflection points is finite. It is hard even to imagine a plane curve in which every point is a flex!---but in positive characteristic there are such ``strange'' curves \cite[]{Hartshorne}.

The Pl\"ucker formula leaves many questions unanswered: we don't know, for example, what combinations of inflection points are possible on the canonical series. \fix{Original statement was stronger -- made me nervous.} One thing we do know is the behavior of the inflection points for a general linear series; we'll state this here and prove it as a corollary to the proof of the Brill-Noether theorem later in this chapter.

\begin{theorem}
If $C$ is a general curve of genus $g$, $L \in \pic^d(C)$ a general line bundle of degree $d$ with $h^0(L) = r+1$ and $V = H^0(L)$, then every inflection point of the linear series $\cD = (L, V)$ has weight 1 and hence ramification sequence $(0, \dots, 0, 1)$.
\end{theorem}

As a sample application of the Pl\"ucker formula, we can give a simple proof of Proposition~\ref{rnc unique uninflected}, which says that the rational normal curve is the only curve with no inflection points. 

\begin{proof}
Suppose $C \subset \PP^r$ is a curve of degree $d$ and genus $g$. If $C$ has no inflection points then, by the Pl\"ucker formula, we must have
$$
(r+1)d + r(r+1)(g-1) = 0.
$$
This immediately implies that $g=0$, so that we must have $(r+1)(d-r) = 0$ and hence $d=r$; thus $C$ is a rational normal curve.
\end{proof}

\subsection{Weierstrass points}

As with any extrinsic invariant of a curve in projective space, we can derive an intrinsic invariant of an abstract curve by applying the invariant to the canonical linear series. We define a \emph{Weierstrass point} of a curve $C$ to be an inflection point of the canonical linear series $|K_C|$. 

Thus $p$ is a Weierstrass point of $C$ if there exists a  differential form on $C$ vanishing to order $g$ or more at $p$. The \emph{weight} $w_p$ of a Weierstrass point $p \in C$  is defined to be the weight $w(|K_C|,p)$ of $p$ as an inflection point of the canonical series. 

The Pl\"ucker formula tells us  the total weight of the Weierstrass points on a given curve $C$:

\begin{corollary}\label{plucker formula}
The sum of the weights of the Weierstrass points on a curve $C$ of genus $g$ is
$$
\sum_{p \in C} w_p = g^3-g.
$$\qed
\end{corollary}

It will follow from the proof of the Brill-Noether theorem below that on a general curve $C$, every Weierstrass point has weight 1; thus there are $g^3-g$ distinct Weierstrass points on $C$

%Since there are only finitely many ramification points of the canonical series, a general point $p$ on any curve $C$ has gap sequence $(1,2,\dots,g)$, and correspondingly its Weierstrass semigroup
%is $W_p = (0, g+1, g+2, \dots)$. A Weierstrass point is called \emph{normal} if it has weight 1; this is tantamount to saying that the gap sequence is $(1,2,\dots,g-1,g+1)$, or that the semigroup is $(0, g, g+2, g+3, \dots)$. (The full Brill-Noether theorem tells us that a general curve $C$ has only normal Weierstrass points; this will be a consequence of Theorem~\ref{BN with inflection and dimension} below.)

\subsubsection{The Weierstrass semigroup} 

By definition, a point $p \in C$ on a curve $C$ of genus $g$ is a Weierstrass point iff $h^0(K_C(-gp)) \neq 0$; that is, iff there exists a holomorphic differential on $C$ vanishing to order at least $g$ at $p$. Now, the Riemann-Roch formula tells us that
$$
h^0(\cO_C(gp)) = g - g + 1 + h^0(K_C(-gp))
$$
so the condition $h^0(K_C(-gp)) \neq 0$ is equivalent to saying that $h^0(\cO_C(gp)) > 1$; in other words, there exists a nonconstant rational function on $C$, regular on $C \setminus \{p\}$ and having a pole of order at most $g$ at $p$.

This suggests that we look at the set of all possible orders of pole at $p$ of rational functions regular on $C \setminus \{p\}$; that is,
$$
W(C,p) := \left\{ \ord_p(f) \mid f \in K(C) \text{ with $f$ regular on } C \setminus \{p\} \right\}.
$$
This is clearly a sub-semigroup of the natural numbers $\NN$; it is called the \emph{Weierstrass semigroup} of the point $p$.  

Now, another way to characterize the condition that there exists a rational function on $C$, regular on $C \setminus \{p\}$, with a pole of order exactly $k$ at $p$ is to say that
$$
h^0(\cO_C(kp)) = h^0(\cO_C((k-1)p)) + 1.
$$
Applying Riemann-Roch to both sides of this equation, we see that it is equivalent to the condition
$$
h^0(K_C(-kp)) = h^0(K_C((-k+1)p)).
$$
This holds  if and only if there does \emph{not} exist a regular differential on $C$ with a zero of order exactly $k-1$ at $p$. In other words, the complement $\NN \setminus W(C,p)$ is exactly the vanishing sequence of the canonical series at $p$, shifted by 1; in particular, it has cardinality  exactly $g$. This is called the \emph{Weierstrass gap sequence} of the point $p$.

%
%By the Riemann-Roch theorem, $p\in C$ is a Weierstrass point if and only if $h^0(\cO_C(gp)) \geq 2$, or in other words to saying that there exists a non-constant rational function on $C$, regular away from $p$ and having a pole of order $g$ or less at $p$. The same argument shows more generally that for any $k \geq 0$, there exists a rational function on $C$, regular on $C \setminus \{p\}$ and having a pole of order exactly $k$ at $p$---that is,
%$$
%h^0(\cO_C(kp)) > h^0(\cO_C((k-1)p))
%$$
%if and only if 
%$$
%h^0(K_C(-kp)) = h^0(K_C((-k+1)p)).
%$$
%This holds  if and only if there does \emph{not} exist a regular differential on $C$ with a zero of order exactly $k-1$ at $p$. 
%
%In sum, for each point $p\in C$ there exist exactly $g$ values of $k$ such that there does \emph{not} exist a rational function on $C$ with a pole of order exactly $k$ at $p$; these are called the \emph{gap values} of the point $p \in C$, and by the above they comprise exactly the vanishing sequence of the canonical series $|K_C|$ at $p$, shifted by 1. Note that the set of $k$ such that there \emph{does} exist a rational function on $C$ with a pole of order exactly $k$ at $p$, regular on $C \setminus \{p\}$---forms a subsemigroup of $\NN$, called the \emph{Weierstrass semigroup} of $p \in C$. Its complement in $\NN$ is the set of $g$ gap values.  

There is still much we don't know about Weierstrass points in general. Most notably, we don't know what semigroups of finite index in $\NN$ occur as Weierstrass semigroups; an example of Buchweitz shows that not all semigroups occur, but there are also positive results, such as the statement ([EH]) that every semigroup of weight $w \leq g/2$ occurs, and its refinement and strengthening by Pflueger (\cite{**}).

\begin{exercise}
Buchweitz' example (the semigroup of $\sL^2$ has the wrong number of elements)
\end{exercise}

 



\section{Finiteness of the automorphism group}

As an application of just a rudimentary knowledge of Weierstrass points \fix{plus a bunch of other stuff!} we will deduce a fundamental fact: that the automorphism group of a curve of genus $g\geq 2$ is finite. The idea behind the argument is simple: because the Weierstrass points of a curve $C$ are intrinsically defined, \emph{any automorphism of $C$ must carry Weierstrass points to Weierstrass points}. Since there are only finitely many Weierstrass points, then, it will suffice to show that the subgroup of $Aut(C)$ of automorphisms of $C$ that fix all the Weierstrass points individually is finite. \fix{True, but this isn't exactly our path.}In fact, the following two lemmas establish a strong version of this:

\begin{lemma}
Let $C$ be a smooth projective curve of genus $g \geq 2$, and $f: C \to C$ an automorphism of $C$.
\begin{enumerate}
\item If $f$ has $2g+3$ or more distinct fixed points, then $f$ is the identity; and
\item If $f$ has $2g+2$ distinct fixed points, then either $f$ is the identity or $C$ is hyperelliptic and $f$ is the hyperelliptic involution.
\end{enumerate}
\end{lemma}

We give two arguments, one invoking the classical topology and applying the Lefschetz fixed point formula and the other more algebro-geometric.

\begin{proof}[Proof  using the classical topology]
For the first, we recall the definition of the \emph{Lefschetz number} of a map $f : M \to M$ of a compact oriented real $n$-manifold $M$. This is the alternating sum of the traces of the action of $f$ on $H^i(X,\CC)$:
$$
L(f) := \sum_{i=0}^n {\rm Trace}\left(f^* : H^i(X,\CC) \to H^i(X,\CC)\right).
$$
The Lefschetz fixed point formula then says that if $f$ has isolated fixed points, the number of those points, properly counted, is equal to $L(f)$.

The fixed points of a map $\phi$ of $C$ to $C$ are the intersection of the graph of $\phi$ with the diagonal in $C\times C$, and are thus a Zariski closed subset of $\phi$. It follows that any map other than the identity has isolated fixed points. Moreover, any complex analytic map is orientation-preserving \fix{somewhere we need to talk about the classical topology, and  orientation in particular} so each fixed point of an automorphism $f$ contributes positively to $L(f)$. In particular,  the number of distinct fixed points is at most $L(f)$.

Now suppose $f: C \to C$ is any automorphism. Of necessity, $f$ acts as the identity on $H^0(C, \CC)$ and $H^2(C, \CC)$, so if we want to bound $L(f)$ we just have to say something about the action of $f$ on $H^1(C,\CC)$. To do this, note that \fix{note that the reader might not know about the Hodge decomposition!} the action of $f$ on $H^1(C,\CC)$ respects the \emph{Hodge decomposition}
$$
H^1(C,\CC)  = H^0(K_C) \oplus H^1(\cO_C).
$$  
Moreover, the action of $f$ on $H^0(K_C)$ preserves the definite Hermitian inner product
$$
H(\eta, \phi) = \int_C \eta \wedge \overline \phi,
$$
and it follows that \emph{the eigenvalues of the action of $f$ on $H^0(K_C)$ are all complex numbers of absolute value 1}, and likewise for the action on $H^1(\cO_C)$ \fix{this all needs to be spelled out} . The absolute value of the trace of $f^*: H^1(C,\CC) \to H^1(C,\CC)$ is thus at most 2g, and hence
$$
L(f) \leq 2 + 2g,
$$
\fix{where did the 2 come from??} proving the stated inequality in general.

Finally, if we have equality then $f$ must act as $-1$ on $H^1(C,\CC)$, \fix{why? How about +1?} and it follows (again from Lefschetz) that $f^2$ is the identity; applying the Riemann-Hurwitz formula to the map from $C$ to the quotient $B = C/\langle f \rangle$ we may deduce that $B = \PP^1$, so $C$ is hyperelliptic and $f$ the hyperelliptic involution.
\end{proof}

An alternative argument showing that if $f$ is not the identity, then it has  for the lemma may be given using the intersection pairing on the surface $S = C \times C$ and applying the index theorem for surfaces.
 \fix{need to define N(S), state the index formula.}

\begin{proof}[Proof of 1. using intersection theory]
Let $\Delta$ and $\Gamma \subset S$ be the diagonal and the graph of $f$ respectively, and let $\Phi_1$ and $\Phi_2 \subset S$ be fibers of the two projection maps; let $\delta, \gamma, \varphi_1$ and $\varphi_2$ be the classes of these curves in the Neron-Severi group $N(S)$  of $S$. The number of fixed points of $f$ (counted with multiplicities) is the intersection number  $b = \delta \cdot \gamma$.

We know all the other pairwise intersection number of these classes: the ones involving $\varphi_1$ or $\varphi_2$ are obvious; we have
$$
\delta^2 = 2 - 2g
$$
and since the automorphism $id_C \times f : C\times C \to C \times C$ carries $\Delta$ to $\Gamma$, we see that $\gamma^2 = 2-2g$ as well.

We can now apply the index theorem for surfaces to deduce our inequality. To keep things relatively simple, let's introduce two new classes: set
$$
\delta' = \delta - \varphi_1 - \varphi_2 \quad \text{and} \quad \gamma' = \gamma - \varphi_1 - \varphi_2,
$$
so that $\delta'$ and $\gamma'$ are orthogonal to the class $\varphi_1 + \varphi_2$. Since $\varphi_1 + \varphi_2$ has positive self-intersection, the index theorem\fix{give the statement} tells us that the intersection pairing must be negative definite on the span $\langle \delta',\gamma' \rangle \subset N(S)$. In particular, the determinant of the intersection matrix
\begin{center}
\begin{tabular}{c|c|c}
& $\delta'$ &  $\gamma'$  \\
\hline
$\delta'$ & $-2g$ & $b-2$ \\
\hline
$\gamma'$ & $b-2$ & $-2g$ 
\end{tabular}
\end{center}
(where again $b = \gamma \cdot \delta$) is nonnegative, or equivalently, $b\leq 2g+2$.
\end{proof}

Having established an upper  bound on the number of fixed points an automorphism $f$ of $C$ (other than the identity) may have, it remains to find a lower bound on the number of distinct Weierstrass points; this is the content of the next lemma.


\begin{lemma}
If $C$ is a smooth projective curve of genus $g \geq 2$, then $C$ has at least $2g+2$ distinct Weierstrass points; and if it has exactly $2g+2$ Weierstrass points it is hyperelliptic.
\end{lemma}

\begin{proof}
Let $p \in C$ be any point, and $w_1=w_1(p),\dots,w_g = w_g(p)$ the ramification sequence of the canonical series $|K_C|$ at $p$. By definition, 
$$
h^0(K_C(-(w_i+i)p)) = g - i.
$$
Applying Clifford's theorem we have
$$
g-i \leq \frac{2g - 2 - w_i - i}{2} + 1;
$$
solving, we see that
$$
w_i \leq i
$$
and hence
$$
w_p \leq \binom{g}{2}
$$
where $w_p$ is the total weight of $p$ as a Weierstrass point. Since the total weight of the Weierstrass points on $C$ is $g^3-g$ by the Pl\"ucker formula~\ref{***}, we see that the number of distinct Weierstrass points must be at least
$$
\frac{g^3-g}{\binom{g}{2}} = 2g+2.
$$
Finally, by the strong form of Clifford's Theorem \ref{***}, equality here implies that the curve is hyperelliptic.\fix{still need to say why only the hyperelliptic involution fixes all the W-points.}
\end{proof}



\fix{DE revised to here May 22}


\section{Proof of (half of) the Brill-Noether theorem}

In its most basic form, the Brill-Noether theorem asserts for any $d, g$ and $r$ that
\begin{enumerate}
\item if $\rho(g,r,d) := g - (r+1)(g-d+r) \geq 0$, then every curve $C$ of genus $g$ possesses a $g^r_d$; and
\item if $\rho < 0$ then a general curve of genus $g$ does not possess a $g^r_d$.
\end{enumerate}

The first part, often called the ``existence half" of Brill-Noether, was originally proved by Kempf (\cite{}) and Kleiman-Laksov (\cite{}); both proofs relied on an application of the Thom-Porteous formula to a particular map of vector bundles on the Jacobian of $C$. An account of this argument may  be found in Appendix A of \cite{}. 

We will now apply the notion of inflection points, and the Pl\"ucker formula in particular, to deduce the second half of the statement above---the ``nonexistence half." We will then go back and deduce some of the other parts of the  full Brill-Noether statement from the same set-up.\fix{full B-N is your language; I don't think it's standard. I guess it will have been stated by now? so there should be a backward ref.}

Our proof will use a construction first suggested by Castelnuovo. His idea is ingenious: a ``general curve of genus $g$" may be a mysterious object for large $g$. But we can construct nodal curves of arithmetic genus $g$ readily: we just start with $\PP^1$, pick $2g$ distinct points $p_1,\dots,p_g, q_1,\dots,q_g \in \PP^1$, and identify each pair of points $p_i, q_i$ to form a $g$-nodal curve $C_0$. We will see below that there exists a family of smooth curves of genus $g$ specializing to such a curve, and we can analyze linear series on the general member of such a family by considering how they specialize to $C_0$.

Interestingly, Castelnuovo introduced this construction not to prove the Brill-Noether theorem---at the time, that was considered well-established, or at least not in need of further verification---but to answer an enumerative question: Castelnuovo asked, if a general curve of genus $g = 2k$ has a finite number of $g^1_{k+1}$s, what is the number? As we've seen, for genera $g = 2, 4$ and 6 the answers are 1, 2 and 5 respectively; Castelnuovo used his construction to find the number in general, as we'll describe below. It was Altman and Kleiman, in~\cite{}, who first proposed that Castelnuovo's construction could be used to prove the Brill-Noether theorem, and carried out the reduction given in Proposition~\ref{} below.

Before we go further: we spoke above of the curve obtained by identifying pairs of points on $\PP^1$; our first task is to say what we mean by this.

\subsubsection{Step 1: Constructing nodal curves}

Let $C$ be a smooth curve, and $p, q \in C$ two distinct smooth points of $C$. We would like to identify the two points $p$ and $q$ to form a nodal curve $C'$. To start, as a topological space, $C'$ will be simply the quotient of $C$ by the equivalence relation $p \sim q$; we'll denote by $r \in C'$ the point corresponding to the equivalence class $\{p, q\}$. Next, we have to specify the structure sheaf of $C'$, which is likewise straightforward: away from the point $r$, $C'$ will be isomorphic to $C \setminus \{p,q\}$; in other words, for a point $s \neq r \in C'$, we take the local ring $\cO_{s, C_0} = \cO_{s,C}$. As for the local ring $\cO_{r, C_0}$, we set
$$
\cO_{r, C_0} := \left\{ (f,g) \in \cO_{p,C} \oplus \cO_{q,C} \mid f(p) = g(q) \right\}.
$$

To see that this is indeed a scheme, we can make a construction. First, embed $C$ in a high-dimensional projective space $\PP^n$ by a linear series of degree $d > 2g+2$, and then project from a general point on the secant line $\overline{p,q}$. We claim that the image of the projection map is the curve described above; we leave the verification as the following exercise.

\begin{exercise}
\begin{enumerate}
\item Using the fact that $d > 2g+2$, show that no two secant or tangent lines to $C \subset \PP^n$ meet; and
\item Deduce from this that the image $C' = \pi(C)$ is the curve constructed above.
\end{enumerate}
\end{exercise}

In fact, an extension of this construction may be applied when $C$ is singular, as long as $p$ and $q$ are smooth points of $C$, so that in particular we can identify multiple pairs of smooth point $p_i, q_i \in C$. For example, if $p_1,\dots,p_g, q_1,\dots, q_g \in \PP^1$ are general points, we can form a curve $C_0$ of arithmetic genus $g$ by identifying each pair $p_i, q_i$ in this way; we may call the resulting curve the \emph{general $g$-nodal rational curve}.

\subsubsection{Step 2: Castelnuovo's specialization}

Next, Castelnuovo proposed analyzing a family of smooth curves specializing to a $g$-nodal one; in order to use this construction in a proof of Brill-Noether, we have to prove that such families exist. We'll state the lemma we need here:

\begin{lemma}\label{specialization to nodal curve}
Let $p_1,\dots,p_g, q_1,\dots, q_g \in \PP^1$ be distinct points, and $C_0$ the curve obtained by identifying $p_i$ with $q_i$ for $i = 1,\dots,g$. There exists a family of curves $\pi : \cC \to B$, where
\begin{enumerate}
\item $B$ is a smooth curve, with distinguished point $0 \in B$;
\item for all $b \neq 0 \in B$, the fiber $C_b = \pi^{-1}(b)$ is a smooth, projective curve of genus $g$;  and
\item the fiber over $0$ is the curve $C_0$.
\end{enumerate}
\end{lemma}

This lemma can be proved by deformation theory, as in Hartshorne (\cite{}) and Sernesi (\cite{}) \fix{or so I thought! I was unable to find this statement, or anything implying it, in either book. Do you have a better reference?}. Alternatively, it will follow from the description of the local geometry of Severi varieties, as worked out in Chapter~\ref{PlaneCurvesChapter}, and we'll defer the proof to that chapter.

%
%\begin{lemma}\label{BN in family}
%If $p_1,\dots,p_g, q_1,\dots, q_g \in \PP^1$ are general points and $\cC \to B$ is a family of curves as described in Lemma~\ref{specialization to nodal curve} above, then for general $b \in B$ the fiber $C_b$ does not possess a $g^r_d$ with $\rho < 0$.
%\end{lemma}
%
%From this, we deduce the basic
%
%\begin{theorem}\label{bare-bones BN}
%A general curve $C$ of genus $g$ does not possess a $g^r_d$ with $\rho(g,r,d) < 0$.
%\end{theorem}

\subsubsection{Step 3: The Altman-Kleiman analysis}

The analysis proposed by Altman and Kleiman considers the question: given a family $\pi : \cC \to B$, and a family $\{ \cD_b = (L_b, V_b)\}_{b \neq 0 \in B}$ of linear series on the curves $C_b$ with $b \neq 0$, how can we describe the ``limit" of the linear series $\cD_b$ as $b \to 0$? 

Suppose now we have a family $\pi : \cC \to B$ and in Lemma~\ref{specialization to nodal curve}, and suppose that the general curve $C_b$ in the family does have a line bundle  of degree $d$ with $r+1$ sections. The first thing to observe is that, possibly after a base change, we can pick out one such line bundle $\cL_b$ for each $b \neq 0$, varying regularly with $b$; or, in other words \emph{there exists a  line bundle $\cL^\circ$ on the complement $\cC^\circ$ of the special fiber such that}
$$
\deg(\cL^\circ|_{C_b}) = d \quad \text{and} \quad h^0(\cL^\circ|_{C_b}) \geq r+1
$$
for all $b \neq 0 \in B$. \fix{need to give argument for this assertion}



%The basic outline of the argument is by contradiction, but straightforward: we assume that the general curve $C_b$ in the family does have a $g^r_d$, consider what the limit of those $g^r_d$s might look like and, using our knowledge of the relatively simple curve $C_0$, arrive at a contradiction. By way of notation, let $B^\circ = B \setminus \{0\}$ and let $\cC^\circ = \pi^{-1}(B^\circ)$ be the complement in $\cC$ of the special fiber. The proof proceeds essentially in four/five steps.
%
%\
%
%\noindent {\bf Step 0: Existence of such a family}
%
%\begin{lemma}\label{cusp smoothing lemma}
%For each $g$, there exists a family $\cC \to B$ of curves with $B$ smooth and one-dimensional; $C_b$ a smooth curve of genus $g$ for $b \neq 0 \in B$ and $C_0$ a rational curve with $g$ cusps.
%\end{lemma}

%\
%
%\noindent {\bf Step 1: Finding a family of $g^r_d$s over $B^\circ$}
%
%Suppose now that the general curve $C_b$ in the family does have a line bundle  of degree $d$ with $r+1$ sections. The first thing to observe is that, possibly after a base change, we can pick out one such line bundle $\cL_b$ for each $b \neq 0$, varying regularly with $b$; or, in other words \emph{there exists a  line bundle $\cL^\circ$ on the complement $\cC^\circ$ of the special fiber such that}
%$$
%\deg(\cL^\circ|_{C_b}) = d \quad \text{and} \quad h^0(\cL^\circ|_{C_b}) \geq r+1
%$$
%for all $b \neq 0 \in B$. \fix{need to give argument for this assertion}

%\
%
%\noindent {\bf Step 2: Extending the line bundle $\cL^\circ$ to a sheaf on all of $\cC$}

Now, we want to extend $\cL^\circ$ to a sheaf on all of $\cC$. The key thing to observe is that we may not be able to extend $\cL^\circ$ to a line bundle on $\cC$. But we do have a positive result along these lines: we claim that \emph{there exists a torsion-free sheaf $\cL$ on all of $\cC$ such that $\cL|_{\cC^\circ} \cong \cL^\circ$}.

To see this, we choose an auxiliary line bundle $\cM$ on $\cC$ with relative degree $e > d + 2g$ (for example, embed $\cC$ in projective space and take $\cM = \cO_{\cC}(m)$ for large $m$); in keeping with our notational conventions, let $\cM^\circ$ be the restriction of $\cM$ to $\cL^\circ$. Consider the line bundle 
$$
\cN^\circ = (\cL^\circ)^* \otimes \cM^\circ.
$$
The bundle $\cN^\circ$ has lots of sections: the direct image is locally free of rank $e-g+1 > 0$, and after restricting to an open neighborhood of $0 \in B$ we can assume it's generated by them. Choose a section $\sigma$ of $\cN^\circ$; let $D^\circ \subset \cC^\circ$ be its zero divisor, and let $D \subset \cC$ be the closure of $D^\circ$ in $\cC$. Now, away from $C_0$ we can write
$$
\cL^\circ = (\cN^\circ)^* \otimes \cM^\circ = \cI_{D^\circ/\cC^\circ} \otimes \cM^\circ
$$
and accordingly the sheaf
$$
\cL := \cI_{D/\cC} \otimes \cM
$$
is the desired sheaf. Note that this need not be locally free: the total space $\cC$ of our family may not be smooth at the nodes of the special fiber $C_0$ (even if the family we originally started with had smooth total space, the base change called for in the first step would yield a family with  total space singular at the nodes of $C_0$), and if $D$ passes through any of these points it need not be Cartier.

In sum, if the general fiber $C_b$ of our family has a $g^r_d$, we can conclude that the special fiber $C_0$ has a torsion-free sheaf $\cL_0$ with 
$$
c_1(\cL_0) = d;
$$
and, by upper-semicontinuity of cohomology,
$$
h^0(\cL_0) \geq r+1.
$$

The next question is, what does $\cL_0$ look like if it's not locally free? Here we have a basic lemma:

\begin{lemma}\label{torsion free at node}
Let $p$ be a  node of a curve $C$. If $\cF$ is a torsion-free sheaf on $C$, then in a neighborhood of $p$ in $C$ the sheaf $\cF$ is either locally free or isomorphic to the ideal sheaf $\cI_{p/C}$ of $p$ in $C$.
\end{lemma}

\begin{proof}
Consider the endomorphism ring of $\cF$, and note that it is commutative and integral over $\cO_{p/C}$; thus it is either 
$\cO_{p/C}$ or its integral closure $\cO_{p/\tilde C}$. In the latter case it is  free over $\cO_{p/\tilde C}\cong \cI_{p/C}$.
In the former case....\fix{complete the argument}

\end{proof}

%\begin{exercise}
%\begin{enumerate}
%\item Show  that the conclusion of Lemma~\ref{torsion free at node} holds in case $p$ is a node of $C$
%\item Show by example that the conclusion of Lemma~\ref{torsion free at cusp} is false in case $p$ is either a tacnode or a triple point of $C$.
%\end{enumerate}
%\end{exercise}
%


\subsubsection{Step 4: The reduction to Schubert calculus}

The point of the argument thus far has been to reduce a problem involving linear series on a smooth curve---that is, a line bundle of degree $d$ with $r+1$ global sections---to one involving sections of a torsion-free sheaf on a $g$-nodal curve. Why is this an improvement? 

The answer is, if we have a linear system on a $g$-nodal curve $C_0$, we can look at its pullback to the normalization $\PP^1$ of $C_0$, where we really do understand the geometry of linear systems much better. To see how this goes, suppose that as above we have a family $\cC \to B$ of curves specializing from a smooth curve of genus $g$ to a $g$-nodal curve $C_0$; suppose that we have a linear series $\cD_b = (\cL_b, V_b)$ of degree $d$ and dimension $r$ on the smooth fibers $C_b$. In addition, suppose for the moment that the family of line bundles $\cL_b$ extends to a line bundle on all of $\cC$. (Of course, once we've done this we'll double back and consider what happens if the limit of $\cL_b$ is torsion-free but not locally free.) What has this reduction bought us?

Quite a lot, actually. The point is, if $\cL_0$ is a line bundle of degree $d$ on the $g$-nodal curve $C_0$, and $\nu : \PP^1 \to C_0$ the normalization map, the pullback $\nu^*\cL$ is simply the line bundle $\cO_{\PP^1}(d)$; and if $V_0 \subset H^0(\cL_0)$ is an $(r+1)$-dimensional vector space of sections of $\cL_0$, then the pullback $\nu^*(V_0)$ is just an $(r+1)$-dimensional subspace of $H^0(\cO_{\PP^1}(d))$. There are of course a lot of these---if we imagine $\PP^1$ as embedded in $\PP^d$ by the complete linear series $|\cO_{\PP^1}(d)|$, they correspond exactly to linear spaces $\Lambda \cong \PP^{d-r-1} \subset \PP^d$, so that we have a Grassmannian of them. The question is, \emph{when is the $g^r_d$ on $\PP^1$ associated to a linear space $\Lambda \subset \PP^d$ the pullback of a $g^r_d$ on $C_0$}?

The answer is straightforward: in order for a linear series on $\PP^1$ to be the pullback of a linear series on $C_0$, it has to be the case that for each $i = 1,\dots, g$, \emph{every divisor of the linear series containing $p_i$ must also contain $q_i$ and vice versa}. In terms of the geometry of the linear space $\Lambda \subset \PP^d$, this is tantamount to saying that \emph{$\Lambda$ must intersect the secant line $\overline{p_i,q_i}$ to the rational normal curve $\PP^1 \subset \PP^d$ for each $i=1,\dots,g$}.

Do we expect there to exist linear spaces $\Lambda \cong \PP^{d-r-1} \subset \PP^d$ meeting each of $g$ general chords to a rational normal curve? To answer this, we can make a dimension count. To start, the Grassmannian $\GG(d-r-1,d)$ has dimension $(r+1)(d-r)$. Next, for a given line $L \subset \PP^d$, the locus of $(d-r-1)$-planes $\Lambda \subset \PP^d$ meeting $L$ is a Schubert cycle, denoted $\Sigma_r(L)$; it has codimension $r$ in $\GG(d-r-1,d)$. If in fact the Schubert cycles $\Sigma_r(\overline{p_i,q_i})$ associated to $g$ general chords to the rational normal curve in $\PP^d$ intersect properly, or dimensionally transversely, this will be the case only if
$$
rg \leq (r+1)(d-r),
$$
which is exactly the condition $\rho(g,r,d) \geq 0$! We must be on the right track.

While we're at it, let's ask what would happen if the limit $\cL_0$ of the line bundles $\cL_b$ in our family were in fact not locally free at some subset $r_1, \dots, r_\delta$ of the nodes of $C_0$? In this case, let $\tilde C_0$ be the partial normalization of $C_0$ at these nodes, and $\tilde \nu : \tilde C_0 \to C_0$ the partial normalization map, so that $\tilde C_0$ is the ($g-\delta$)-nodal curve obtained by identifying $g-\delta$ pairs of general points on $\PP^1$. The pullback $\tilde \nu^*(\cL_0)$ is then locally free of degree $d-2\delta$ on $\tilde C_0$, so that an $(r+1)$-dimensional space of sections of $\cL_0$ will correspond to a plane $\Lambda \cong \PP^{d-2\delta-r-1} \subset \PP^{d-2\delta}$, meeting each of $g-\delta$ general chords to a rational normal curve in $\PP^{d-2\delta}$. Again, if we knew that the corresponding Schubert cycles intersected properly, the existence of such planes would imply that
$$
r(g-\delta) \leq \dim\GG(d-2\delta-r-1,d-2\delta) = (r+1)(d-r-2\delta),
$$
which amounts to the inequality $\rho \geq \delta \geq 0$. 

In sum, we have established the 

\begin{theorem}[Altman-Kleiman]
Assuming that the Schubert cycles $\Sigma_r(\overline{p_i,q_i})$ associated to general chords to a rational normal curve intersect properly, a general curve of genus $g$ will possess no linear series with $\rho < 0$.
\end{theorem}

In other words, we have reduced the nonexistence half of classical Brill-Noether to an assertion about chords to a rational normal curve!

\

\noindent {\bf Step 5: Applying the Pl\"ucker formula}

Specialization is one the core techniques of algebraic geometry, especially if we're trying to prove a statement about ``general" objects; the goal is exactly to get rid of the hypothesis of generality. In the present context, we want to prove that the Schubert cycles $\Sigma_r(\overline{p_i,q_i})$ associated to general chords to a rational normal curve intersect properly; and the first question we might ask is, is the hypothesis of generality necessary?

The answer, unfortunately, is ``yes:" there do exist collections of chords to a rational normal curve such that the corresponding Schubert cycles fail to intersect properly. A simple but very instructive example would be the case $g=3$, $d=2$ and $r=1$ (corresponding to the assertion that ``a general curve of genus 3 is not hyperelliptic"). Here, we are asking simply if three chords to a plane conic in $\PP^2$ are concurrent. Very clearly, three \emph{general} chords to a conic are not concurrent, but what is concerning is that there do exist triple of chords that are, meaning that the hypothesis of generality is still necessary.

(picture here)

At this point, though, we are saved by a remarkable observation: \emph{three secant lines to a conic may be concurrent, but three tangent lines can never be}! There are many ways to see this, but the one most relevant to our present circumstances is: if the tangent lines to a plane conic $C \subset \PP^2$ met at a point $r \in \PP^2$, then the projection map $\pi_r : C \to \PP^1$ would be a degree 2 map with three or more ramification points, a violation of the Riemann-Hurwitz formula.

This suggests both the statement we should be proving, and how to prove it. We will establish the key

\begin{lemma}
Let $C \cong \PP^1 \subset \PP^d$ be a rational normal curve, and $p_1,\dots,p_g \in C$ any $g$ points of $\PP^1$. If $L_i = \overline{2p_i}$ is the tangent line to $C$ at $p_i$, then the Schubert cycles $\Sigma_r(L_i)$ intersect properly; in particular, if $rg > (r+1)(d-r)$, then the intersection $\cap \Sigma_r(L_i)$ is empty.
\end{lemma}

Note that since tangent lines to a curve are specializations of secant lines, this implies that for $g$ general chords $L_i$ to a rational normal curve, the corresponding Schubert cycles $\Sigma_r(L_i)$ will intersect properly. Thus, once we establish the lemma we will have completed the proof of the nonexistence half of Brill Noether.

\begin{proof}
We will just prove the ``in particular" part here; the more general statement will follow, as we'll indicate in Exercise~\ref{} below.

The proof is head-slappingly easy, and indeed is already suggested by the simple example above. Suppose $\Lambda \cong \PP^{d-r-1} \subset \PP^d$ is a linear space meeting each of the tangent lines $L_i$, and consider the linear series cut on $\PP^1$ by hyperplanes containing $\Lambda$. The condition that $\Lambda \cap \overline{2p_i} \neq \emptyset$ means exactly that no hyperplane containing $\Lambda$ can be transverse to $\PP^1$ at $p_i$, or in other words that the linear series $\cD$ cut on $\PP^1$ by hyperplanes containing $\Lambda$ will have ramification sequence at least $(0, 1, 1,\dots,1)$ at $p_i$. Thus the sum of the weights of $\cD$ at the points $p_1,\dots,p_g$ is at least $rg$, and applying the Pl\"ucker formula we arrive at the inequality
$$
rg \leq (r+1)(d-r).
$$
\end{proof}


%
%Now, back to our family $\pi : \cC \to B$ of curves. We have assumed that for some $d$ and $r$ with $\rho(g,r,d) < 0$ the general curve $C_b$ has a $g^r_d$, and deduced that the special fiber $C_0$ has a rank 1 torsion-free sheaf $\cL_0$ of degree $d$ with at least $r+1$ sections; we now have to derive from this a contradiction.
%
%To see most clearly where this contradiction comes from, let's start with the simplest case: where $\cL_0$ is indeed locally free. In this case, let $\nu :  C^\nu \cong \PP^1 \to C$ be the normalization of $C$ and let $q_1,\dots, q_g \in \PP^1$ be the points lying over the cusps of $C_0$. We have
%$$
%\nu^*(\cL) \cong \cO_{\PP^1}(d)
%$$  
%and 
%$$
%V = \nu^*(H^0(\cL_0)) \subset H^0(\cO_{\PP^1}(d))
%$$
%is an $(r+1)$-dimensional space of sections. (If $H^0(\cL_0) > r+1$, just choose any $(r+1)$-dimensional subspace.) 
%
%Now, given that any section $\sigma \in V \subset H^0(\cO_{\PP^1}(d))$ is pulled back from the cuspidal curve $C$, we see that \emph{$\sigma$ cannot vanish to order exactly 1 at the point $q_i \in \PP^1$ lying over any of the cusps of $C_0$}. It follows that for each $i$ the ramification index 
%$$
%\alpha_1(V,q_i) \geq 1 
%$$
%and hence in  general $\alpha_1(q_i,V) \geq 1$ for all $i \geq 1$. In particular, the weight of the inflection point $q_i$ for the linear series $V$ satisfies
%$$
%w(V, q_i) \geq r
%$$
%and correspondingly
%$$
%\sum_{i=1}^g w(V, q_i) \geq rg
%$$
%But the Pl\"ucker formula~\ref{} tells us that the total weight of all inflection points for the series $V$ is
%$$
%\sum_{p \in \PP^1} w(V,p) = (r+1)(d-r)
%$$
%and there's our contradiction: by the hypothesis that 
%$$
%\rho(g,r,d) := g - (r+1)(g-d+r) < 0
%$$
%we have $rg > (r+1)(d-r)$.
%
%Finally, the case where $\cL_0$ is not locally free is if anything even easier. Suppose now that the sheaf $\cL$ fails to be locally free at $l$ of the cusps of $C_0$, say $\nu(p_1),\dots,\nu(p_l)$. Again, we can pull $\cL$ back to $\PP^1$; again we have
%$$
%\nu^*(\cL_0) \cong \cO_{\PP^1}(d);
%$$  
%and again we pull back section of $\cL$ to arrive at a linear system
%$$
%V = \nu^*(H^0(\cL_0)) \subset H^0(\cO_{\PP^1}(d))
%$$
%of degree $d$ and genus $g$ on $\PP^1$. The only difference here is that sections of $V$ all vanish at $p_1,\dots,p_l$, so that we have
%$$
%w(V,p_k) \geq 
%\begin{cases}
%r+1 &\text{ if } k \leq l; \text{ and} \\
%r  &\text{ if } k > l.
%\end{cases}
%$$
%so that
%$$
%\sum_{k=1}^g w(V, p_k) \geq rg + l
%$$
%and our contradiction is even more of a contradiction!

\section{Corollaries and extensions of our proof}

We have proved a bare-bones version of the nonexistence half of Brill-Noether; in particular, the argument does not a priori tell us anything about the geometry of $W^r_d(C)$ for a general curve $C$, or the geometry of $C$ as embedded by a general linear system. In the final section of this chapter, we'll see how we can deduce stronger forms of Brill-Noether from the basic set-up above.

\subsection{Brill-Noether with inflection}

Let's start with the low-hanging fruit. Since the basic ingredient of the proof of Theorem~\ref{bare-bones BN} is the Pl\"ucker formula, the argument tells us something about the inflection points of linear series on a general curve $C$ as well.

We start with a definition.

\begin{definition}
Let $C$ be a smooth curve of genus $g$ and $p_1,\dots,p_n \in C$ distinct points of $C$. If $\cD = (L,V)$ is a linear system on $C$ of degree $d$ and dimension $r$, we define the \emph{adjusted Brill-Noether number} of $\cD$ relative to the points $p_k$ to be
$$
\rho(\cD; p_1,\dots,p_k) := g - (r+1)(g-d+r) - \sum_{k=1}^n w(\cD,p_k).
$$
\end{definition}

In these terms, we have the basic (but powerful) extension of the Brill-Noether theorem proved above:

\begin{theorem}\label{Brill-Noether with inflection}
Let $(C;p_1,\dots,p_n)$ be a general $n$-pointed curve of genus $g$ (that is, let $C$ be a general curve and $p_1,\dots,p_n \in C$ general points; equivalently, let $(C;p_1,\dots,p_n)$ correspond to a general point of $M_{g,n}$). If $\cD$ is any linear system on $C$, then
$$
\rho(\cD; p_1,\dots,p_k) \geq 0.
$$
\end{theorem}

\begin{proof}
The proof is just an extension of the argument for Lemma~\ref{BN in family}. To start, let $\cC \to B$ be a family of curves as in the proof of Lemma~\ref{cusp smoothing lemma}. Let $\sigma_1, \dots, \sigma_n : B \to \cC$ be sections of $\cC \to B$ with $\sigma_k(0)$ a smooth point of $C_0$ for all $k$ (such sections can always be found after passing to an \'etale open neighborhood of $0 \in B$). Exactly as in the proof of Lemma~\ref{BN in family}, if the general curve $C_b$ in our family admits a $g^r_d$ $\cD$ with
$$
\rho(\cD;\sigma_1(b),\dots,\sigma_n(b)) < 0
$$
we can choose a family $\{\cD_b\}$ of such linear series on the fibers $C_b$ for $b \neq 0$ and, taking limits, we arrive at a $g^r_d$ $\cD_0$ on $\PP^1$ with
$$
w(\cD_0, q_i) \geq r
$$
for each of the $g$ points $q_i \in \PP^1$ lying over the cusps of $C_0$, and in addition
$$
w(\cD_0, r_k) \geq w(\cD_b,\sigma_k(b))
$$
where $r_k \in \PP^1$ is the point in $\PP^1$ lying over $\sigma_k(0) \in C_0$. Adding up, we have
\begin{align*}
\sum_{i=1}^g w(\cD_0, q_i) + \sum_{k=1}^n w(\cD_0, r_i) &\geq rg + \sum_{k=1}^n w(\cD_b,\sigma_k(b)) \\
&> rg + g - (r+1)(g-d+r) = (r+1)(d-r)
\end{align*}
since we assumed that 
$$
\rho(\cD_b;\sigma_1(b),\dots,\sigma_n(b)) = g - (r+1)(g-d+r) - \sum_{k=1}^n w(\cD_b,\sigma_k(b)) < 0.
$$
But as before the Pl\"ucker formula for $\PP^1$ tells us that
$$
\sum_{p \in \PP^1} w(\cD_0, p) = (r+1)(d-r),
$$
a contradiction.
\end{proof}

\subsection{Brill-Noether with dimension}

Theorem~\ref{Brill-Noether with inflection} might at first glance seem relevant only to problems involving inflection, but in fact in can be used to prove results that have nothing to do with inflection points. For example, one consequence is the stronger form of Brill-Noether:

\begin{theorem}\label{BN with dimension}
If $C$ is a general curve of genus $g$, then for any $d$ and $r$ with $\rho(g,r,d) \geq 0$,
$$
\dim W^r_d(C) = \rho(g,r,d).
$$
\end{theorem}

\begin{proof}
The basic idea of the proof is simple: basically, we argue that if we had a $(\rho+1)$-dimensional family of $g^r_ds$ on $C$, then we could find one with nonzero ramification at $\rho+1$ general points of $C$, violating Theorem~\ref{Brill-Noether with inflection}.

This idea is easier to implement after specializing, so once more we go back to our family $\cC \to B$ of smooth curves specializing to a $g$-cuspidal curve $C_0$, with normalization $\PP^1$. The basic lemma is:

\begin{lemma}\label{forced ramification}
Let $\Sigma$ be a complete curve and let $\{ \cD_\lambda \}_{\lambda \in \Sigma}$ be a (nonconstant) family of $g^r_ds$ on $\PP^1$ parametrized by $\Sigma$. If $p \in \PP^1$ is any fixed point, then for at least one $\lambda \in \Sigma$ we have $w(\cD_\lambda,p)>0$.
\end{lemma}

\begin{proof}
Embed $\PP^1$ in $\PP^d$ as a rational normal curve of degree $d$. Given a $(d-r-1)$-plane $\Lambda \subset \PP^d$, the hyperplanes in $\PP^d$ containing $\Lambda$ cut out a $g^r_d$ $\cD_\Lambda$ on $\PP^1$, and indeed every $g^r_d$ on $\PP^1$ can be described in this way for a unique $\Lambda$. The $g^r_d$s on $\PP^1$ are thus parametrized by the Grassmannian $\GG(d-r-1,d)$, and we can think of $\Sigma$ as a complete curve in $\GG(d-r-1,d)$.

Consider now the hyperplanes $H \subset \PP^d$ such that the divisor $H \cap \PP^1$ has multiplicity $\geq r+1$ at $p$. These correspond to points in a linear space of codimension $r+1$ in $(\PP^d)^*$; in particular, their intersection is an $r$-plane $\Omega \subset \PP^d$, called the \emph{osculating plane} to the rational normal curve at $p$. The condition that a  $g^r_d$ $\cD_\Lambda$ have non-zero ramification at $p$---in other words, that $\cD_\Lambda$ contains a divisor with multiplicity $\geq r+1$ at $p$---is simply that $\Lambda \cap \Omega \neq \emptyset$. But the set of such $\Lambda$ is a hyperplane section of $\GG(d-r-1,d)$ under trhe Pl\"ucker embedding; in particular, any complete curve $\Sigma \subset \GG(d-r-1,d)$ must intersect it.
\end{proof}

Given this lemma, the proof of Theorem~\ref{BN with dimension} proceeds as follows. We know from the basic dimension estimates of Chapter~\ref{} that $\dim W^r_d(C) \geq \rho(g,r,d)$ for any $C$; we have to show that we cannot have $\dim W^r_d(C) > \rho(g,r,d)$ for a general curve $C$. We argue as follows:

First: if it were the case that $\dim W^r_d(C) > \rho(g,r,d)$ for a general curve $C$, we would have, after specializing and pulling back to $\PP^1$, at least a $(\rho + 1)$-dimensional family of $g^r_d$s on $\PP^1$, all of which had ramification weight at least $r$ at the points $q_i$ of $\PP^1$ lying over the cusps of $C_0$. 

Secondly, we pick any $\rho + 1$ points $r_k \in \PP^1$ other than the $q_i$. Applying Lemma~\ref{forced ramification} repeatedly, we find that there is at least a $\rho$-dimensional subfamily of $g^r_d$s having nonzero ramification at $p_1$, a $(\rho - 1)$-dimensional subfamily of $g^r_d$s having nonzero ramification at $p_1$ and $p_2$, and so on; ultimately, we conclude that there is a $g^r_d$ $\cD$ on $\PP^1$ with ramification index at least $r$ at each $q_i$ and nonzero ramification index at each $r_k$. 

Finally, we observe that the linear series $\cD$ has total ramification at least
$$
rg + \rho + 1 = (r+1)(d-r)+1
$$
at the points $q_i$ and $r_k$, once more violating the Pl\"ucker formula.
\end{proof}

We can combine Theorem~\ref{BN with dimension} and Theorem~\ref{Brill-Noether with inflection} into one theorem, more complicated but more inclusive:

\begin{theorem}\label{BN with inflection and dimension}
Let $C$ be a smooth curve of genus $g$ and $p_1,\dots,p_n \in C$ distinct points; for $k = 1,\dots,n$ let $\alpha^k = (\alpha^k_0,\dots\alpha^k_r)$ be a nondecreasing sequence of nonnegative integers, and let
$$
G^r_d(p_1,\dots,p_n; \alpha^1,\dots,\alpha^n) = \{\cD \in G^r_d(D) \mid \alpha_i(\cD, p_k) \geq \alpha^k_i \}.
$$
If $(C, p_1,\dots,p_n)$ is a general $n$-pointed curve, then either $G^r_d(p_1,\dots,p_n; \alpha^1,\dots,\alpha^n)$ is empty or
$$
\dim G^r_d(p_1,\dots,p_n; \alpha^1,\dots,\alpha^n) = \rho(g,r,d) - \sum_{k+1}^n \sum_{i=0}^r \alpha^k_i.
$$
\end{theorem}

Finally, we can combine this last theorem with a little dimension-counting to deduce a simple fact:

\begin{theorem}
If $\cD$ is a general $g^r_d$ on a general curve, then $\cD$ has only simple ramification; that is,
$$
w(\cD, p) \leq 1 \quad \text{for all } p \in C.
$$
\end{theorem}

Note that applying this in case $d=2g-2$ and $r = g-1$, we arrive at the statement made earlier: that a general curve $C$ of genus $g$ has only normal Weierstrass points!

\input{footer}