\input{header.tex}


\chapter{Uniform position and its applications}\label{uniform position}

In this chapter, we will introduce an elementary but surprisingly useful tool for studying curves (and higher-dimensional varieties as well!). We'll see how to apply it in several contexts: to prove the strong form of Clifford's theorem, the strong form of Martens' theorem, and the Castelnuovo bound on the genus of a curve in projective space.

The basic Uniform Position lemma deals with the \emph{monodromy group} of the points of a general hyperplane section of a curve $C \subset \PP^r$. We'll define this precisely below, but the basic idea is straightforward if one is willing to adopt the classical topology (an equivalent definition is described in Cheerful Fact~\ref{Galois equals monodromy} below). To set it up, suppose $C \subset \PP^r$ is an irreducible curve, and $H_0 \subset \PP^r$ a hyperplane transverse to $C$; say the intersection $C \cap H_0 = \{p_1,\dots,p_d\}$. As we vary $H_0$ continuously along a real arc $\{H_t\}$, staying within the open subset $U \subset {\PP^r}^*$ of hyperplanes transverse to $C$, we can ``follow" each of the points $p_i(t)$ of intersection of $C$ with the hyperplane $H_t$.

Now imagine that the hyperplanes $H_t$ come back to the original $H_0$ at some time; that is, we have a continuous family $\{H_t\}_{0 \leq t \leq 1}$ with $H_1 = H_0$. Again, we can follow each of the points $p_i$ as $t$ varies, arriving at an arc $\{p_i(t) \in C \cap H_t\}_{0 \leq t \leq 1}$. Since $H_1 = H_0$, the end point $p_i(1)$ will be one of the original points $p_j \in C \cap H_0$ we started with, but it need not be the point $p_i$ we started with! In this way, we get a permutation of the set $C \cap H_0$; the group of all permutations arrived at in this way will be called the \emph{monodromy group} of the points $C \cap H_0$. 

This may seem a rather esoteric construction, but as we said it is surprisingly useful. We'll start by giving a precise definition of the mondromy group in a more general setting, then prove the \emph{uniform position lemma} for the points of a general hyperplane section of a curve; the rest of the chapter will be a series of applications of this lemma.

\begin{exercise}
(To go later, obviously) Let $C \subset \PP^r$ be any curve, not assumed irreducible; let $C = \cup C_i$ be its irreducible decomposition, and suppose that $C_i$ has degree $d_i$. Prove that the monodromy group of the points of a general hyperplane section of $C$ is the product $\prod S_{d_i}$.
\end{exercise}

\section{Uniform position} \label{uniformSection}

\subsection{The monodromy group of a generically finite morphism}

Let $f : Y \to X$ be a dominant map between varieties of the same dimension over $\CC$, and suppose that $X$ is irreducible. There is then an open subset $U \subset X$ such that $U$ and 
its preimage $V = f^{-1}(U)$ are smooth, and the restriction of $f$ to $V$ is a covering space in the classical topology. Let $d$ be the number of sheets. This is the degree of the extension $K(Y)/K(X)$.

A basic construction from homotopy theory now associates a monodromy group to any finite topological covering map: Choose a base point $p_0 \in U \subset X$, and suppose $\Gamma := f^{-1}(p_0)  = \{q_1,\dots,q_d\}$. If $\gamma$ is any loop in $U$ with base point $p_0$, for any $i = 1, \dots, d$ there is a unique lifting of $\gamma$ to an arc $\tilde \gamma_i$ in $V$ with initial point $\tilde \gamma_i(0) = q_i$ and end point $\tilde \gamma_i(1) = q_j$ for some $j \in \{1,2,\dots,d\}$. Since we could traverse the loop in the opposite direction, the index $j$ determines $i$, and the map $i\mapsto j$ is a permutation of $\{1,2,\dots,d\}$. 
Since the set $\Gamma$ is discreet, the permutation depends only on the class of $\gamma$ in $\pi_1(U,p_0)$ so we have defined a homomorphism to the symmetric group:
$$
\pi_1(U,p_0)  \to {\rm Perm}(\Gamma) \cong S_d.
$$
The image $M$ of this map is called the \emph{monodromy group} of the map $f$. It depends on the labeling of the points of $\Gamma$, but is well-defined  up to conjugation. Moreover, it is independent of the choice of open set $U$: if $U' \subset U$ is a Zariski open subset, the map $\pi_1(U', p_0) \to \pi_1(U,p_0)$ is surjective,  so the image of $\pi_1(U', p_0)$ in $S_d$ is the same.

\begin{fact}\label{Galois equals monodromy}
The theory of finite coverings of algebraic varieties is not only analogous to Galois theory, it \emph{is} Galois theory: In the situation described above, if we assume that $Y$ is irreducible, then the pullback map $f^*$ expresses the function field $K(Y)$ as a finite algebraic extension of $K(X)$. The degree of this extension is the degree of the map $f$. The monodromy group of $f$  the Galois group of the Galois normalization of $K(Y)$ over $K(X)$. (see \cite{Harris1979}.)
\end{fact}

Since we assumed that $X$ is irreducible, the space $U$ is (path) connected, and it follows that the monodromy group is transitive if and only if the space $V$ is (path) connected. But $V$ is a smooth
variety, so this is the case if and only if $V$ is irreducible.

We will next compute the monodromy group of the  universal hyperplane section, constructed as follows:
Let $C \subset \PP^r$ be an irreducible, nondegenerate curve of degree $d$, and let $X = {\PP^r}^*$ be the space of hyperplanes in $\PP^r$. We define the \emph{universal hyperplane section of $C$} to be the projection to $f: Y\to {\PP^r}^*$ of the incidence variety
$$
Y = \{ (H, p) \in {\PP^r}^* \times C \mid p \in H \}.
$$
The fibers of $f$ are the hyperplane
sections of $C$, so it is a dominant finite map. If we let $U\subset {\PP^r}^*$ be the open subset of hyperplanes
meeting $C$ transversely, then the restriction of $f$ to the preimage $V$ of $U$ is a covering space
whose fibers each consist of $d$ distinct points. The preimage in $Y$ of a point $p\in C$ is the set of hyperplanes containing
$p$, a copy of $\PP^{r-1}$, and thus $Y$ is irreducible. Thus the monodromy group of $f$ is transitive. But much more is true:

\begin{theorem}[Uniform Position Theorem]\label{uniform position lemma}
The monodromy group of the universal hyperplane section of an irreducible curve $C \subset \PP^r_\CC$ is the full symmetric group $S_d$.
\end{theorem}

Informally, this theorem says that two subsets of the same cardinality in the general hyperplane section of $C$
are indistinguishable from the point of view of any discrete invariant that is semicontinuous in the Zariski topology. 
The Uniform position result fails over fields of finite characteristic; see \cite{Rathmann} and \cite{Kadets} for what is known. 

The following is a good example of how the theorem is applied:

\begin{corollary}[numerical uniform position lemma]\label{numerical uniform position lemma}
Let $C\subset \PP^r$ be an irreducible curve, and let $H\cap C$ be a general hyperplane section. Any two subsets of $H\cap C$ with the same cardinality impose the same number of conditions on forms of any degree; that is, any two subsets of the same cardinality have the same Hilbert function.
\end{corollary}

\begin{proof} Let $U = {\PP^r}^* \setminus C^*$ be the open subset of hyperplanes transverse to $C$, and let $Y\to U$ be the universal hyperplane section.
Thus
\begin{align*}
 Y^{k*} :=&Y\times_U Y\times_U \cdots Y\times_U \setminus \Delta  =\\
                                                 & \{ (H, \Gamma) \in {\PP^r}^* \times C^k \mid \Gamma \subset H, \hbox{ and $\Gamma$ consists of $k$ distinct points.}\}\\
                                                 &\to U
\end{align*}
is the family subsets $\Gamma$ of cardinality $k$ in hyperplane sections $H \cap C$ of $C$. Because the monodromy is of the universal hyperplane section of $C$ is
$k$-times transitive, $Y_k^*$ is irreducible.

For each $m$ the number of conditions that $\Gamma$ imposes on forms of degree $m$ is lower semicontinuous, so it achieves its maximum on a Zariski open subset of $Y^{k*}$. Since $Y^{k*}$ is irreducible, the complement of this open set has dimension strictly less than $\dim Y^{k*} = r$. Thus a general hyperplane $H \in {\PP^r}^*$ will lie outside the image of $Y^{k*}$, meaning that the number of conditions imposed by all the $k$-element subsets $\Gamma \subset C \cap H$ will have this maximal value.
\end{proof}

Lemma~\ref{general position lemma} follows:

\begin{proof}[Proof of Lemma~\ref{general position lemma}] As in the proof above, let $Y^{k*}$ be the family of $k$-element subsets
of transverse hyperplane sections of $C$, and suppose that $k$ is maximal such that for general hyperplanes $H$ the
$k$-element subsets of $H\cap \Gamma$ imposes indedpendent conditions on linear forms, or in other words, is linearly independent.
It follows from Corollary~\ref{numerical uniform position lemma} that all subsets of $k+1$ elements are dependent, so $H\cap C$ is
contained in the span of a single $k$-element subset. Since $C$ is nondegenerate, a general hyperplane is spanned
by its intersection with $C$, so $k=r+1$.
\end{proof}

Since the symmetric group $S_d$ is generated by transpositions, we may prove that a subgroup $M\subset S_d$ is equal to $S_d$ by showing that $M$ is doubly transitive and contains a transposition (of course it will then follow that $M$ is $k$-fold transitive for any $k\leq d$.) We can translate these properties into geometry as follows:

\begin{lemma}\label{transitivity lemma}
Let $f : Y \to X$ be a generically finite cover of degree $d$, with  monodromy group $M \subset S_d$.
The $k$-fold fiber product $\prod_1^k(Y/X)$ is irreducible if and only if $M$ is $k$ times transitive.
\end{lemma}

\begin{proof}
The monodromy of the covering $f^k: \times^k(Y/X) \to X$ is the the diagonal action of
$M$ in the product $\prod_1^k S_d$.
\end{proof}

Thus a consequence of the Uniform Position Theorem is an irreducibility result:

\begin{corollary} If $C\subset \PP^r$ is a smooth curve of degree $d$, then 
for $k\leq \deg C$, the fibered products $\prod_1^k(Y/{\PP^r}^*)$  of the universal hyperplane section 
of $C$ are irreducible.
\end{corollary}

Here is a sufficient condition for double transitivity:

\begin{lemma}\label{transposition lemma}
Let $f : Y \to X$ be a generically finite cover of degree $d$ over an irreducible variety $X$, with  monodromy group $M \subset S_d$.  
If,  for some smooth point $p \in X$ the fiber $f^{-1}(p)\subset V$ consists of $d-2$ reduced points $p_1,\dots, p_{d-2}$ and one point $q$ of multiplicity 2, where $q$ is also a smooth point of $Y$, then $M$ contains a transposition.
\end{lemma}

\begin{proof} To begin with, let $U \subset X$ be a Zariski open subset of the smooth locus in $X$, as in the definition of the monodromy group, so that  $V := f^{-1}(U)$ is also smooth and the restriction $f|_V : V \to U$ expresses $V$ as a $d$-sheeted covering space of $U$, with $U,V$ smooth and $p\in X$

Now, let $A \subset X$ be a small neighborhood of $p$ in the classical topology. The preimage $f^{-1}(A)$ will have $d-1$ connected components, one containing each of the points $p_1,\dots,p_{d-2}$ and $q$; let
 $B \subset f^{-1}(A) \subset Y$ be the connected component of $f^{-1}(A)$ containing the point $q$.

Let $p' \in A \cap U$. Two of the $d$ points of $f^{-1}(p')$ will lie in the component $B$ of $f^{-1}(A)$ containing $q$; call these $q'$ and $q''$. Since $B \cap V$ is connected (it's the complement of a proper subvariety in the neighborhood $B$ of the smooth point $q \in Y$), we can draw a real arc $\gamma : [0,1] \to B \cap V$ joining $q'$ to $q''$; by construction, the permutation of $f^{-1}(p')$ associated to the loop $f \circ \gamma$ will exchange $q'$ and $q''$ and fix each of the remaining $d-2$ points of $f^{-1}(p')$.
\end{proof}

%For every $k\leq \deg C$ we consider the fiber product over $U$ minus the main diagonal $\Delta$
%\begin{align*}
% Y^{k*} :=&Y\times_U Y\times_U \cdots Y\times_U \setminus \Delta  =\\
%                                                 & \{ (H, p_1,\dots, p_k) \in {\PP^r}^* \times C^k \mid p_1,\dots, p_k \in H, p_i \neq p_j \hbox{ for all } i,j \}.
%\end{align*}

\begin{proof}[Proof of the Uniform Position Theorem]
We show first that the monodromy group $M$ is twice transitive, and then that it contains a transposition. It follows that $M$ contains all the transpositions so that  $M = S_d$.
Restricting to an open subset of $C$, we may assume that $C$ is smooth.

For the double transitivity, we introduce a related cover: set
$$
\Phi = \{ (H, p, q) \in {\PP^r}^* \times C \times C \mid p + q \subset H \}
$$
where $p+q$ is the divisor $p+q$ on $C$, viewed as a subscheme of $C$. The projection $\pi_{2,3} : \Phi \to C \times C$ is a $\PP^{r-2}$-bundle, and so irreducible; applying Lemma~\ref{transitivity lemma}, we deduce that $M$ is twice transitive.

To prove that $M$ contains a transposition, we will use the characteristic 0 hypothesis. It follows from Bertini's theorem and Lemma~\ref{tangent not bitangent} that a general hyperplane $H$ containing
the tangent line at a general point $p\in C$ meets $C$ with multiplicity exactly 2 at $p$ and meets $C$ transversely elsewhere. 
Thus the fiber of $\Phi$ over the point $H \in {\PP^r}^*$ consists of the point $p$ with multiplicity 2, and $d-2$ reduced points; applying Lemma~\ref{transposition lemma}, we deduce that $M$ contains a transposition.
\end{proof}

\section{Applications I: the strong form of Clifford's theorem}

Another consequence of the uniform position lemma is a result about sums of linear series.
Recall that if $D$ is a divisor on a curve $C$ we write $r(D) = \dim |D| = h^0(\cO_C(D))-1$.

\fix{Give a proof of Bertini irreducibility as Cor}

\begin{corollary}\label{Clifford equality plus}
If $D,E$ are effective divisors on a curve $C$ then
$$
r(D+E) \geq r(D)+r(E).
$$
If the genus of $C$ is $>0$ and $D+E$ is birationally very ample, then the inequality is strict.
\end{corollary}.

Note that on $\PP^1$ any effective divisor $D$ has $r(D) = \deg D$, so the inequality above is
always an equality for $C = \PP^1$.

\begin{proof}
 The inequality follows in general because the sums of divisors in $|D|$ and divisors in $|E|$ already move in 
 a family of dimension $r(D)+r(E)$; the key point is the strict inequality in case $D+E$ is birationally very ample.
 
In the case where $D+E$ is birationally very ample then, restricting to an open set,
we may identify $C$ with its image under the complete linear series $|D+E|$, and we see that a general hyperplane section $H\cap C$ contains a divisor equivalent to $D$. Since the monodromy of the universal hyperplane section
is the symmetric group, this implies that \emph{every} subset of $\deg D$ elements in $H\cap C$ is
equivalent to $D$. But if $p\in D$ and $q\notin D$, then $D-p+q \equiv D$, whence $q\equiv p$. Thus
$r(p)\geq 1$, so $C\cong \PP^1.$
\end{proof}

A special case of Corollary~\ref{Clifford equality plus} completes the ``equality'' case of Clifford's Theorem:

\begin{corollary}\label{Clifford equality}
If $D$ is an effective divisor on a curve of genus $g\geq 2$ and $r(D) = d/2$, then either $D= 0$ or 
$D=K$ or $C$ is hyperelliptic and $D$ is a multiple of the $g^1_2$.
\end{corollary}

\begin{proof}
  By the argument given in Theorem~\ref{Clifford}, if $r(D) = d/2$, but $D\neq 0,K$, then
$$
r(D) + r(K-D) = r(K) = g-1;
$$
which implies that every canonical divisor is expressible as a sum of a divisor in $|D|$ and a divisor in $|K-D|$.
By Corollary~\ref{Clifford equality plus}, $K$ is not very ample; so $C$ is hyperelliptic. Moreover the
divisors equivalent to $D$ are sums of the fibers of the map given by $|K|$.
\end{proof}

\section{Applications II: the strong form of Martens' theorem}

Another consequence of the uniform position lemma is that it allows us to give a proof of the strong form of Martens' theorem bounding the dimension of the variety $W^r_d(C)$ parametrizing divisor classes of degree $d$ on a curve $C$ with $r(D) \geq r$.
To start, recall the statement of Martens' theorem from Chapter~\ref{new Jacobians chapter}

\begin{theorem}[Martens' theorem]
If $C$ is any smooth projective curve of genus $g$, then for any $d$ and $g$ we have
$$
\dim(W^r_d(C)) \leq d-2r;
$$
moreover, if we have equality for any $r > 0$ and $d < 2g-2$ the curve $C$ must be hyperelliptic.
\end{theorem}

Note that the inequality $\dim(W^r_d(C)) \leq d-2r$ is equivalent to the inequality $\dim C^r_d \; \leq \; d-r$, which is what we actually showed in Chapter~\ref{new Jacobians chapter}. This in turn followed by combining the geometric form of Riemann-Roch with an elementary bound on the dimension of the variety of secant planes to a curve in projective space:

\begin{lemma}[elementary secant plane lemma]
Let $C \subset \PP^n$ be a smooth, irreducible and nondegenerate curve. If we denote by $\Sigma \subset C_d$ the locus of effective divisors $D$ of degree $d$ on $C$ with $\dim \overline D \leq d-r-1$, then for any $d \leq n$ and $r > 0$,
$$
\dim \Sigma \leq d-r.
$$
\end{lemma}

The uniform position lemma allows us to do exactly one better: using it we can, without difficulty, prove the

\begin{lemma}[Strong secant plane lemma]\label{Strong secant plane lemma}
Let $C \subset \PP^n$ be a smooth, irreducible and nondegenerate curve. If we denote by $\Sigma \subset C_d$ the locus of effective divisors $D$ of degree $d$ on $C$ with $\dim \overline D \leq d-r-1$, then for any $d \leq n$ and $r > 0$,
$$
\dim \Sigma \leq d-r-1.
$$
\end{lemma}

\begin{proof}
We set up an incidence correspondence: 
$$
\Gamma := \left\{ (D, H) \in \Sigma \times {\PP^n}^* \mid \overline D \subset H \right\}.
$$
The curve being nondegenerate, the projection map $\Gamma \to  {\PP^n}^*$ is finite. But the fibers of $\Gamma$ over $\Sigma$ have dimension at least $n-d+r$; if we had $\dim \Sigma \geq d-r$, it would follow that $\dim \Gamma \geq n$, and hence that the projection map $\Gamma \to  {\PP^n}^*$ is dominant---contradicting the general position lemma~\ref{general position lemma}.
\end{proof}

Now, if a curve $C$ is non-hyperelliptic, we can apply the strong secant plane lemma to conclude that $\dim W^r_d(C) \leq d-2r-1$; it follows that if we have $\dim W^r_d(C) = d-2r$ the curve $C$ in question must be hyperelliptic.

There are extensions of Martens' theorem to the case $\dim(W^r_d(C)) = d-2r-1$ (Mumford) and $d-2r-2$ (Keem).



\section{Castelnuovo's theorem}

Clifford's theorem gives a complete and sharp answer to the question, ``what linear series can exist on a curve of genus $g$?
But maybe that wasn't the question we meant to ask! After all, we're interested in describing curves in projective space as images of abstract curves $C$ under maps given by linear systems on $C$. Observing that the linear series that achieve equality in Clifford's theorem give maps to $\PP^r$ that are 2 to 1 onto a rational curve, we might hope that we would have a different---and more meaningful---answer if we  restrict our attention to linear series $\cD = (\cL,V)$ for which the associated map $\phi_\cD$ is at least  birational. 

A classical result of Castelnuovo gives a sharp bound, which is in some sense explained by Castelnuovo's classification of the curves that achieve it. For positive integers $d$ and $r$, we write
$$
 d = M(r-1) + 1 + \epsilon \quad \text{ with } \quad 0 \leq \epsilon \leq r-2
$$
and set
$$
\pi(d,r) = \frac{M(M-1)}{2}(r-1) + M\epsilon.
$$

\begin{theorem}[Castelnuovo's bound]\label{Castelnuovo's bound}
If $C \subset \PP^r$ is a reduced, irreducible, nondegenerate curve of degree $d$ and arithmetic genus $p_a$, then
$$
p_a \leq \pi(d,r).
$$
\end{theorem}

In our proof we will use characteristic 0, but in fact the result holds in all characteristics.

We will say that a curve achieving the bound is a \emph{Castelnuovo curve}. We will see in Chapter~\ref{} that the bound is sharp: for every $r$ and $d \geq r$, there exist curves with genus exactly $\pi(d,r)$. 

\begin{example}
For plane curves, $r=2, \ \epsilon = 0, \ M = d-1$ so we recover the necessary $p = \pi(d,2) =  {d-1\choose 2}$. Thus every reduced irreducible
plane curve is a Castelnuovo curve.
\end{example}

\begin{example}For a more typical situation, consider the case $r=3$. We have
$$
p_a \leq \pi(d,3) = \lfloor d^2/4 \rfloor-d+1\, .
$$
It is easy to check that curves on a smooth quadric surface of classes
$(d/2, d/2)$ or $((d-1)/2, (d+1)/2)$
achieve this bound, and a variant of this works for singular quadrics as well. The classification  explained in Chapter~\ref{ScrollsChapter}
shows that these are the only Castelnuovo curves in $\PP^3$.
\end{example}

\begin{fact}
A famous theorem of Gruson and Peskine~\ref{Gruson-Peskine} (see \ref{Hartshorne-report} for an exposition and also the cases of characteristic $>0$ and small fields) completes the picture of the possibilities for the degree $d$ and  genus $g$  of a smooth curve in $\PP^3$. First, if the curve does not lie on a plane or a quadric, then
$$
g\leq \pi_1(d,3) := \frac{d^2-3d}{6} +1
$$
so there are  sometimes gaps in the possible genera for a given degree: for example  $\pi_1(9,3) = 10<\pi(9,3) =12$ but there is
no curve of degree 9 and genus 11: an irreducible curve of class $(a,b)$ on a smooth quadric has arithmetic genus $11 = (a-1)(b-1)$
if and only if $\{a,b\} = \{2,12\}$, so such a curve must have degree 14, and a similar computation holds for a curve on a quadratic cone.

On the other hand, there are smooth curves in $\PP^3$ of every genus and degree realizing $g\leq \pi_1(d)$; and these can be realized as curves
on cubic or quartic surfaces.

The case of $\PP^n$ for $n\geq 3$ remains open. 
\end{fact}
To prove Theorem~\ref{Castelnuovo's bound}, we will give lower bounds for the dimensions of the linear series  cut on a curve $C$ by hypersurfaces of degree $m$. For large values of $m$ the line bundle $\cO_C(m)$ is non-special and $H^1(\sI_{C/\PP^r}(m) = 0$, so a lower bound on the dimension of its space of sections translates, via the Riemann-Roch theorem, into an upper bound on the genus $g$ and this is the Castelnuovo bound.

To state the proof, the following definition will be convenient:

\begin{definition}
If $\sV = (V,cL)$ is a linear system on a variety and $\Gamma$ is a subscheme then the number of conditions
imposed by $\Gamma$ on $\sV$ is the dimension of the image of $V$ in $H^0(\sL\mid_\Gamma) = H^0(\sL \otimes \sO_\Gamma)$; or, numerically,
$$
\dim(V) - \dim \left(V \cap H^0(\cL\otimes \cI_{\Gamma/X}) \right).
$$\end{definition}

Thus, for example, if $\Gamma \subset \PP^r$, then the number of conditions imposed by $\Gamma$ on $H^0(\cO_{\PP^r}(m))$ is the value $h_\Gamma(m)$ of the Hilbert function of $\Gamma$ at $m$.
Note that the number of conditions imposed by $\Gamma$ on a linear system $V$ is necessarily less than or equal to the degree $d$ of $\Gamma$; if it is equal we say that $\Gamma$ \emph{imposes independent conditions on $V$}.

\begin{proof}[Proof of Theorem~\ref{Castelnuovo's bound}]
Suppose that $C \subset \PP^r$ is an irreducible, nondegenerate curve, and let $\Gamma = C \cap H$ be a general hyperplane section of $C$. Let $V_m \subset H^0(\cO_C(m))$ be the linear series cut on $C$ by hypersurfaces of degree $m$ in $\PP^r$, that is, the image of the restriction map
$$
H^0(\cO_{\PP^r}(m)) \to H^0(\cO_C(m)).
$$
Since the ideal sheaf of the hyperplane section is $\cI_{\Gamma/C} = \sO_C(-1)$ we have:
\begin{align*}
h^0(\cO_C(m)) - h^0(\cO_C(m-1)) & \geq \text{\# of conditions imposed by $\Gamma$ on $H^0(\cO_C(m))$} \\
&\geq \text{\# of conditions imposed by $\Gamma$ on $V_m$} \\
&\geq \text{\# of conditions imposed by $\Gamma$ on $H^0(\cO_{\PP^r}(m))$} \\
&\geq h_\Gamma(m).
\end{align*}
Summing these relations, we see that the dimension $h^0(\cO_C(m))$ is bounded below by
$$
h^0(\cO_C(m)) \geq \sum_{k=0}^m h_\Gamma(k).
$$

 To make use of this inequality, we need to know more about $\Gamma$.
What we need follows from a linear independence result:

\begin{lemma}[Linear General Position]\label{general position lemma}
If $C \subset \PP^r$ is an irreducible, nondegenerate curve and $\Gamma = C \cap H$ a general hyperplane section of $C$, then the points of $\Gamma$ are in linearly general position in $H \cong \PP^{r-1}$; that is, no $r$ points of $\Gamma$ lie in a hyperplane $\PP^{r-2} \subset H$.
\end{lemma}


Thus, for example, if $C \subset \PP^3$ is a space curve, the general position lemma says that no three points of a general plane section $\Gamma = H \cap C$ of $C$ are colinear. Even this case is surprisingly tricky to prove (and is false in characteristic $p$); Exercise~\ref{weak general position} sketches a proof.
See \cite[Lemma 1.1]{Rathmann} for a fairly direct proof, valid for smooth curves in every characteristic, except for the plane conic in characteristic 2. We will prove Lemma~\ref{general position lemma} in the next section as a consequence of the far stronger Uniform Position theorem. Remarkably, Lemma~\ref{general position lemma} already suffices for the necessary bound on the number of conditions imposed by hyperplane sections:

\begin{proposition}\label{min hilb}
If $\Gamma \subset \PP^n$ is a collection of $d$ points in linearly general position that span $\PP^n$, then 
$$
h_\Gamma(m) \geq \min\{d, mn+1\}
$$
\end{proposition}

\begin{proof}
Suppose first that $d \geq mn+1$, and let $p_1,\dots,p_{mn+1} \in \Gamma$ be any subset of $mn+1$ points. It suffices to show that $\Gamma' = \{p_1,\dots,p_{mn+1}\}$ imposes independent conditions of $H^0(\cO_{\PP^n}(m))$, that is, for any $p_i \in \Gamma'$ there is a hypersurface $X \subset \PP^n$ of degree $m$ containing all the points $p_1,\dots, \hat{p_i},\dots,p_{mn+1}$ but not containing $p_i$.

To construct such an $X$, group the $mn$ points of $\Gamma' \setminus \{p_i\}$ into $m$ subsets $\Gamma_k$ of cardinality $n$; each set $\Gamma_k$ will span a hyperplane $H_k \subset \PP^n$, and we can take $X = H_1 \cup \dots \cup H_m$. 

In the case where $d<mn+1$, we add $mn+1-d$ general points; each one imposes exactly one
additional condition on hypersurfaces of degree $m$.
\end{proof}
This may seem like a crude argument, but the bound derived is sharp, as shown by the example in Exercise~\ref{linear bound is sharp}.

To complete the proof of Theorem~\ref{Castelnuovo's bound} all that remains is to add up the lower bounds in the proposition. To this end, let $C \subset \PP^r$ be as above an irreducible, nondegenerate curve of degree $d$, and set $M = \lfloor{\frac{d-1}{r-1}}\rfloor$, so that we can write
$$
d = M(r-1) + 1 + \epsilon \quad \text{ with } \quad 0 \leq \epsilon \leq r-2.
$$
We have 
\begin{align*}
h^0(\cO_C(M)) &\geq \sum_{k=0}^M h^0(\cO_C(k)) - h^0(\cO_C(k-1)) \\
&\geq  \sum_{k=0}^M k(r-1)+1 \\
&= \frac{M(M+1)}{2}(r-1) + M + 1
\end{align*}
and similarly
$$
h^0(\cO_C(M+m)) \geq \frac{M(M+1)}{2}(r-1) + M + 1 + md.
$$
For sufficiently large $m$, the line bundle $\cO_C(M+m)$ will be nonspecial, so by the Riemann-Roch Theorem,
\begin{align*}
g &= (M+m)d - h^0(\cO_C(M+m)) + 1 \\
&\leq (M+m)d - \bigl(  \frac{M(M+1)}{2}(r-1) + M + 1 + md \bigr) \\
& = M\bigl( M(r-1) + 1 + \epsilon \bigr) - \bigl(  \frac{M(M+1)}{2}(r-1) + M + 1 \bigr) \\
&= \frac{M(M-1)}{2}(r-1) + M\epsilon.
\end{align*}
 \end{proof}

From the proof we see that if  $C\subset \PP^r$ who degree and genus achieve equality, then all the inequalities in the proof
must be equalities; in particular, the map $H^0(\sO_{\PP^r}(d)) \to H^0(\sO_C(d))$ is surjective for every $d$ so the curve is
arithmetically Cohen-Macaulay in the sense of ****. \fix{give good reference for "arithmetically Cohen-Macaulay"}

\begin{exercise}\label{weak general position}
Let $C \subset \PP^3$ be an irreducible, nondegenerate space curve. Assuming characteristic 0,
\begin{enumerate}
\item Show that if $p, q \in C$ are general, then the tangent lines $\TT_pC$ and $\TT_qC \subset \PP^3$ do not intersect.
\item Using this, show that for general $p, q \in C$ the line $\overline{p,q}$ is not a trisecant; that is, it does not intersect $C$ a third time.
\item Using the irreducibility of $C \times C$, show that $C$ can have at most a 1-parameter family of trisecant lines.
\item Deduce that a general plane $H \subset \PP^3$ does not contain three colinear points of $C$.
\end{enumerate}
\end{exercise}

\begin{exercise}\label{linear bound is sharp}
Let $D \subset \PP^n$ be a rational normal curve. If $\Gamma \subset D$ is any collection of $d$ points on $D$ (or for that matter any subscheme of $D$ of degree $d$) then the Hilbert function of $\Gamma$ is
$$
h_\Gamma(m) = \min\{d, mn+1\}
$$
\end{exercise} 

Castelnuovo in fact proved a converse to this observation:

\begin{fact}
If $d \geq 2n+3$, then any collection of $d$ points in $\PP^n$ in linearly general position with Hilbert function $h_\Gamma(2) = \min\{d, 2n+1\}$ must lie on a rational normal curve (and then $h_\Gamma(m) = \min\{d, mn+1\}$ for all $m$.) See~\cite{****} for a proof.
\fix{reference for Castelnuovo's $2n+3$}
\end{fact}

This was a key ingredient in Castelnuovo's theorem characterizing curves of maximal genus; we'll see how it's applied in Chapter~\ref{} (scrolls, if we include it). For a proof, see for example~\cite{ACGH} or~\cite{Montreal}

\begin{exercise}
Show that with $C$ and $M$ as above, the line bundle $\cO_C(M)$ is nonspecial. (We will see in Section~\ref{} that this is sharp; that is, there exist curves $C \subset \PP^r$ with $\cO_C(M-1)$ special).
\end{exercise}

The truth of Lemma~\ref{general position lemma} was originally asserted by Castelnuovo. It follows easily from the far stronger Uniform Position Theorem, proved in Section~\ref{uniformSection} below.


\section{Existence of good projections}\label{projection section}\label{good projections}

\fix{reconcile with the exercise currently right after $g+3$ in the Jacobians chapter}

Another application of the Linear General Position lemma~\ref{general position lemma} is to show that every smooth curve $C$ is birational to a nodal plane curve $C_0 \subset \PP^2$. Even better, such a birational map can be obtained by following any embedding in projective space with a  linear projection $\pi_\Lambda : C \to \PP^2$ from a general $(n-3)$-plane $\Lambda \subset \PP^n$ i. 

\begin{proposition}\label{nodal projection}
If $C \subset \PP^n$ is a smooth nondegenerate curve in projective space, and $\Lambda \cong \PP^{n-3} \subset \PP^n$ a general $(n-3)$-plane, then the projection $\pi_\Lambda : C \to \PP^2$ is birational onto its image, a curve with only ordinary nodes.
\end{proposition}

\begin{proof} Recall that the secant variety of $C$ consists of the union of the lines $\overline{q,r}$ joining pairs of distinct points $q,r \in C$, plus the tangent lines ${\mathbb T}_q(C)$; altogether, these lines form a single family, parametrized by the symmetric square of $C$. The secant variety thus has dimension 3, so that a general $(n-3)$-plane $\Lambda$ will meet it in finitely many points $p \in \PP^n$. The images of these points are the singularities of the image curve $C_0 \subset \PP^2$: if $p \in \overline{q,r}$ lies on a secant line, then the projection $\pi_\Lambda$ fails to be one-to-one at the image point $\pi_\Lambda(q) = \pi_\Lambda(r)$, while if $p$ lies on a tangent line ${\mathbb T}_q(C)$ the differential of $\pi_\Lambda$ will vanish at $q$.

To prove the proposition, we will show:
\begin{enumerate}\label{needed for nodes}
\item there do not exist three points $p,q,r \in \widetilde C$ such that $p, q, r$ and $\Lambda$ are all contained in a $\PP^{N-2}$; and
\item there does not exist a pair of points $p, q \in \widetilde C$ such that 
$p, q$ and $\Lambda$ are contained in a $\PP^{N-2}$ and $\TT_p(\widetilde C), \TT_q(\widetilde C)$ and $\Lambda$ are contained in a $\PP^{N-1}$.
\end{enumerate}

It will be easier to visualize what follows if we first reduce to the case $n=3$. If $\Gamma \subset \PP^n$ is a general $(n-4)$-plane, then a general $(n-3)$-plane $\Lambda$ containing $\Gamma$ is a general $(n-3)$-plane in $\PP^n$, so we can view the projection $\pi_\Lambda : C \to \PP^2$ as the composition of the projection $\pi_\Gamma : C \to \PP^3$ with the projection $\pi_p$ of the image $\pi_\Gamma(C)$ from a general point $p \in \PP^3$. Moreover, since a general  $(n-4)$-plane $\Gamma \subset \PP^n$ is disjoint from the secant variety of $C \subset \PP^n$, the projection $\pi_\Gamma : C \to \PP^3$ is an embedding. Thus we may assume that $C \subset \PP^3$ is a smooth curve to start with, and show that the projection from a general point $p \in \PP^3$
is birational with image a nodal curve.


Since the family of secant lines to $C$ is two-dimensional,  their union is at most 3-dimensional, so only a finite number of them will contain a general point $p \in \PP^3$; thus the map $\pi_p$ is birational onto its image. Moreover, since the family of tangent lines is 1-dimensional, a general point $p \in \PP^3$ will not lie on any tangent lines, so the differential of $\pi_p : C \to \PP^2$ does not vanish; in other words, the map $\pi_p$ is an immersion.

To see that there are no more than two points of $C$ in a fiber of the map to $C_0$ we invoke Lemma~\ref{general position lemma}. That is, we show that a general point $p \in \PP^3$ does not lie on any trisecant line. Since the lines that meet $C$ at least 3 time form a closed subscheme of the 2-dimensional secant locus, it suffices to to show that not every secant line to $C \subset \PP^3$ is a trisecant. By the General Position lemma, a general plane $H \subset \PP^3$ meets $C$ in a set of point in linearly general position; in particular, no three are collinear. Thus the line joining any two points of $H\cap C$ is a secant  to $C$ that is not  trisecant.

We next observe that the tangent lines to $C$ at general points do not meet, since otherwise the projection from any tangent line
$C\to \PP^1$ would be inseparable and---since $C$ is nondegenerate---not constant.
\begin{exercise}
Show that if every secant line to a nondegenerate curve in $\PP^3$ were a trisecant, then every pair of tangent lines would be coplanar. Thus projection from a tangent line would be inseparable, which cannot happen in characteristic 0.
\end{exercise}

We have shown that the map $pi_p$ is an immersion, and at most two-to-one everywhere; thus the image curve $C_0 \subset \PP^2$ will have at most double points, and an analytic neighborhood of each double point will consist of two smooth branches.
To complete the proof of Proposition~\ref{nodal projection} we have to show that those two branches have distinct tangent lines; that is, that
if $q, r \in C$ are any two points collinear with $p$, then the images of the tangent lines ${\mathbb T}_q(C)$ and ${\mathbb T}_r(C)$ in $\PP^2$ are distinct. But if  $\pi_p({\mathbb T}_q(C)) = \pi_p({\mathbb T}_r(C))$ then  ${\mathbb T}_q(C)$ and ${\mathbb T}_r(C)$ lie in a plane, and thus intersect.

Classically, a secant line $\overline{q,r}$ was called a \emph{stationary secant} if the tangent lines ${\mathbb T}_q(C)$ and ${\mathbb T}_r(C)$ met; in these terms, it remains to show that a general point $p \in \PP^3$ does not lie on any stationary secant. Again, this follows from the assertion that the family of stationary secants has dimension at most 1; and again, since the family of all secant lines is irreducible of dimension 2, it will suffice to show that not every secant line to $C$ is a stationary secant.

Suppose the contrary, that for a general pair of points $q, r \in C$ we have $\TT_qC \cap \TT_rC \neq \emptyset$; otherwise the projection $\pi_{\TT_qC} : C \to \PP^1$ would have derivative everywhere 0, but be nonconstant. Thus the space of  stationary secants is at most 1-dimensional, and a general $p \in \PP^3$ will not lie on one.
\end{proof}


In characteristic $p > 0$ it we do not know whether a general projection of a smooth curve to $\PP^2$ is nodal---but it's still true that every curve can be birationally mapped onto a nodal curve in $\PP^2$:

\begin{exercise}
Let $C\subset \PP^n$ be a smooth curve. If we re-embed $C$ by a Veronese map of sufficiently high degree---that is, we let $\nu_m : \PP^n \to \PP^N$ be the $m$th Veronese map, and let $\widetilde C = \nu_m(C)$ be image of $C$. Prove:

\begin{proposition}[Proposition~\ref{nodal projection} in characteristic $p$]\label{positive characteristic nodes}
The projection of $\widetilde C$ from a general $\PP^{N-3}$ is a nodal plane curve.
\end{proposition}

Hint: show that the items of~\ref{needed for nodes}  are true in this situation.
\end{exercise}

\begin{exercise}
Let $C_0$ be a plane quartic curve with two nodes $q_1, q_2$; let $\nu : C \to C_0$ be its normalization, and let $o \in C$ be any point not lying over a node of $C_0$.
By the genus formula, $C$ has genus 1. Using the construction above, describe the group law on $C$ with $o$ as origin.
\end{exercise}


%\subsection{Uniform position} \label{uniformSection}
%
%\subsubsection{The monodromy group of a generically finite morphism}
%
%Let $f : Y \to X$ be a dominant map between varieties of the same dimension over $\CC$, and suppose that $X$ is irreducible. There is then an open subset $U \subset X$ such that $U$ and 
%its preimage $V = f^{-1}(U)$ are smooth, and the restriction of $f$ to $V$ is a covering space in the classical topology. Let $d$ be the number of sheets. This is the degree of the extension $K(Y)/K(X)$.
%
%A basic construction from homotopy theory now associates a monodromy group to any finite topological covering map: Choose a base point $p_0 \in U \subset X$, and suppose $\Gamma := f^{-1}(p_0)  = \{q_1,\dots,q_d\}$. If $\gamma$ is any loop in $U$ with base point $p_0$, for any $i = 1, \dots, d$ there is a unique lifting of $\gamma$ to an arc $\tilde \gamma_i$ in $V$ with initial point $\tilde \gamma_i(0) = q_i$ and end point $\tilde \gamma_i(1) = q_j$ for some $j \in \{1,2,\dots,d\}$. Since we could traverse the loop in the opposite direction, the index $j$ determines $i$, and the map $i\mapsto j$ is a permutation of $\{1,2,\dots,d\}$. 
%Since the set $\Gamma$ is discreet, the permutation depends only on the class of $\gamma$ in $\pi_1(U,p_0)$ so we have defined a homomorphism to the symmetric group:
%$$
%\pi_1(U,p_0)  \to {\rm Perm}(\Gamma) \cong S_d.
%$$
%The image $M$ of this map is called the \emph{monodromy group} of the map $f$. It depends on the labeling of the points of $\Gamma$, but is well-defined  up to conjugation. Moreover, it is independent of the choice of open set $U$: if $U' \subset U$ is a Zariski open subset, the map $\pi_1(U', p_0) \to \pi_1(U,p_0)$ is surjective,  so the image of $\pi_1(U', p_0)$ in $S_d$ is the same.
%
%\begin{fact}
%The theory of finite coverings of algebraic varieties is not only analogous to Galois theory, is \emph{is} Galois theory: In the situation described above, if we assume that $Y$ is irreducible, then the pullback map $f^*$ expresses the function field $K(Y)$ as a finite algebraic extension of $K(X)$. The degree of this extension is the degree of the map $f$. The monodromy group of $f$  the Galois group of the Galois normalization of $K(Y)$ over $K(X)$. (see \cite{Harris1979}.)
%\end{fact}
%
%Since we assumed that $X$ is irreducible, the space $U$ is (path) connected, and it follows that the monodromy group is transitive if and only if the space $V$ is (path) connected. But $V$ is a smooth
%variety, so this is the case if and only if $V$ is irreducible.
%
%We will next compute the monodromy group of the  universal hyperplane section, constructed as follows:
%Let $C \subset \PP^r$ be an irreducible, nondegenerate curve of degree $d$, and let $X = {\PP^r}^*$ be the space of hyperplanes in $\PP^r$. We define the \emph{universal hyperplane section of $C$} to be the projection to $f: Y\to {\PP^r}^*$ of the incidence variety
%$$
%Y = \{ (H, p) \in {\PP^r}^* \times C \mid p \in H \}.
%$$
%The fibers of $f$ are the hyperplane
%sections of $C$, so it is a dominant finite map. If we let $U\subset {\PP^r}^*$ be an open subset of hyperplanes
%meeting $C$ transversely, then the restriction of $f$ to the preimage $V$ of $U$ is a covering space
%whose fibers each consist of $d$ distinct points. The preimage in $Y$ of a point $p\in C$ is the set of hyperplanes containing
%$p$, a copy of $\PP^{r-1}$, and thus $Y$ is irreducible. Thus the monodromy group of $f$ is transitive. But much more is true:
%
%\begin{theorem}[Uniform Position Theorem]\label{uniform position lemma}
%The monodromy group of the universal hyperplane section of an irreducible curve $C \subset \PP^r_\CC$ is the full symmetric group $S_d$.
%\end{theorem}
%
%Informally, this theorem says that two subsets of the same cardinality in the general hyperplane section of $C$
%are indistinguishable from the point of view of any discrete invariant that is semicontinuous in the Zariski topology. 
%The Uniform position result fails over fields of finite characteristic; see \cite{Rathmann} and \cite{Kadets} for what is known. 
%
%The following is a good example of how the theorem is applied:
%
%\begin{corollary}[numerical uniform position lemma]\label{numerical uniform position lemma}
%Let $C\subset \PP^r$ be an irreducible curve, and let $H\cap C$ be a general hyperplane section. Any two subsets of $H\cap C$ with the same cardinality impose the same number of conditions on forms of any degree; that is, any two subsets of the same cardinality have the same Hilbert function.
%\end{corollary}
%
%\begin{proof} Let $U = {\PP^r}^* \setminus C^*$ be the open subset of hyperplanes transverse to $C$, and let $Y\to U$ be the universal hyperplane section.
%Thus
%\begin{align*}
% Y^{k*} :=&Y\times_U Y\times_U \cdots Y\times_U \setminus \Delta  =\\
%                                                 & \{ (H, \Gamma) \in {\PP^r}^* \times C^k \mid \Gamma \subset H, \hbox{ and $\Gamma$ consists of $k$ distinct points.}\}\\
%                                                 &\to U
%\end{align*}
%is the family subsets $\Gamma$ of cardinality $k$ in hyperplane sections $H \cap C$ of $C$. Because the monodromy is of the universal hyperplane section of $C$ is
%$k$-times transitive, $Y_k^*$ is irreducible.
%
%For each $m$ the number of conditions that $\Gamma$ imposes on forms of degree $m$ is lower semicontinuous, so it achieves its maximum on a Zariski open subset of $Y^{k*}$. Since $Y^{k*}$ is irreducible, the complement of this open set has dimension strictly less than $\dim Y^{k*} = r$. Thus a general hyperplane $H \in {\PP^r}^*$ will lie outside the image of $Y^{k*}$, meaning that the number of conditions imposed by all the $k$-element subsets $\Gamma \subset C \cap H$ will have this maximal value.
%\end{proof}
%
%Lemma~\ref{general position lemma} follows:
%
%\begin{proof}[Proof of Lemma~\ref{general position lemma}] As in the proof above, let $Y^{k*}$ be the family of $k$-element subsets
%of transverse hyperplane sections of $C$, and suppose that $k$ is maximal such that for general hyperplanes $H$ the
%$k$-element subsets of $H\cap \Gamma$ imposes indedpendent conditions on linear forms, or in other words, is linearly independent.
%It follows from Corollary~\ref{numerical uniform position lemma} that all subsets of $k+1$ elements are dependent, so $H\cap C$ is
%contained in the span of a single $k$-element subset. Since $C$ is nondegenerate, a general hyperplane is spanned
%by its intersection with $C$, so $k=r+1$.
%\end{proof}
%
%Since the symmetric group $S_d$ is generated by transpositions, we may prove that a subgroup $M\subset S_d$ is equal to $S_d$ by showing that $M$ is doubly transitive and contains a transposition (of course it will then follow that $M$ is $k$-fold transitive for any $k\leq d$.) We can translate these properties into geometry as follows:
%
%\begin{lemma}\label{transitivity lemma}
%Let $f : Y \to X$ be a generically finite cover of degree $d$, with  monodromy group $M \subset S_d$.
%The $k$-fold fiber product $\prod_1^k(Y/X)$ is irreducible if and only if $M$ is $k$ times transitive.
%\end{lemma}
%
%\begin{proof}
%The monodromy of the covering $f^k: \times^k(Y/X) \to X$ is the the diagonal action of
%$M$ in the product $\prod_1^k S_d$.
%\end{proof}
%
%Thus a consequence of the Uniform Position Theorem is an irreducibility result:
%
%\begin{corollary} If $C\subset \PP^r$ is a smooth curve of degree $d$, then 
%for $k\leq \deg C$, the fibered products $\prod_1^k(Y/{\PP^r}^*)$  of the universal hyperplane section 
%of $C$ are irreducible.
%\end{corollary}
%
%Here is a sufficient condition for double transitivity:
%
%\begin{lemma}\label{transposition lemma}
%Let $f : Y \to X$ be a generically finite cover of degree $d$ over an irreducible variety $X$, with  monodromy group $M \subset S_d$.  
%If,  for some smooth point $p \in X$ the fiber $f^{-1}(p)\subset V$ consists of $d-2$ reduced points $p_1,\dots, p_{d-2}$ and one point $q$ of multiplicity 2, where $q$ is also a smooth point of $Y$, then $M$ contains a transposition.
%\end{lemma}
%
%\begin{proof} To begin with, let $U \subset X$ be a Zariski open subset of the smooth locus in $X$, as in the definition of the monodromy group, so that  $V := f^{-1}(U)$ is also smooth and the restriction $f|_V : V \to U$ expresses $V$ as a $d$-sheeted covering space of $U$, with $U,V$ smooth and $p\in X$
%
%Now, let $A \subset X$ be a small neighborhood of $p$ in the classical topology. The preimage $f^{-1}(A)$ will have $d-1$ connected components, one containing each of the points $p_1,\dots,p_{d-2}$ and $q$; let
% $B \subset f^{-1}(A) \subset Y$ be the connected component of $f^{-1}(A)$ containing the point $q$.
%
%Let $p' \in A \cap U$. Two of the $d$ points of $f^{-1}(p')$ will lie in the component $B$ of $f^{-1}(A)$ containing $q$; call these $q'$ and $q''$. Since $B \cap V$ is connected (it's the complement of a proper subvariety in the neighborhood $B$ of the smooth point $q \in Y$), we can draw a real arc $\gamma : [0,1] \to B \cap V$ joining $q'$ to $q''$; by construction, the permutation of $f^{-1}(p')$ associated to the loop $f \circ \gamma$ will exchange $q'$ and $q''$ and fix each of the remaining $d-2$ points of $f^{-1}(p')$.
%\end{proof}
%
%%For every $k\leq \deg C$ we consider the fiber product over $U$ minus the main diagonal $\Delta$
%%\begin{align*}
%% Y^{k*} :=&Y\times_U Y\times_U \cdots Y\times_U \setminus \Delta  =\\
%%                                                 & \{ (H, p_1,\dots, p_k) \in {\PP^r}^* \times C^k \mid p_1,\dots, p_k \in H, p_i \neq p_j \hbox{ for all } i,j \}.
%%\end{align*}
%
%\begin{proof}[Proof of the Uniform Position Theorem]
%We show first that the monodromy group $M$ is twice transitive, and then that it contains a transposition. It follows that $M$ contains all the transpositions so that  $M = S_d$.
%Restricting to an open subset of $C$, we may assume that $C$ is smooth.
%
%For the double transitivity, we introduce a related cover: set
%$$
%\Phi = \{ (H, p, q) \in {\PP^r}^* \times C \times C \mid p + q \subset H \}
%$$
%where $p+q$ is the divisor $p+q$ on $C$, viewed as a subscheme of $C$. The projection $\pi_{2,3} : \Phi \to C \times C$ is a $\PP^{r-2}$-bundle, and so irreducible; applying Lemma~\ref{transitivity lemma}, we deduce that $M$ is twice transitive.
%
%To prove that $M$ contains a transposition, we will use the characteristic 0 hypothesis. It follows from Bertini's theorem and Lemma~\ref{tangent not bitangent} that a general hyperplane $H$ containing
%the tangent line at a general point $p\in C$ meets $C$ with multiplicity exactly 2 at $p$ and meets $C$ transversely elsewhere. 
%Thus the fiber of $\Phi$ over the point $H \in {\PP^r}^*$ consists of the point $p$ with multiplicity 2, and $d-2$ reduced points; applying Lemma~\ref{transposition lemma}, we deduce that $M$ contains a transposition.
%\end{proof}

\input footer.tex