
\input{header.tex}


\chapter{Hyperplane sections of a curve}\label{uniform position}

One way to study a curve in $\PP^n$ is to use properties of its hyperplane sections. In this chapter, we will take up the question: if $C \subset \PP^n$ is a reduced, irreducible and nondegenerate curve, what can we say about the geometry of the points in the  hyperplane section $\Gamma = C \cap H$ of $C$ of a curve, and what more can we say about a general hyperplane section?

%\fix{which of our applications is good for integral curves, not just smooth ones?}

We begin by showing that the points in every hyperplane section span the hyperplane. Then, in the second section, we prove a result originally due to Castelnuovo, that in the case of a general hyperplane section, \emph{every} subset of $n$ points in a spans the hyperplane; that is, any $n$ points of $\Gamma$ are linearly independent. (This 
holds for smooth curves in arbitrary characteristic (but see Exercise~\ref{strange curves} for
interesting counterexamples of singular examples in positive characteristic). 

In the following sections we give a series of applications of this result. Then, in Section~\ref{uniform position section} we return to the characteristic 0 case to prove the much stronger \emph{uniform position lemma}. Finally we give some applications
of this stronger result.
\section{Arbitrary hyperplane sections}

\begin{proposition}\label{arbitrary hyperplane}
If $C\subset \PP^r$ is a nondegenerate, reduced, irreducible curve, and $H$ is a hyperplane, then the subscheme
$\Gamma := H\cap C$ spans $H$; that is, $H^0(\sI_{\Gamma/H}(1)) = 0$.
\end{proposition}

See Exercise~\ref{arbitrary hyperplane examples} for the necessity of the hypotheses.

Geometrically: since $C$ is nondegenerate, reduced, and irreducible, every hyperplane meets $C$ in a finite subscheme.
if $\Gamma$ spanned only a codimension 2 plane $\Lambda$, then in the family of hyperplanes
containing $\Lambda$ there would be some meeting $C$ in a finite scheme of degree more than $\deg \Gamma = \deg C$
points, a contradiction. Here is a more formal argument:

\begin{proof}
Write $h\in H^0\sO_{\PP^n}(1)$ for the linear form vanishing on $H$.
Because $C$ is reduced irreducible and nondegenerate,
$h$ is a nonzerodivisor modulo $\sI_C$, and thus
we have 
$$
\begin{aligned}
\sI_{C/\PP^n}/h\sI_{C/\PP^n} &= \sI_{C/\PP^n}/h\sI_{C/\PP^n} \\
&= \sI_{C/\PP^n}/((h)\cap\sI_{C/\PP^n})\\
 &=(\sI_{C/\PP^n}+(h))/(h)\\
 &= \sI_{\Gamma/H}.
\end{aligned}
 $$
 From this equality we get an exact sequence
 $$
0\to \sI_C \rTo^h \sI_C(1)\to \sI_\Gamma(1)\to 0.
$$
Again because $C$ is reduced and irreducible, $H^0\sO_C$ contains only the constant function, so the map $H^0(\sO_{\PP^n}) \to H^0(\sO_C)$ is surjective, 
from which it follows that $H^1(\sI_{C/\PP^n}) = 0.$ 
Thus the restriction map $H^0(\sI_C(1))\to H^0(\sI_\Gamma(1))$ is surjective. Since
$C$ is nondegenerate, $H^0(\sI_C(1)) = 0$, and the desired result follows.
\end{proof}


\section{Linearly general position}
In this section we will allow our algebraically closed ground field to have arbitrary characteristic. 

An irreducible and reduced curve
$C\subset \PP^n$ over an algebraically closed field $k$, other than a line, is called \emph{strange} if
all the tangent lines at smo gitoth points of $C$ meet in a point $p\in \PP^n$. If this conditions seems somewhat, well, strange, that may be because no such curves exist in characteristic 0: if it were the case that all the tangent lines at smooth points of $C$ met in a point, then the projection of $C$ from that point would have derivative 0 everywhere. But strange curves do exist in characteristic $p > 0$, albeit rarely: Pierre Samuel showed that the only \emph{smooth} curve that is strange (in any characteristic) is the conic in characteristic 2; a proof is given in \cite[Theorem IV.3.9]{Hartshorne1977}. Without the smoothness hypothesis, however there are other examples; see Exercise~ref{strange curves}.

Given this language, we can state our first result. Recall that if $C\subset \PP^n$ is a reduced curve then, by Bertini's theorem, a general hyperplane
section of $C$ consists of smooth points. 

\begin{theorem} \label{basic linear independence}(\cite[Lemma 1.1]{Rathmann})\label{linear general position}
Suppose that $n\geq 3$. If $C\subset \PP^n$ is a nondegenerate irreducible, reduced curve, that is not neither strange
and $H$ is a general hyperplane, then any $n$ points of $H\cap C$ are
linearly independent.
\end{theorem}

For example, the theorem says that for $C \subset \PP^3$, the general hyperplane section does not contain a line
that meets $C$ 3 or more times---a \emph{multisecant}. Another special configuration which is not present, as it turns out,
is any pair of points $p\neq q$ on $C$ such that the the tangent lines $T_p(C), T_q(C)$ to $C$ at $p$ and $q$ intersect---what was classically
called a \emph{staionary secant}. More formally we define two subsets of the symmetric square $C_2$ of $C$:
$$
\begin{aligned}
 mult(C) &:=\{(p,q) \in C_2 \mid \overline{p,q} \hbox{meets $C$ in a scheme of length $\geq 3$}\};\\
stat(C) &:= \{(p,q) \in C_2\setminus \Delta \mid  \hbox{meets $C$ in a scheme of length $\geq 3$}\},
\end{aligned}
$$
where $\overline{p,q}$ denotes the secant line through $p,q$ (or the tangent line if $p=q$ and $\Delta\subset C_2$
is the diagnal.
Clearly if $C$ is strange or planar then $stat(C) = C_2$. If $C$ is planar with $\deg C>2$ then $mult(C) = C_2$ as well.
It turns out that these are the only cases:

\begin{proposition}\label{mult and stat}
 If $C\subset \PP^n$ is a reduced, irreducible, curve that is neither planar nor strange, then $mult(C)\subset C_2$
 and $stat(C)\subset C_2\setminus \Delta$ are closed subsets of dimension $\leq 1$.
\end{proposition}

\begin{proof}
To show that $mult(C)$ is closed, consider the projection to the first factor, 
$$
\{((p,q), r) \in C_2\times \PP^n \mid r \in \overline{p,q}\} \to C_2 : ((p,q),r)\mapsto (p,q).
$$
By the semicontinuity of the degree of fiber dimension, the set of points lying in fibers
of length $\geq 3$ is closed, and $mult(C)$ is the image of this set. Because the map is finite, 
$mult(C)$  is closed as well.

Similarly since 
$$
\{((p,q), r) \in (C_2\setminus \Delta)\times \PP^n \mid r \in T_p(C)\cup T_q(C) \}
$$
is closed in $(C_2\setminus \Delta)\times \PP^n$, its image under the projection to the first factor,
which is $stat(C)$, is closed in $C_2\setminus \Delta$

Suppose, contrary to the Proposition, that $\dim stat(C) >1$. Since $stat(C)$ is a closed subset of the irreducible variety $C_2\setminus \Delta$ it
follows that $stat(C) = C_2\setminus \Delta$, so every pair of tangent lines meet. Any three
lines in projective space that meet pairwise must be either coplanar or meet in a point. Thus if $C$ is not strange,
then all the tangents to $C$ lie in a common plane, and $C$ is planar, contradicting our hypothesis
and proving that $\dim stat(C)\leq 1$.

Finally suppose, contrary to the Proposition,  that $\dim mult(C)>1$. As in the previous argument, this
implies that every secant is a multisecant.  We will show in this case that $\dim stat(C) =2$, a contradiction. 

To this end, consider the set  $I \subset  \{(p,r,r')\in C^3$ consisting of those triples of distinct points 
such that $\pi_p(r) = \pi_p(r')$ is a smooth point of $\pi_p(C)$,
where $\pi_p$ denotes the projection of $C$ from $p$.

The tangent lines $T_r(C)$ and $T_{r'}(C)$ both project from $p$ to the tangent line $T_{\pi_p(r)}(\pi_C)$,
and thus they lie in the 2-plane spanned by $p$ and this line; it follows that $(r,r') \in stat(C)$.
The projection $I \to C_2: (p,r,r') \mapsto (r,r')$ is finite since the secant line $\overline{r,r'}$ meets $C$ in only finitely many points, 
so $\dim I \leq \dim stat(C)$.

 If the projection from some point $p$ were everywhere ramified,
then all tangents to $C$ would pass through $p$, and $C$ would be strange.
Since also every secant to $C$ is a multisecant it follows that  for every smooth point $p\in C$ there is an open set of points $q\in \pi_p(C)$ such that
the fiber $\pi_p^{-1}(q)$ contains two distinct points of $C$. Thus the projection $I \to C: (p,r,r') \mapsto p$ maps
$I$ onto an open set of $C$, with 1-dimensional fibers, so $2 = \dim I\leq \dim stat(C)$ as claimed. \end{proof}


\begin{proof}[Proof of Theorem~\ref{linear general position}]
First suppose that $n=3$, and consider the incidence correspondence
$$
I := \{ ((p,q), H) \in mult(C) \times \PP^{3*} \mid \overline{p,q}\in H\}.
$$
Since $\dim \overline{p,q} = 1$, the fibers of the projection of $I$ to $mult(C)$ have dimension 1,
so $\dim I = 1+\dim mult(C)  \leq 2$. Consequently the image of  the projection of $I$ to $\PP^{3*}$ has
dimension $\leq 2$, so there is an open set of planes that contain no multisecants, proving the Theorem in this case.

We next do induction on $n\geq 4$.We will show first that the image $C$ of
$C$ under the projection $\pi_p: C\to C'$ from a general point of $C$ is not strange. Since $C\subset \PP^n$ is neither
planar nor strange
Proposition~\ref{mult and stat} shows that we may choose two smooth points
$r,r'\in C\subset \PP^n$ such that $T_r(C)\cap T_{r'}(C) = \emptyset$, and thus 
$T_r(C)$ and $T_{r'}(C)$ together span a 3-plane $L$. Since $C$ is nondegenerate we may
choose a point $p\in C$ outside $L$, and it follows that $\pi_p$ restricted to $L$ is an isomorphism.
Thus $T_{\pi_p(r)}(C') \cap T_{\pi_p({r'})}(C') = \emptyset$, so $C'$ is not strange, 
and since $n\geq 4$ the curve $C'$ is not planar either.

Arguing by contradiction, suppose that the general hyperplane section of $C$ contains a set of $n$ linearly dependent points. Consider the closed subsets  
$$
I_1 \subset I \subset \{(p,H) \in C \times \PP^{n*}\}
$$
where I is defined by the conditions that $p\in H$ and $I_1$ is defined by the condition that, in addition, there exist $p_2,\dots, p_n\in H$
such that $p, p_2, \dots, p_n$ are dependent. Both $I_1$ and $I$ are closed subsets.

By hypothesis, both $I_1$ and $I$ project onto open sets of $\PP^{n*}$, so they have the same dimension.
But $I$ is irreducible: it projects onto an open subset of $C$ with fibers isomorphic to $\PP^{n-1}$. Thus $I_1 = I$,
and  $p$ is part of a dependent set
$\{p, p_2,\dots, p_n\}$ in a general hyperplane containing $p$.

Let $H'$ be a general hyperplane in $\PP^{n-1}$
and let $H = \pi_p^{-1}(H')$, which is a general hyperplane containing $p$. The intersection $H\cap C$
contains $n$ dependent points $\{p, p_2,\dots, p_n\}$, and it follows that $\pi_p(p_2),\dots \pi_p(p_n)$
are dependent points of $H'\cap C'$. This contradicts the induction hypothesis, and proves that
the general hyperplane section of $C$ is in linearly general position.
\end{proof}
 
 

\section{Existence of good projections}\label{projection section}\label{good projections}

We can use Theorem~\ref{basic linear independence} to show that every smooth curve $C$ is birational to a nodal plane curve $C_0 \subset \PP^2$, in many ways.

\begin{proposition}\label{nodal projection}
If $C \subset \PP^n$ is a smooth nondegenerate curve in projective space, let $\Lambda \cong \PP^{k} \subset \PP^n$ be a general $k$-plane, and let
$\pi_\Lambda$ be the projection from $\Lambda$, restricted to $C$. If $n\geq 4$ and $k=n-4$,then
$\pi_\Lambda: C \to \PP^3$ defines an isomorphism of $C$ onto its image, while if $n\geq 3$ then $\pi_\Lambda$ is birational onto its image, which is a curve with only ordinary nodes.
\end{proposition}

\begin{proof} Recall that the secant variety of $C$ consists of the union of the lines $\overline{q,r}$ joining pairs of distinct points $q,r \in C$, plus the tangent lines ${\mathbb T}_q(C)$; altogether, these lines form a family, parametrized by the symmetric square $C^{(2)}$ of $C$. More precisely every subscheme $\lambda$ of
length 2 in $\PP^n$ spans a line $\overline \lambda$. The incidence variety
$$
I:=\{(\lambda, p)\mid \lambda\in C^{(2)} \hbox{ is a divisor of degree 2 on }C,\ p\in \overline \lambda\subset\PP^n\}
$$
projects to $C^{(2)}$ with 1-dimensional fibers isomorphic to $\PP^1$, and thus
is irreducible of dimension 3. Its image in $\PP^n$ under the second projection
is the secant variety of $C$, which is thus irreducible of dimension $\leq 3$.
It follows that a general
$n-4$-plane does not meet the secant variety, and the first statement of the Proposition follows.

If $n>3$ then by first projecting from a general $n-4$-plane inside $\Lambda$ we may reduce to the case $n=3$, and assume that $\Lambda$ is a general point of $\PP^3$. By a variant of the argument above, the union 
of the tangent lines to $C$ is a surface, and thus does not contain $\Lambda$.
It follows that $\pi_\Lambda$ is locally an analytic isomorphism.

To show that the fibers of $\pi_\Lambda$ are subschemes of length at most 2,
we need to show that $\Lambda$ does not lie on any multisecant line. 

By Theorem~\ref{basic linear independence} the family of multisecant lines to $C$ is a proper subscheme of the irreducible two-dimensional family of secant lines, so the union of the trisecant lines is at most 2 dimensional, and we see that the fibers of $\pi_\Lambda$ all have degree $\leq 2$. Furthermore, the general fiber of the projection
from the incidence correspondence $I$ to $\PP^3$ is empty or finite, so only a finite number of secant lines will contain $\Lambda$, and we see that $\pi_\Lambda$ is birational. 

We have shown that the map $\pi_\Lambda$ is an immersion, and at most two-to-one everywhere; thus the image curve $C_0 \subset \PP^2$ will have at most double points, and an analytic neighborhood of each double point will consist of two smooth branches. To complete the proof of Proposition~\ref{nodal projection} we have to show that those two branches have distinct tangent lines; that is, that
if $q, r \in C$ are any two points collinear with $\Lambda$, then the images of the tangent lines ${\mathbb T}_q(C)$ and ${\mathbb T}_r(C)$ in $\PP^2$ are distinct. But if  $\pi_p({\mathbb T}_q(C)) = \pi_p({\mathbb T}_r(C))$ then  ${\mathbb T}_q(C)$ and ${\mathbb T}_r(C)$ lie in a plane, and thus intersect.

Since the family of all secant lines is irreducible of dimension 2,  it will suffice to show that not every secant line to $C$ is a stationary secant or, equivalently, that not every pair of tangent lines to $C$ meet. We saw in Lemma~\ref{incident tangents} that in the contrary case the curve $C$ would be either strange or planar, a contradiction.
\end{proof}


\section{The case of equality in Martens' theorem}

Theorem~\ref{basic linear independence}  allows us to analyze the case of equality in Martens' theorem bounding the dimension of the variety $W^r_d(C)$ parametrizing divisor classes of degree $d$ on a curve $C$ with $r(D) \geq r$
(Theorem~\ref{Martens' inequality}.
To start, recall the statement:

\begin{theorem}(Martens)\label{full Martens}
If $C$ is any smooth projective curve of genus $g$, then for any $r>d-g$ we have
$$
\dim W^r_d(C) \leq d-2r.
$$
Equality holds iff either $C$ is hyperelliptic
or $d=r=0$ or $d=2g-2, r=g-1$; in either of the last two cases $W^r_d$ is
a single point.
\end{theorem}

Note that the inequality $\dim W^r_d(C) \leq d-2r$ is equivalent to the inequality $\dim C^r_d \; \leq \; d-r$, which is what we actually showed in Chapter~\ref{new Jacobians chapter}. This in turn followed by combining the geometric form of Riemann-Roch with an elementary bound on the dimension of the variety of secant planes to a curve in projective space:

Theorem~\ref{basic linear independence} allows us to do exactly one better: 

\begin{lemma}(strong secant plane lemma)\label{Strong secant plane lemma}
Let $C \subset \PP^n$ be a smooth, irreducible and nondegenerate curve. If we denote by $\Sigma^r_d \subset C_d$ the locus of effective divisors $D$ of degree $d$ on $C$ with $\dim \overline D \leq d-r-1$, then for any $d \leq n$ and $r > 0$,
$$
\dim \Sigma^r_d \leq d-r-1.
$$
\end{lemma}

\begin{proof}
Consider the incidence correspondence: 
$$
\Gamma := \left\{ (D, H) \in \Sigma^r_d\times {\PP^n}^* \mid \overline D \subset H \right\}.
$$
The curve being nondegenerate, the projection map $\Gamma \to  {\PP^n}^*$ is finite. But the fibers of $\Gamma$ over $\Sigma^r_d$ have dimension at least $n-d+r$; if we had $\dim \Sigma^r_d \geq d-r$, it would follow that $\dim \Gamma \geq n$, and hence that the projection map $\Gamma \to  {\PP^n}^*$ is dominant---contradicting Theorem~\ref{basic linear independence}.
\end{proof}

\begin{proof}[Proof of the case of equality in Martens theorem]
 Now, if a curve $C$ is non-hyperelliptic, we can apply the strong secant plane lemma to the canonical curve. Except for the trivial cases $d=r=0$ and $d=2g-2, r=g-1$,
 we can apply Lemma~\ref{Strong secan plane lemma} to conclude that $\dim W^r_d(C) \leq d-2r-1$; it follows that if we have $\dim W^r_d(C) = d-2r$ the curve $C$ in question must be hyperelliptic.
\end{proof}

Using the case of equality in Martens' Theorem, we can analyze equality in 
Clifford's Theorem:

\begin{corollary}\label{equality in Clifford from Martens}
If $C$ is a smooth curve of genus $g$ and $D$ a divisor on $C$ of degree $\leq 2g-2$,
such that $\deg D = 2r(D)$, then either $D =0$ or $D=K_C$ or $C$ is hyperelliptic.
\end{corollary}

\begin{proof}
If $C$ has a divisor $D$ with $\deg D =2 r(D)$ then by the Riemann-Roch theorem,  $\deg D  = 2(\deg D-g+h^1(D))$, 
so $\deg D = 2g-2h^1(D)$. If also $\deg D\leq 2g-2$, then $h^1(D) \geq 1$
and thus $r(D) >\deg D-g$. Since $W^{r(D)}_{\deg D}$ contains $D$ its dimension
is $\geq 0$, and we have a case of equality in Martens' theorem.
\end{proof}


\section{The $g+2$ theorem}\label{g+2 section}

Now that we have the strong form of Martens' theorem, we can prove the analogue of Theorem~\ref{g+3 theorem} for general linear series of degree $g+2$. This was stated in Section~\ref{g+3 section}, but we'll reproduce the statement here.

\begin{theorem}
Let $C$ be any smooth projective curve of genus $g$, and let $D$ be a general divisor of degree $g+2$ on $C$. 
The complete linear series $|D|$ is base-point free of dimension 2, and thus defines a map $\phi_D: C\to \PP^2$
which is birational onto its image $C_0$. Moreover, no two branches of $C_0$ share a tangent line. Furthermore,

\begin{enumerate}
\item If $C$ is nonhyperelliptic, then $C_0$ has only $\binom{g}{2}$ ordinary nodes and no other singularities; and
\item If $C$ is hyperelliptic,  $C_0$ has only one singular point, which is an ordinary $g$-fold point.\end{enumerate}
\end{theorem}

\begin{proof}
\begin{itemize}
\item (Dimension) Since $D$ is general, its points impose independent conditions on the $g$-dimensional space $H^0(K_C)$,
so $D$ is nonspecial, whence $h^0(D) = (g+2)-g+1 = 3.$

\item (No base points) If $|D|$ had a base point $p\in C$, then $3=h^0(D) = h^0(D-p) = (g+1)-g+1+h^0(K_C-D+p)$,
so $K_C-D$ is effective of degree $2g-2-(g+1) =g-3$; thus the family of such $D$ depends on only $g-3+1$ parameters,
and $D$ would lie in a proper subvariety of $Pic_{g+2}(C).$

\item (Birationality)
 To say that $\phi_D(p) = \phi_D(q)$ for some pair of points $p, q \in C$   means that $h^0(D-p-q) = 2$. By the Riemann-Roch
 theorem, $D-p-q = K-E$ for some effective divisor $E$ of degree $g-2$; Thus $\sO_C(D)$ is in the image of the map
$$
\nu : C_2 \times C_{g-2} \to \Pic_{g+2}(C)
$$
sending $(p+q, E)$ to $K_C - E + p + q$. 

The map $\nu$ is 
the composition of two surjective maps, $C_2 \times C_{g-2} \to C_g$ and $C_g\to Pic_g \to Pic_{g+2}$ by duality. Since the source and target of $\nu$ have the same dimension, the general fiber of $\nu$ is finite. Thus for general $D$ we  there are only finitely many pairs $p, q \in C$ such that $\phi_D(p) = \phi_D(q)$; it follows that $\phi_D$ is birational onto its image.


\item (Tacnodes):  To say that a pair of points $p, q \in C$ map to a point of $C_0$ such that the 
two branches have a common tangent means two things: that $h^0(D-p-q) \geq 2$; and that $h^0(D-2p-2q) \geq 1$. If this is the case, set $E = D - 2p - 2q$.  The condition $h^0(D-2p-2q) \geq 1$ is equivalent to  $E$ being an effective divisor. Considering the canonical map $\phi_K: C\to \PP^{g-1}$ 
the geometric Riemann-Roch Theorem $r(E) = \deg E -1-\dim\overline{\phi_K(E)}$ implies that $\overline {\phi_K(E)}$ has dimension $g-3$. The additional condition $h^0(D-p-q) \geq 2$ says that $\dim\overline{\phi_K(E+p+q)} = g-2$,
so the secant line $\overline{p,q}$ meets the $g-3$-plane $\overline{\phi_K(E)}$. Now, not every secant line to $C$ can meet a given linear subspace $\Lambda \subset \PP^{g-1}$ of dimension $g-3$--- otherwise, the projection $\pi_\Lambda : C \to \PP^1$ would be constant---so we see that $\mu(D)$ would have to lie on the image of a proper subvariety of $C_{g-2} \times C_2$ under the map $(E, p+q) \mapsto \sO(E+2p+2q)$ from $C_g-2 \times C_2$ to $Pic_{g+2}(C)$;
Since the dimension of this subvariety is $<g$ we see that if $C'$ has
a point at which two branches are tangent, then $\sO_C(D)$ lies in a proper subvariety of $\Pic_{g+2}(C)$. Since
the Abel-Jacobi map $C_{g+2} \to Pic_{g+2}(C)$ is surjective,
 $D$ would not be general. 
 

Let's now suppose that $C$ is non-hyperelliptic and $D$ is a general divisor of degree $g+2$ on $C$. To prove the theorem in this case, we have to show  three things: that the image $C_0 = \phi_D(C)$ does not have cusps or triple points.

\item (No cusps) To say that a point $p \in C$ maps to a cusp of $C_0$ (that is, the differential $d\phi_D$ is zero at $p$) amounts to saying that $h^0(D-2p) \geq 2$; that is, $D-2p$ is a $g^1_g$. But by Riemann-Roch, $W^1_g = K_C - W_{g-2}$; so to say $\phi_D$ has a cusp means that
$$
\mu(D) \in 2W_1 + K_C - W_{g-2},
$$
and since the locus on the right has dimension at most $g-1$, a general point of $J(C)$ will not lie in it. Note that this subsumes the fact that $|D|$ has no base points.

\item (no Triple points) To say that $C_0$ has a triple point means that for some divisor $E = p+q+r$ of degree 3, $h^0(D-E) \geq 1$; thus we must have 
$$
\mu(D) \in W_3 + W^1_{g-1}
$$
Now, to argue that this is not the case, we need to know that $\dim W^1_{g-1} \leq g-4$. Here we have to invoke Marten's theorem, which says that $\dim W^1_{g-1} \leq g-4$ if $C$ is non-hyperelliptic; given this, we conclude that $C_0$ has no triple points.


Thus, in the non-hyperelliptic case, the image curve $C_0 = \phi_D(C)$ has only nodes as singularities; the fact that there are exactly $\binom{g}{2}$ of them follows from the genus formula.
\end{itemize}

\noindent This concludes the proof in the non-hyperelliptic case. Now suppose that $C$ is hyperelliptic, and let $|E|$ be the  $g^1_2$ on $C$. If  $D$ is any divisor of degree $g+2$, the divisor $D - E$ will have degree $g$, and so be effective; thus we can write
$$
D \sim E + p_1 + \dots + p_g
$$
for some $g$-tuple of points $p_i$. If the divisor $D$ is general, the points $p_i$ will be general as well, and in particular distinct.

Now, the fact that
$$
h^0(D - p_1 - \dots - p_g) = h^0(E) = 2 = h^0(D) - 1
$$
implies that $\phi_D$ maps all the points $p_i$ to the same point. The image curve $C_0$ thus has a point with at least $g$ branches. Since these branches cannot share a tangent line,
they are smooth and their tangents are distinct; that is, the point is an ordinary multiple point. By the adjunction
formula $p_a(C) = binomial{g+1}{2}$ and since $C$ has genus $g$ the singularity must have $\delta$ invariant
$binom{g+1}{2} -g = \binom{g}{2}$. Such a singularity cannot have multiplicity $>g$, and is thus an ordinary $g$-fold point. Since the $\delta$-invariant of such a point is $\binom{g}{2}$ by Proposition~\ref{effect of blowup on genus}.
\end{proof}





\section{Castelnuovo's theorem}

Clifford's theorem gives a complete and sharp answer to the question, ``what linear series can exist on a curve of genus $g$?
But maybe that wasn't the question we meant to ask! After all, we're interested in describing curves in projective space as images of abstract curves $C$ under maps given by linear systems on $C$. Observing that the linear series that achieve equality in Clifford's theorem give maps to $\PP^r$ that are 2 to 1 onto a rational curve, we might hope that we would have a different---and more meaningful---answer if we  restrict our attention to linear series $\cD = (\cL,V)$ for which the associated map $\phi_\cD$ is at least  birational. 

A classical result of Castelnuovo gives a sharp bound on the genus of a curve in terms of its degree. For positive integers $d$ and $r$, let $M := M(d,r) := \lfloor(d-1)/(r-1)\rfloor$, so that
$$
 d -1 = M(r-1) + \epsilon \quad \text{ for some $\epsilon = \epsilon(d,r)$ with }0 \leq \epsilon \leq r-2. 
$$

\begin{theorem}[Castelnuovo's bound]\label{Castelnuovo's bound}
If $C \subset \PP^r$ is a reduced, irreducible, nondegenerate curve of degree $d$ then, with $M$ and $\epsilon$
as above,
$$
p_a(C) \leq \pi(d,r) := \frac{M(M-1)}{2}(r-1) + M\epsilon.
$$
\end{theorem}

We will say that a curve achieving the bound is a \emph{Castelnuovo curve}. We will see in Chapter~\ref{ScrollsChapter} that the bound is sharp: for every $r$ and $d \geq r$, there exist curves with genus exactly $\pi(d,r)$. 

\begin{example}
For plane curves, $r=2, M = d-1, \ \epsilon = 0,$ so we recover the necessary $p = \pi(d,2) =  {d-1\choose 2}$. Thus every reduced irreducible
plane curve is a Castelnuovo curve.
\end{example}

\begin{example}If $r=3$ then
$$
\pi(d,3) = \lfloor d^2/4 \rfloor-d+1\, .
$$
It is easy to check that curves on a smooth quadric surface in each of the classes
$(d/2, d/2)$ or $((d-1)/2, (d+1)/2)$
achieve this bound, and a variant of this works for singular quadrics as well. The classification  explained in Chapter~\ref{ScrollsChapter}
shows that these are the only Castelnuovo curves in $\PP^3$.
\end{example}

\begin{fact}
A famous theorem of Gruson and Peskine~\ref{Gruson-Peskine} (see \ref{Hartshorne-report} for an exposition and also the cases of characteristic $>0$ and small fields) completes the picture of the possibilities for the degree $d$ and  genus $g$  of a smooth curve in $\PP^3$. If the curve does not lie on a plane or a quadric, then the genus satisfies the stronger inequality
$$
g\leq \pi_1(d,3) := \frac{d^2-3d}{6} +1
$$
and smooth curves with all such degree and genus exist; they can all be realized as as curves
on cubic or quartic surfaces.

Note that there can be gaps in the possible genera of curves a given degree: for example  $\pi_1(9,3) = 10<\pi(9,3) =12$ but there is
no curve of degree 9 and genus 11. 
The full range of possible degrees and genera for curves in $\PP^n$ remains open for larger $n$. 
\end{fact}

To prove Theorem~\ref{Castelnuovo's bound}, we will give lower bounds for the dimensions of the linear series  cut on a curve $C$ by hypersurfaces of degree $m$. For large values of $m$ the line bundle $\cO_C(m)$ is non-special and $H^1(\sI_{C/\PP^r}(m)) = 0$, so a lower bound on the dimension of its space of sections translates, via the Riemann-Roch theorem, into an upper bound on the genus $g$ and this is the Castelnuovo bound.

For the proof, the following definition will be convenient:

\begin{definition}
If $\sV = (V,\cL)$ is a linear system on a variety $X$ and $\Gamma$ is a subscheme then the number of conditions
imposed by $\Gamma$ on $\sV$ is the dimension of the image of $V$ in $H^0(\sL\mid_\Gamma) = H^0(\sL \otimes \sO_\Gamma)$; or, numerically,
$$
\dim(V) - \dim \left(V \cap H^0(\cL\otimes \cI_{\Gamma/X}) \right).
$$\end{definition}

Thus, for example, if $\Gamma \subset \PP^r$, then the number of conditions imposed by $\Gamma$ on $H^0(\cO_{\PP^r}(m))$ is the value $h_\Gamma(m)$ of the Hilbert function of $\Gamma$ at $m$.
Note that the number of conditions imposed by $\Gamma$ on a linear system $V$ is necessarily less than or equal to the degree $d$ of $\Gamma$; if it is equal we say that $\Gamma$ \emph{imposes independent conditions on $V$}.

\begin{proof}[Proof of Theorem~\ref{Castelnuovo's bound}]
Suppose that $C \subset \PP^r$ is an irreducible, nondegenerate curve, and let $\Gamma = C \cap H$ be a general hyperplane section of $C$. Let $V_m \subset H^0(\cO_C(m))$ be the linear series cut on $C$ by hypersurfaces of degree $m$ in $\PP^r$, that is, the image of the restriction map
$$
H^0(\cO_{\PP^r}(m)) \to H^0(\cO_C(m)).
$$
Since the ideal sheaf of the hyperplane section is $\cI_{\Gamma/C} = \sO_C(-1)$ we have:
\begin{align*}
h^0(\cO_C(m)) - h^0(\cO_C(m-1)) & = \text{\# of conditions imposed by $\Gamma$ on $H^0(\cO_C(m))$} \\
&\geq \text{\# of conditions imposed by $\Gamma$ on $V_m$} \\
&= \text{\# of conditions imposed by $\Gamma$ on $H^0(\cO_{\PP^r}(m))$} \\
&= h_\Gamma(m).
\end{align*}
Summing these relations, we see that the dimension $h^0(\cO_C(m))$ is bounded below by
$$
h^0(\cO_C(m)) \geq \sum_{k=0}^m h_\Gamma(k).
$$

Thus, in order to bound the genus $C$ from above, we have to bound the Hilbert function of its hyperplane section $\Gamma$  from below; and for this, we need to know something about the geometry of $\Gamma$. In fact, all we need to know is Lemma~\ref{general position lemma}, which says that the points of $\Gamma$ are in linear general position! The key fact is the

\begin{proposition}\label{min hilb}
If $\Gamma \subset \PP^n$ is a collection of $d$ points in linearly general position that span $\PP^n$, then 
$$
h_\Gamma(m) \geq 
\begin{cases}
&mn+1\hbox{ if $m\leq M(d,n+1)$}\\
&d \hbox{ otherwise.}
\end{cases}
$$
\end{proposition}

One way to understand the bound $mn+1$ is to realize that if $\Gamma$ is any finite subscheme of a rational normal curve $C\subset\PP^n$ of degree $n$, 
then $h_\Gamma(m) = min(\deg \Gamma, mn+1$ for every $m$ (Exercise~\ref{linear bound is sharp}).
  Thus the bound in Proposition~\ref{min hilb} is best possible.
On the other hand, sets of points on a rational normal curve are almost the only sets for which the bound is sharp:

\begin{fact}
If $d \geq 2n+3$, then any collection of $d$ points in $\PP^n$ in linearly general position with Hilbert function $h_\Gamma(2) = 2n+1$ must lie on a rational normal curve. See~\cite[Lemma 3.9]{MR685427} for a proof.
%\fix{true for any subscheme in lgp, or just points?}
 This is a key ingredient in Castelnuovo's theorem characterizing curves of maximal genus as
those lying in certain divisor classes on rational normal scrolls.
We'll see how it's applied in Chapter~\ref{ScrollsChapter}.
\end{fact}


\begin{proof}
Suppose first that $d \geq mn+1$, and let $p_1,\dots,p_{mn+1} \in \Gamma$ be any subset of $mn+1$ points. It suffices to show that $\Gamma' = \{p_1,\dots,p_{mn+1}\}$ imposes independent conditions of $H^0(\cO_{\PP^n}(m))$, that is, for any $p_i \in \Gamma'$ there is a hypersurface $X \subset \PP^n$ of degree $m$ containing all the points $p_1,\dots, \hat{p_i},\dots,p_{mn+1}$ but not containing $p_i$.

To construct such an $X$, group the $mn$ points of $\Gamma' \setminus \{p_i\}$ into $m$ subsets $\Gamma_k$ of cardinality $n$; each set $\Gamma_k$ will span a hyperplane $H_k \subset \PP^n$, and we can take $X = H_1 \cup \dots \cup H_m$. 

In the case where $d<mn+1$, we add $mn+1-d$ general points; each one imposes exactly one
additional condition on hypersurfaces of degree $m$.
\end{proof}

To complete the proof of Theorem~\ref{Castelnuovo's bound} all that remains is to add up the lower bounds in the proposition. To this end, let $C \subset \PP^r$ be as above an irreducible, nondegenerate curve of degree $d$, and again set 
$M = M(d,r) = \lfloor{\frac{d-1}{r-1}}\rfloor$.
We have 
\begin{align*}
h^0(\cO_C(M)) &\geq \sum_{k=0}^M h^0(\cO_C(k)) - h^0(\cO_C(k-1)) \\
&\geq  \sum_{k=0}^M (k(r-1)+1) \\
&= \frac{M(M+1)}{2}(r-1) + M + 1
\end{align*}
and similarly
$$
h^0(\cO_C(M+m)) \geq \frac{M(M+1)}{2}(r-1) + M  + md.+ 1
$$
For sufficiently large $m$, the line bundle $\cO_C(M+m)$ will be nonspecial, so by the Riemann-Roch Theorem,
\begin{align*}
g &= (M+m)d - h^0(\cO_C(M+m)) + 1 \\
&\leq (M+m)d - \bigl(  \frac{M(M+1)}{2}(r-1) + M + 1 + md \bigr)+1 \\
& = M\bigl( M(r-1) + 1 + \epsilon \bigr) - \bigl(  \frac{M(M+1)}{2}(r-1) + M  \bigr) \\
&= \frac{M(M-1)}{2}(r-1) + M\epsilon.
\end{align*}
 \end{proof}

From the proof we see that if  $C\subset \PP^r$ is a curve whose degree and genus achieve equality, then all the inequalities in the proof
must be equalities; in particular, the map $H^0(\sO_{\PP^r}(d)) \to H^0(\sO_C(d))$ is surjective for every $d$ so, as we'll explain
in Chapter~\ref{linkage}, the curve is arithmetically Cohen-Macaulay.



\section{Uniform position} \label{uniformSection}
We now return to the situation where the ground field $k$ is algebraically closed of characteristic 0.

The Uniform Position lemma deals with the \emph{monodromy group} of the points of a general hyperplane section of a curve $C \subset \PP^r$. To prove it we will use the classical topology (an equivalent definition is described in Cheerful Fact~\ref{Galois equals monodromy} below). 

We may describe the monodromy group informally as follows: Suppose that $C \subset \PP^r$ is an irreducible curve over $\CC$, and $H_0 \subset \PP^r$ a hyperplane transverse to $C$; say the intersection $C \cap H_0 = \{p_1,\dots,p_d\}$. As we vary $H_0$ continuously along a real arc $\{H_t\}$, staying within the open subset $U \subset {\PP^r}^*$ of hyperplanes transverse to $C$, we can ``follow" each of the points $p_i(t)$ of intersection of $C$ with the hyperplane $H_t$.

Now imagine that the hyperplanes $H_t$ come back to the original $H_0$ at some time; that is, we have a continuous family $\{H_t\}_{0 \leq t \leq 1}$ with $H_1 = H_0$. Each of the points $p_i$ then traces out a continuous real arc 
$\{p_i(t) \in C \cap H_t\}_{0 \leq t \leq 1}$. Since $H_1 = H_0$, the end point $p_i(1)$ is one of the original points $p_j \in C \cap H_0$. In this way, we get a permutation of the set $C \cap H_0$; the group of all permutations arrived at in this way will be called the \emph{monodromy group} of the points $C \cap H_0$. 

We will now give a precise definition of the monodromy group in a more general setting, and prove that points of a general hyperplane section of an irreducible curve
the monodromy group is the full symmetric group; this is the uniform position lemma. The rest of the chapter will be a series of applications.

\subsection{The monodromy group of a generically finite morphism}

Let $f : Y \to X$ be a dominant map between varieties of the same dimension over $\CC$, and suppose that $X$ is irreducible. There is then an open subset $U \subset X$ such that $U$ and 
its preimage $V = f^{-1}(U)$ are smooth, and the restriction of $f$ to $V$ is a covering space in the classical topology. Let $d$ be the number of sheets. This is the degree of the extension $K(Y)/K(X)$. %\fix{should we put in the proof?}

Homotopy theory  associates a monodromy group to any finite topological covering map $f : V \to U$, defined as follows: Choose a base point $p_0 \in U$, and suppose $\Gamma := f^{-1}(p_0)  = \{q_1,\dots,q_d\}$. If $\gamma$ is any loop in $U$ with base point $p_0$, for any $i = 1, \dots, d$ there is a unique lifting of $\gamma$ to an arc $\tilde \gamma_i$ in $V$ with initial point $\tilde \gamma_i(0) = q_i$ and end point $\tilde \gamma_i(1) = q_j$ for some $j \in \{1,2,\dots,d\}$. Since we could traverse the loop in the opposite direction, the index $j$ determines $i$, and the map $i\mapsto j$ is a permutation of $\{1,2,\dots,d\}$. 
Since the set $\Gamma$ is discreet, the permutation depends only on the class of $\gamma$ in $\pi_1(U,p_0)$ so we have defined a homomorphism to the symmetric group:
$$
\pi_1(U,p_0)  \to {\rm Perm}(\Gamma) \cong S_d.
$$
The image $M$ of this map is called the \emph{monodromy group} of the map $f$. It depends on the labeling of the points of $\Gamma$, but a change in labeling
only changes the group by conjugation with the corresponding permutation, and the subgroup is well defined up to 
conjugation.

\begin{fact}\label{Galois equals monodromy}
In our setting $\pi_1(U,p_0)$, and thus the monodromy group, is independent of the choice of open set $U$: if $U' \subset U$ is a Zariski open subset, the complement of $U\setminus U'$ has
real codimension 2 so the map $\pi_1(U', p_0) \to \pi_1(U,p_0)$ is surjective. Thus the image of $\pi_1(U', p_0)$ in $S_d$ is the same. 

The theory of finite coverings of algebraic varieties is not only analogous to Galois theory, it \emph{is} Galois theory: In the situation described above, if we assume that $Y$ is irreducible, then the pullback map $f^*$ expresses the function field $K(Y)$ as a finite algebraic extension of $K(X)$. The degree of this extension is the degree of the map $f$. The monodromy group of $f$  is the Galois group of the Galois normalization of $K(Y)$ over $K(X)$ (see \cite{Harris1979}). Indeed, in early treatments of Galois theory, such as the famous \emph{Trait\'e des Substitutions}~\cite{MR1188877}, originally published in 1870, function fields played as large a role as number fields.
%\fix{what's the embedding in the symmetric group? must be the full perm group of the roots of primitive poly.}
\end{fact}

Since we assumed that $X$ is irreducible, the space $U$ is (path) connected, and it follows that the monodromy group is transitive if and only if the space $V$ is (path) connected. But $V$ is a smooth
variety, so this is the case if and only if $V$ is irreducible.

We will next compute the monodromy group of the  universal hyperplane section of a curve, constructed as follows:
Let $C \subset \PP^r$ be an irreducible, nondegenerate curve of degree $d$, and let $X = {\PP^r}^*$ be the space of hyperplanes in $\PP^r$. We define the \emph{universal hyperplane section of $C$} to be the projection  $f: Y\to {\PP^r}^*$ of the incidence variety
$$
Y = \{ (H, p) \in {\PP^r}^* \times C \mid p \in H \}.
$$
The fibers of $f$ are the hyperplane
sections of $C$, so $f$ is a dominant finite map. If we let $U\subset {\PP^r}^*$ be the open subset of hyperplanes
meeting $C$ transversely, then the restriction of $f$ to the preimage $V$ of $U$ is a covering space
whose fibers each consist of $d$ distinct points. The preimage in $Y$ of a point $p\in C$ is the set of hyperplanes containing
$p$, a copy of $\PP^{r-1}$, and thus $Y$ is irreducible. Thus the monodromy group of $f$ is transitive. But much more is true:

\begin{theorem}[Uniform Position Theorem]\label{uniform position lemma}
The monodromy group of the universal hyperplane section of an irreducible curve $C \subset \PP^r$ is the full symmetric group $S_d$.
\end{theorem}

Theorem~\ref{uniform position lemma} fails over fields of finite characteristic, though there is no known counterexample for smooth curves; see Exercise~\ref{strange curves} for singular examples, and \cite{Rathmann} and \cite{Kadets} for what is known. 

Theorem~\ref{uniform position lemma} implies that two subsets of the same cardinality in the general hyperplane section of $C$
are indistinguishable from the point of view of any discrete invariant that is semicontinuous in the Zariski topology. To make this precise, we introduce a definition:

\begin{definition}
Let $\phi : Y \to X$ be a finite morphism. By the \emph{restricted fiber power} $\tilde Y^n/X$ we will mean the complement of all diagonals in the ordinary fiber power; that is,
$$
\tilde Y^n/X := \{ (x, y_1,\dots, y_n) \in X \times Y^n \mid \phi(y_i) = x \text{ and } y_i \neq y_j \; \forall i \neq j \}
$$
\end{definition}

In down-to-earth terms, a point of $\tilde Y^n/X$ is a set of $n$ distinct points in a fiber of $\phi$ together with
the choice of a total order on these points. 

\begin{lemma}\label{transitivity lemma}
Let $f : Y \to X$ be a generically finite cover of degree $d$, with  monodromy group $M \subset S_d$.
$M$ is $n$ times transitive if and only if the restricted fiber power $\tilde Y^n/X$ is irreducible.
\end{lemma}

\begin{proof}
If we restrict ourselves to open subsets $U \subset X$ and $V = f^{-1}(U) \subset Y$ such that $U$ and $V$ are smooth and the restriction $f|_V : V \to U$ is a covering space map in the classical topology, then the restricted fiber powers are unions of connected component of the usual fiber power $Y^n/X$. The condition that the monodromy if $n$-times transitive is equivalent to the condition that the restricted fiber power $\tilde Y^n/X$ is connected; since the fiber powers are all smooth, this is equivalent to $\tilde Y^n/X$ being irreducible.
\end{proof}


\subsection{Proof of the uniform position theorem}

Let $C \subset \PP^r$ be an irreducible, nondegenerate curve of degree $d$ and $f : Y \subset {\PP^r}^* \times C \to  X = {\PP^r}^*$ its universal hyperplane section; let $U \subset {\PP^r}^*$ be the open subset of hyperplanes transverse to $C$ and $V = f^{-1}(U)$; let $M \subset S_d$ be the monodromy group of $V$ over $U$.
To show that  $M$ is the full symmetric group, it suffices to show that $M$ contains all transpositions, for this it is enough to show that $M$ is doubly transitive and contains one transposition.

To prove that $M$ is doubly transitive, we can give a concrete description of the restricted fiber power $\tilde V^2/U$: let
$$
\Sigma := \left\{ (H, p, q) \in {\PP^r}^* \times C \times C \mid p, q \in H \text{ and } p \neq q \right\}.
$$
Projection on the second and third factor expresses $\Sigma$ as a $\PP^{r-2}$-bundle over the complement $C \times C \setminus \Delta$ of the diagonal in $C \times C$. Thus $\Sigma$ is irreducible, and it follows that the restricted fiber square $\tilde V^2/U$, which is a Zariski open subset of $\Sigma$, is as well. Note that this part of the argument does not rely on any assumption about the characteristic.

Next we give a criterion for a monodromy group to contain a transposition:

\begin{lemma}\label{transposition lemma}
Let $f : Y \to X$ be a generically finite cover of degree $d$ over an irreducible variety $X$, with  monodromy group $M \subset S_d$.  
If,  for some smooth point $p \in X$ the fiber $f^{-1}(p)\subset V$ consists of $d-2$ reduced points $p_1,\dots, p_{d-2}$ and one point $q$ of multiplicity 2, where $q$ is also a smooth point of $Y$, then $M$ contains a transposition.
\end{lemma}

\begin{proof} Note that the hypothesis implies that $Y$ is smooth
locally near the fiber of $p$. Let $U \subset X$ be a Zariski open subset of the smooth locus in $X$, as in the definition of the monodromy group, so that  $V := f^{-1}(U)$ is also smooth and the restriction $f|_V : V \to U$ expresses $V$ as a finite $d$-sheeted covering space of $U$, with $U,V$ smooth and $p\in X$.

Let $B_i\subset Y$ be disjoint small connected closed neighborhoods of the points
in $f^{-1}(p)$. Because $f|_V$  is finite, the image $A := \cap f(B_i)$ has is a locally
closed subset of the same dimension as $X$ and thus $A$
 contains a neighborhood
of $p$ in the classical topology.

Let $p' \in A \cap U$. Two of the $d$ points of $f^{-1}(p')$ will lie in the component  of $B$ containing $q$; call these $q'$ and $q''$. Since $B \cap V$ is the complement of a proper subvariety in $B$ it is connected, and we can draw a real arc $\gamma : [0,1] \to B \cap V$ joining $q'$ to $q''$; by construction, the permutation of $f^{-1}(p')$ associated to the loop $f \circ \gamma$ will exchange $q'$ and $q''$ and fix each of the remaining $d-2$ points of $f^{-1}(p')$.
\end{proof}

\begin{proof}[Completion of the proof of the uniform position lemma]
 It remains to show that a reduced irreducible curve $C$ of degree $d$ (in characteristic 0)
 has a hyperplane section consisting of $d-2$ smooth points and one double point; that is, a hyperplane simply tangent to the curve at one point and transverse everywhere else. By the results of 
 Section~\ref{isolated flexes and bitangents} there is a tangent line $L$ to $C$ at a smooth point that is not flex tangent, and is not tangent to $C$ at any other point. A general hyperplane $H$ containing $L$ meets $C$ doubly in the point of
 tangency. At each of the other points of $L\cap C$ the intersection $C\cap H$ is transverse because there are only finitely many lines to avoid; and at points not on $L$ $H\cap C$ is transverse by Bertini's theorem. This completes the proof.
\end{proof}

\begin{remark}
 The fact that not every tangent line is bitangent is true
even for holomorphic arcs. Using the theory of correspondences it is possible to show that
a general tangent line to $C$ does not meet $C$ at any further points; but
this is not true of analytic arcs.
\end{remark}

 \section{Applications of uniform position}

Applying this to the universal hyperplane section of a curve $C \subset \PP^r$, a consequence of the Uniform Position Theorem is an irreducibility result:

\begin{corollary}\label{hyperplane section monodromy} If $C\subset \PP^r$ is a smooth curve of degree $d$, then 
for $k\leq d$, the restricted fibered powers $\tilde Y^k/X$  of the universal hyperplane section 
of $C$ are irreducible.
\end{corollary}


\begin{corollary}[numerical uniform position lemma]\label{numerical uniform position lemma}
Let $C\subset \PP^r$ be an irreducible curve, and let $\Gamma = H\cap C$ be a general hyperplane section. Any two subsets of $\Gamma$ with the same cardinality impose the same number of conditions on forms of any degree; that is, any two subsets of the same cardinality have the same Hilbert function.
\end{corollary}

\begin{proof} Let $U = {\PP^r}^* \setminus C^*$ be the open subset of hyperplanes transverse to $C$, and let $Y\to U$ be the universal hyperplane section.
Corollary~\ref{hyperplane section monodromy} says that the restricted fiber powers $\tilde V^n/U$ are irreducible.

Now, for each $m$ the number of conditions that $\Gamma$ imposes on forms of degree $m$ is lower semicontinuous, so it achieves its maximum on a Zariski open subset of $\tilde V^n/U$. Since $\tilde V^n/U$ is irreducible, the complement $Z$ of this open set has dimension strictly less than $\dim \tilde V^n/U = \dim U$. Thus a general hyperplane $H \in {\PP^r}^*$ will lie outside the image of $Z$, meaning that the number of conditions imposed by all the $k$-element subsets $\Gamma \subset C \cap H$ will have this maximal value.
\end{proof}

This result may be seen as an important strengthening of Theorem~\ref{basic linear independence}, since if $C$ is a reduced, irreducible nondegenerate curve in $\PP^n$ then a general subset of $n$ points of $C$ is linearly independent and spans a hyperplane; Corollary~\ref{numerical uniform position lemma} says that his must
be true for every subset of $n$ points of every general hyperplane section, 
which reproves Theorem~\ref{basic linear independence}, though only in characteristic 0. 

\section{Sums of linear series}

Another consequence of the uniform position lemma is a result about sums of linear series.
Recall that if $D$ is a divisor on a curve $C$ we write $r(D) = \dim |D| = h^0(\cO_C(D))-1$.

\begin{corollary}\label{Clifford equality plus}
If $D,E$ are effective divisors on a curve $C$ then
$$
r(D+E) \geq r(D)+r(E).
$$
If the genus of $C$ is $>0$ and $D+E$ is birationally very ample, then the inequality is strict.
\end{corollary}.

Note that on $\PP^1$ any effective divisor $D$ has $r(D) = \deg D$, so the inequality above is
always an equality for $C = \PP^1$.

\begin{proof}
 The inequality follows in general because the sums of divisors in $|D|$ and divisors in $|E|$ already move in 
 a family of dimension $r(D)+r(E)$; the key point is the strict inequality in case $D+E$ is birationally very ample.
 
If $D+E$ is birationally very ample then, restricting to an open set,
we may identify $C$ with its image under the complete linear series $|D+E|$, and we see that a general hyperplane section $H\cap C$ contains a divisor equivalent to $D$.

Let $Y$ be the $\deg D +\deg E$ restricted fiber power of the universal hyperplane.
A point $y\in Y$ is a hyperplane section plus an ordering of its points.  Let $\phi: Y \to \Pic_d(C)$ be the Abel-Jacobi map taking $y$
 to the class of the divisor that is the sum of first $d$ points in this order. The preimage  $Y'$ of the point of $\Pic_d(C)$ corresponding to the class of $D$ is a closed subset, and
since every divisor in the class of a hyperplane section contains a divisor
linearly equivalent to  $D$, the subvariety $Y'$ dominates $\PP^{n*}$, and thus
has the same dimension as $Y$. Consequently $Y'=Y$, and the sum of the first $d$ points
in any ordering of the general hyperplane section---that is, the sum of any $d$
of the points---is equivalent to $D$.

Thus if $p\in D$ and $q\notin D$, then $D-p+q \equiv D$, whence $q\equiv p$. Thus
$r(p)\geq 1$, so $C\cong \PP^1.$
\end{proof}



\section{Exercises}
\begin{exercise}\label{{arbitrary hyperplane examples}}
Let $C_1$ be the union of two skew lines in $\PP^3$, and let $C_2$ be the double line on a smooth quadric in $\PP^3$.
Show that there are hyperplane sections of $C_1$ and of $C_2$ that violate the conclusion of Proposition~\ref{arbitrary hyperplane}.
\end{exercise}

\begin{exercise}

Let $C\subset \PP^n$ be a smooth curve. If we re-embed $C$ by a Veronese map of sufficiently high degree---that is, we let $\nu_m : \PP^n \to \PP^N$ be the $m$th Veronese map, and let $\widetilde C = \nu_m(C)$ be image of $C$. Prove:

\begin{proposition}[Proposition~\ref{nodal projection} in characteristic $p$]\label{positive characteristic nodes}
The projection of $\widetilde C$ from a general $\PP^{N-3}$ is a nodal plane curve.
\end{proposition}

Hint: show that the items of~\ref{needed for nodes}  are true in this situation.
\end{exercise}

\begin{exercise}
Let $C_0$ be a plane quartic curve with two nodes $q_1, q_2$; let $\nu : C \to C_0$ be its normalization, and let $o \in C$ be any point not lying over a node of $C_0$.
By the genus formula, $C$ has genus 1. Using the construction above, describe the group law on $C$ with $o$ as origin.
\end{exercise}

\begin{exercise}\label{strange curves} \cite{Rathmann}. Let $k$ be an algebraically closed field of characteristic $p>0$, and let $q=p^e$ for some $e\geq 1$. Let $C\subset \PP^n$
be the closure of the image $C_0$ of the morphism
$$
\AA^1 \ni t \mapsto (t, t^q, t^{q^2}, \dots , t^{q^{n}}) \in \AA^n
$$
where $\AA^n\subset \PP^n$ is the open set $x_0=1$. 
\begin{enumerate}
\item Show that $C$ is a complete intersection, defined by the equations
$$
x_0^{q-1}x_2 - x_1^q, x_0^{q-1}x_3 - x_2^q,\dots, 
x_0^{q-1}x_n - x_{n-1}^q.
$$
\item Show that $C$ is singular unless $q = n = 2$.
\item Show every secant line to $C_0$ contains $q$ points of $C_0$; more generally, if
$a_1, \dots, a_r$ are linearly independent points of $C_0$, show that the linear span of
$a_1, \dots, a_r$ contains $q^{(r-1)}$ points of $C_0$.  Compare this with the configuration of
points in affine $n$-space over a field of $q$ elements.
\end{enumerate}
\end{exercise}

Here is a generalization of the uniform position theorem for varieties of higher dimension:

\begin{exercise}
Let $X \subset \PP^r$ be an irreducible variety of dimension $k$, and let $\Phi \subset \GG(r-k, r) \times X$ the universal $(r-k)$-plane section; that is,
$$
\Phi := \{ (\Lambda, p) \in \GG(r-k,r) \times X \mid p \in \Lambda \}.
$$
Show that the monodromy of the projection $\Phi \to \GG(r-k,r)$ is the full symmetric group.
\end{exercise}

As a consequence of this, we can deduce the Bertini irreducibility theorem:

\begin{exercise}
Let $X \subset \PP^r$ be an irreducible variety of dimension $k \geq 2$. Show that a general hyperplane section of $X$ is irreducible.
\end{exercise}

\begin{exercise}
Let $C \subset \PP^r$ be a union of irreducible curves $C_i$ of degrees $d_i$. Prove that the monodromy group of the points of a general hyperplane section of $C$ is the product $\prod S_{d_i}$.
\end{exercise}

Hint: We need to know that the dual hypersurfaces $C_i^* \subset {\PP^r}^*$ are all distinct; given this, we can exhibit loops that induce a given permutation of the points of $H \cap C_i$ while fixing the points of $H \cap C_j$ for $j \neq i$.


\begin{exercise}
Here is another approach to the $g+2$ theorem in the hyperelliptic case: 
Let $C$ be a hyperelliptic curve of genus $g$ and $D$ a general divisor of degree $g+1$ on $C$; let $|E|$ be the $g^1_2$ on $C$.
Consider the map $\phi : C \to \PP^1 \times \PP^1$ given as the product of the maps $\phi_D : C \to \PP^1$ and $\phi_E : C \to \PP^1$ given by the pencils $|D|$ and $|E|$.
\begin{enumerate}
\item Show that $\phi$ embeds the curve $C$ as a curve of bidegree $(g+1,2)$ on $\PP^1 \times \PP^1$.
\item Now embed $\PP^1 \times \PP^1$ into $\PP^3$ as a quadric surface $Q$; pick a general point $p \in C \subset Q$ and project $C$ from the point $p$. Show that the image curve $C_0$ is a plane curve of degree $g+2$ with one ordinary $g$-fold point.
\end{enumerate}
\end{exercise}

\begin{exercise}
Show that with $C$ and $M$ as above, the line bundle $\cO_C(M)$ is nonspecial. (We will see in Section~\ref{} that this is sharp; that is, there exist curves $C \subset \PP^r$ with $\cO_C(M-1)$ special).
\end{exercise}

\begin{exercise}\label{extremal m-ics}
Establish the analog of Proposition~\ref{rnc on most quadrics} for hypersurfaces of any degree $m$, that is to say no irreducible, nondegenerate curve in $\PP^r$ lies on more hypersurfaces of degree $m$ than the rational normal curve.
To do this, let $C\subset \PP^d$ be any irreducible nondegenerate curve. Let $\Gamma$ be a general hyperplane section
of $C$, and use the exact sequences
$$
0 \to \cI_{C/\PP^d}(l-1) \to \cI_{C/\PP^d}(l) \to \cI_{\Gamma/\PP^{d-1}}(l) \to 0.
$$ 
with $2 \leq l \leq m$ to show that
$$
h^0(\cI_{C/\PP^d}(m)) \leq  \binom{d+m}{m} - (md+1)
$$
with equality only if $C$ is a rational normal curve.
\end{exercise}

\begin{exercise}\label{linear bound is sharp}
Let $D \subset \PP^n$ be a rational normal curve. If $\Gamma \subset D$ is any collection of $d$ points on $D$ (or for that matter any subscheme of $D$ of degree $d$) then the Hilbert function of $\Gamma$ is
$$
h_\Gamma(m) = \min\{d, mn+1\}
$$
\end{exercise} 

\begin{exercise}
Let $C \subset \PP^r$ be an irreducible, nondegenerate curve of degree $d$, and set $M = \lfloor{\frac{d-1}{r-1}}\rfloor$ as in the proof of Castelnuovo's theorem.
Show that the line bundle $\cO_C(M)$ is nonspecial. (We will see in Section~\ref{} that this is sharp; that is, there exist curves $C \subset \PP^r$ with $\cO_C(M-1)$ special).
\end{exercise}

\input footer.tex