\input{header}

\chapter{History} 
\label{History}

In this Chapter we reproduce an excerpt of an unpublished essay by Jeremy Gray, sketching some of the landmarks in the history of algebraic geometry before the complex analytic work of
Abel, Jacobi and Riemann, and before Clebsch began the program of incorporating Riemann's insights into the theory of projective plane curves.

%\section{Greek mathematicians and conic sections}
Surviving sources suggest that the study of conic sections, literally sections of a cone, may well have arisen with work of Menaechmus (380--320 BCE) on doubling the cube. This was a long-standing problem with a theological spin -- double the size of an altar (the Delian problem).  Earlier, Hippocrates of Chios (470--410 BCE) had reduced the problem to that finding two mean proportionals between 1 and 2; i.e. the numbers $x$ and $y$ such that 
\begin{equation}~\label{Menmus}
1:x= x:y = y: 2.
\end{equation}
a quadratic equation.
For Greek mathematicians, these would have been line segments of those lengths. 

%Menaechmus may have been the first to connect the problem of doubling the cube to the idea of conic sections, or perhaps we should say quadratic curves, because it has been argued that he considered equation \eqref{Menmus} as defining or generating curves that would have been drawn and understood pointwise. I'll return to this when discussing graphing of curves. At all events, he expressed the solution to the Delian problem in terms of the intersection of a hyperbola and a parabola. 

About a century later, Diocles (290--180 BCE), and soon after him Apollonius of Perga (240--190 BCE), made significant progress with the conic sections.
Apollonius produced the first \emph{theory} of conic sections as sections of a cone. The names of the three types of section, the hyperbola, parabola, and ellipse,  are due to him.
% and the transformation from what Manaechmus did in solving problems to identifying a family of curves may be due to one Aristaeus, whose work is now lost. Diocles identified the curve that would focus the sun's rays to a point as the parabola, and knew precisely how to cut a cone so as to obtain it. 

%Apollonius produced the first \emph{theory} of conic sections as sections of a cone. The books are very dry. Van der Waerden called him a virtuoso ``in dealing with geometric algebra, and also a virtuoso in hiding his original line of thought'', while going on to say that ``his reasoning was crystal clear and elegant.'' It's hard to summarise Apollonius's work. He began with basic definitions of a cone (on a circle) and its three types of section: the hyperbola, parabola, and ellipse; the names are due to him. He showed that all the known sections that had been studied as sections of a cone with vertical angle a right angle could be obtained as sections of a suitable but otherwise arbitrary cone.  

%It's  arguable that a system of coordinates was present in his work, inasmuch as he referred everything to a pair of distinguished lines in the plane of section, but it is buried in the heavy use of proportion theory between different line segments, and you have to be very smart, as Apollonius undoubtedly was, to handle it. That said, he produced a theory of the principal diameters of conics, and studied such problems as finding the tangents to a conic from an exterior point, and came close to observing the properties of cross-ratio that modern writers detected. He also had a theory about normals to conics from which, apparently, it's a short step to obtaining their evolutes. 

We know this because four books of his conics survive in Greek and three more in Arabic; the eighth and final volume is lost. 

%A number of other Greek mathematicians pitched in. Sometimes their work survives directly, more often in the work of later writes such as Pappus; much of it, including a book by Euclid, who lived a little before Apollonius, is lost. 
%
%The biggest weakness in Apollonius's theory was that very often he had to treat the three kinds of conic section separately, although you can see him trying for greater generality.

%\section{Descartes,  coordinatising the plane,  graphing functions and curves}
%A common way to think of curves in antiquity was pointwise: some length depends in a given way on some other length. Accordingly, at least in principle, if you know the independent length
%(the ordinate, or $x$ coordinate) you know the dependent length (the abscissa, or $y$ coordinate).  

By the time of Descartes (1596--1650) there was already some sophisticated algebra expressed in a formalism that hadn't quite shaken off the Greek insistence on seeing everything as geometrical magnitudes: lengths, areas, volumes, and, well, what exactly? It was possible to write polynomial equations in this language. First Fermat (1607--1665) in 1636, and then much more boldly Descartes in 1637, realised that you could extend the language to two variables and so describe curves in the plane.

Descartes' s first achievement was to eliminate the dimensional aspect. A simple use of similar triangles allowed him to show that the product of two lengths could be seen as another length (not an area) so all geometrical quantities could be regarded as one-dimensional and the idea of dimension quietly dropped. Then came the real work. Almost all mathematical problems in his day were expressed in the language of geometry, except for some problems we would call diophantine and were implicitly about integers and rational numbers. Accordingly, the answer had to be expressed geometrically. Descartes's idea was to give letters to all the lengths involved in a problem, use the statement of the problem to express relationships between the letters, and reduce the equations to a single equation. Then solve the equation and express the answer again in geometrical terms.  

%His achievements include replacing the cumbersome algebra of his day, which was written in capital letters with abbreviations for the algebraic operations, with something much more like   what we use today. He wrote $x$ and $y$ for the key variables (Not $A$ and $E$ as hitherto). He was clear that he was using coordinates, although his $x$ and $y$ coordinates could have oblique axes. He could find normals to a curve at a point if the curve had an algebraic equation, and described a system of sliding rulers that he said could be adapted to draw any such curve. (Curves like the cycloid that were patently not algebraic he sought to exclude from geometry.) In this way he solved the famous Pappus problem: given four lines and four angles (nothing is lost if take these angles to be right angles), find the locus of points $P$ such that the product of the distances of $P$ from the first two lines is proportional to the product of the distances of $P$ from the last two lines. As he showed, and Pappus had known, the answer is a conic section, but 
%Descartes went further and claimed that in this way he could solve the Pappus problem for any number of lines. 

However, before Euler (1707--1783) in the late 1740s there was no concept of a function, everything was geometry, a relationship between two varying lengths.

The conic sections remained the primary object of study in algebraic geometry well into the 19th century, and
the transition to complex numbers happened only falteringly before 1850.\footnote{Euler's attitude that there was nothing to explain. We have these expressions of the form $a+bi$ and they behave arithmetically like numbers, so let's call them numbers, and as for what the symbol $i$ means well, to sure, you'll never meet a length of $i$ but we can imagine these expressions in our mind and control them, so what more do you want? Euler, and Cauchy after him, was not one for definitions of a philosophical nature, but what he was rejecting was the idea that ultimately mathematical quantities must be exhibited in nature: three sheep, a length of $\sqrt{2}$, and so on.} Several features promoted the use of synthetic methods (though algebraic methods were also often used.)  They can be elegant when algebraic methods are blunt; they correspond to the visual form of the conics; they provide a language for describing what is apparent or to be found in a problem. Against them is the obstinate fact that algebra is more general: it does not care if some quantities become negative, but what is a negative length? Once ways round that were found (by Poncelet and then Chasles, around 1820) the way was open to a truly systematic synthetic theory of conics. 

However, the synthetic methods did not extend to higher-order curves.  Although Euler and Cramer (1707--1752) had written on cubic and quartic curves, Pl\"ucker (1801--1868) saw that much needed to be done. His key idea to study families of curves, using the symbolic notation he devised: If $S_1$ and $S_2$ stand for the equations of two curves of the same degree, then $S_1 + \lambda S_2$ is the equation of another degree of that kind. But In his study of curves he discussed them first in  the plane, and then as they went off to infinity; he didn't say that the line at infinity could be mapped by a projective transformation into the finite part of the plane.

The way forward was indicated by M\"obius (1790--1868), who introduced barycentric coordinates: pick three points forming a triangle, and attach weights, positive, zero, or negative (not all zero) to these points to compute a point that is the barycenter.  The line at infinity, which is invisible in Cartesian coordinates, is a perfectly sensible line in barycentric coordinates. In this way M\"obius obtained a simple theory of conics and duality in the plane.  

However the first complete treatment of complex projective space is that of  von Staudt (1798--1867) in
his book Geometrie der Lage (Geometry of Position) (1847), and subsequent treatments were based on this. In 1922 H. F. Baker, the leading British algebraic geometer of his time, wrote: 
\begin{quotation}
 It was von Staudt to whom the elimination of the ideas of distance and congruence was a conscious aim, if, also, the recognition of the importance of this might have been much delayed save for the work of Cayley and Klein upon the projective theory of distance. Generalised, and combined with the subsequent Dissertation of Riemann, v. Staudt's volumes must be held to be the foundation of what, on its geometrical side, the Theory of Relativity, in Physics, may yet become \cite[p. 176]{MR2849917}.
\end{quotation}
\input{footer}