
\chapter*{Preface}
\label{preface}


In writing this book we have been keenly aware of  the influence of 
our great teachers in this subject,
\index{Buchsbaum, David A.}%
\index{Griffiths, Phillip}%
\index{Ven@van de Ven, Antonius}%
\index{Mumford, David}%
David Buchsbaum, Phillip Griffiths, David Mumford, and Antonius van de Ven. Great stylists
as well as great mathematicians, we can only hope that, consciously or not, the reader will occasionally hear their voices in our writing. 

We have taught large parts of this material at Harvard and Berkeley,
\index{Harvard University}%
\index{University of California, Berkeley}%
and our students' contributions to whatever virtues our text may have
are considerable. \.{I}zzet Co\c{s}kun, in particular, read an earlier
\index{Coskun@Co\c{s}kun, \.{I}zzet}%
version of this manuscript in detail, and made myriad useful comments.

We have had the good fortune to have, in the
\index{Levy, Silvio}%
role of copy-editor, a mathematician and friend, Silvio Levy, whose sense of style and publishing expertise informs the
layout of this book. Silvio also converted our sketches, done (to quote James Thurber) with a dull
pencil, into professional graphics that are mathematically accurate when possible.

\index{Eisenbud, Monika}%
\index{Robson, Vicky}%
Writing this book was an enterprise of years that involved the patience of more than just the authors, and we have been fortunate to have had the support of Monika and Vicky, wives who understood and valued our work during the hours ``away'' that were necessary.

To all these, our heartfelt thanks!

\vskip0.8\baselineskip
David and Joe, August, 2024

