%header and footer for separate chapter files

\documentclass[12pt, leqno]{report}
%\input style-for-curves.sty
   
   \usepackage{amsmath,amscd,amsthm,amssymb,amsxtra,latexsym,epsfig,epic,graphics}
\usepackage[matrix,arrow,curve]{xy}
\usepackage{graphicx}
\usepackage{diagrams}

\def\PP{{\mathbb P}}

\errorcontextlines=1000
%\usepackage{makeidx}
\let\see\relax

\overfullrule=5pt
\catcode`\@\active
\def@{\mskip1.5mu} %produce a small space in math with an @

\title{Prerequisites for Sanya Summer Graduate School on Algebraic Curves}
\author{ David Eisenbud and Joe Harris}

%\date{\today}
%%\date{}
%\title{Curves}
%%{\normalsize ***Preliminary Version***}} 
%\author{David Eisenbud and Joe Harris }
%
%\begin{document}

\begin{document}
\maketitle

\pagenumbering{roman}
\setcounter{page}{5}
%\begin{5}
%\end{5}
\pagenumbering{arabic}
%\tableofcontents
%\fi

Students should be comfortable with the following ideas, constructions, and results:

\section*{Algebraic Geometry}

\begin{itemize}
\item Abstract varieties and subvarieties of projective space at the level of \cite[Chapters 1--3 ]{S}.
%\item subscheme of projective space $=$ saturated homogeneous ideal in a polynomial ring \fix{Do we need this?}
\item The relations between (rational) maps to projective space, divisors and divisor classes, line bundle and invertible sheaves at the level of \cite[Chapter 8 and Appendix]{S}.
\item Definition of cohomology, cohomology of line bundles on $\PP^{n}$ \cite{FAC}.
\item Serre's theorems: vanishing of cohomology above the dimension; vanishing of higher cohomology of high twists \cite{FAC} or \cite[Ch 3]{H}.
\item Riemann-Roch for curves. \cite[Section IV.1]{H}.
\end{itemize}

\section*{Commutative Algebra}
\begin{itemize}
\item Basic Commutative algebra at the level of \cite{A}.
\item Some homological ideas: complexes, free resolutions,  Hilbert's Syzygy Theorem \cite[Chapter 1]{E}.
\end{itemize}

\begin{thebibliography}{99}
\bibitem{A} Atiyah, M. F. and McDonald, I. G. Introduction to commutative algebra. Addison-Wesley Publishing Co., Reading, Mass., 1969.

\bibitem{E} Eisenbud, D., Commutative algebra. With a view toward algebraic geometry. Graduate Texts in Mathematics, 150. Springer-Verlag, New York, 1995
\bibitem{H} Hartshorne, R., Algebraic geometry. Graduate Texts in Mathematics, No. 52. Springer-Verlag, New York-Heidelberg, 1977.
\bibitem{FAC} Serre, J.-P. Faisceau Alg\'ebriques Coh\'erents
Ann. of Math. (2) 61, (1955). 197--278. English Tranlation (Achinger) at https://achinger.impan.pl/fac/fac.pdf .
\bibitem{S} Smith, K., Kahanp\"a\"a, L., Kek\"al\"ainen, P., and Traves, W., An invitation to algebraic geometry. Universitext. Springer-Verlag, New York, 2000.
\end{thebibliography}
\end{document}

